### 2024-09-12 Docs/linear-reduction.tex
Proved equivalence between algorithm and eval

\input pdfToolbox
\input preamble

\setlayout{horizontal margin=2cm, vertical margin=2cm}

{\setbox0=\hbox{\setfontandscale{bf}{25pt}Linear Reduction}
\centerline{
    \vbox{
        \copy0
        \smallskip
        \hbox to\wd0{\hfil Ari Feiglin and Noam Kaplinski}
    }
}}

\bigskip
\hbox to\hsize{\hfil\vbox{\hsize=.7\hsize
\leftskip=0pt plus 1fil \rightskip=\leftskip \parfillskip=\z@
\hrule
\kern5pt

In this paper we will define the concept of linear reduction in the context of syntax parsing.
We will progress through more and more complicated examples, beginning from the programming of a simple calculator until we ultimately have created an extensible programming language, which we call Lang-X.

\kern5pt
\hrule
}\hfil}

\section*{Table of Contents}

\tableofcontents

\vfill\break

\setcounter{section}{-1}
\section{Notation}

\benum
    \item ${\bb N}$ denotes the set of natural numbers, including $0$.
    \item $\overline{\bb N}$ is defined to be ${\bb N}\cup\set\infty$.
    \item ${\bb Z}$ denotes the set of integers.
    \item $\overline{\bb Z}$ is defined to be ${\bb Z}\cup\set{\pm\infty}$.
    \item $f\colon A\plongto B$ means that $f$ is a partial function from $A$ to $B$.
    \item If ${\sf X}$ is a set and ${\sf x}$ is some symbol, then ${\sf X}_{\sf x}={\sf X}^{\sf x}={\sf X}\cup\set{{\sf x}}$.
    \item $\epsilon$ is the empty string, it and $\varnothing$ are also used to denote ``nothing'' in whatever context that may be.
    \item $(x_1,\dots,x_n)$ denotes a list.
    \item If $\ell_1,\ell_2$ are lists, $\ell_1@\ell_2$ is their concatenation.
    \item $t{::}\ell$ is the list whose first element is $t$ and whose tail is $\ell$.
\eenum
\vfill\break

\section{The Algorithm}

\subsection{Stateless Reduction}

The idea of linear reduction is simple: given a string $\xi$ the first character looks if it can bind with the second character to produce a new character, and the process repeats itself.
There is of course, nuance.
This nuance hides in the statement ``if it can bind'': we must define the rules for binding.

Let us define an {\it reducer} to be a tuple $(\Sigma,\beta,\pi)$ where $\Sigma$ is an alphabet; $\beta\colon\overline\Sigma\times\overline\Sigma\plongto\overline\Sigma$ is a partial function called the
{\it reduction function} where $\overline\Sigma=\Sigma\times\overline{\bb N}$; and $\pi$ is the {\it initial priority function}.
A {\it program} over an reducer is a string over $\overline\Sigma$.
We write a program like $\sigma^1_{i_1}\cdots\sigma^n_{i_n}$ instead of as pairs $(\sigma^1,i_1)\dots(\sigma^n,i_n)$.
In the character $\sigma_i$, we call $i$ the {\it priority} of $\sigma$.

Then the rules of reduction are as follows, meaning we define $\beta(\xi)$ for a program:
We do so in cases:
\benum
    \item If $\xi=\sigma_i$ then $\beta(\xi)=\sigma_0$.
    \item If $\xi=\sigma^1_i\sigma^2_j\xi'$ where $i\geq j$ and $\beta(\sigma^1_i,\sigma^2_j)=\sigma^3_k$ is defined then $\beta(\xi)=\sigma^3_k\xi'$.
    \item Otherwise, for $\xi=\sigma^1_i\sigma^2_j\xi'$, $\beta(\xi)=\sigma^1_i\beta(\sigma^2_j\xi')$.
\eenum

A string $\xi$ such that $\beta(\xi)=\xi$ is called {\it irreducible}.
Notice that it is possible for a string of length more than $1$ to be irreducible: for example if $\beta(\sigma^1,\sigma^2)$ is not defined then $\sigma^1_i\sigma^2_j$ is irreducible.
$$ \beta(\sigma_1\tau_2)\xvarrightarrow{(3)}\sigma_1\beta(\tau_2)\xvarrightarrow{(1)}\sigma_1\tau_2 $$
But such strings are not desired, since in the end we'd like a string to give us a value.
So an irreducible string which is not a single character is called {\it ill-written}, and a string which is not ill-written is {\it well-written}.

Now the initial priority function is $\pi\colon\Sigma\plongto\overline{\bb N}$ which gives characters their initial priority.
We can then canonically extend this to a function $\pi\colon\Sigma^*\plongto(\Sigma\times\overline{\bb N})^*$ defined by $\pi(\sigma^1\cdots\sigma^n)=\sigma^1_{\pi(\sigma^1)}\cdots\sigma^n_{\pi(\sigma^n)}$.
Then a $\beta$-reduction of a string $\xi\in\Sigma^*$ is taken to mean a $\beta$-reduction of $\pi(\xi)$.

Notice that once again we require that $\pi$ only be a partial function.
This is since that we don't always need every character in $\Sigma$ to have an initial priority; some symbols are only given their priority through the $\beta$-reduction of another pair of symbols.
So we now provide a new definition of a {\it program}, which is a string $\xi=\sigma^1\cdots\sigma^n\in\Sigma^*$ such that $\pi(\sigma^i)$ exists for all $1\leq i\leq n$.
We can only of course discuss the reductions of programs, as $\pi(\xi)$ is only defined if $\xi$ is a program.

\Example let $\Sigma={\bb N}\cup\set{+,\cdot}\cup\set{(n+),(n\cdot)}[n\in{\bb N}]$.
$\beta$ as follows:

\medskip
\centerline{
    \vtop{\ialign{\hfil$#$\hfil\tabskip=.25cm&\hfil$#$\hfil\cr
        \sigma^1_i,\sigma^2_j & \beta(\sigma^1_i,\sigma^2_j)\cr\noalign{\kern3pt\hrule\kern3pt}
        n,+ & (n+)\cr
        n,\cdot & (n\cdot)\cr
        (n+),m & n+m\cr
        (n\cdot),m & n\cdot m\cr
        (n\cdot),(m+) & (n\cdot m,+)\cr
        (n+),(m+) & (n+m,+)\cr
        (n\cdot),(m\cdot) & (n\cdot m,\cdot)\cr
    }}
}
\medskip
\noindent Where $n,m$ range over all values in ${\bb N}$.
Here $\beta(\sigma_i,\sigma_j)$'s priority is $j$.
We define the initial priorities
$$ \pi(n) = \infty,\quad \pi(+) = 1,\quad \pi(\cdot) = 2 $$

Now let us look at the string $1+2\cdot3+4;$.
Here,
$$ \eqalign{
    1_\infty+_12_\infty\cdot_23_\infty+_14_\infty &\longto (1+)_12_\infty\cdot_23_\infty+_14_\infty\cr
    &\longto (1+)_1(2\cdot)_23_\infty+_14_\infty\cr
    &\longto(1+)_1(2\cdot)_2(3+)_14_\infty\cr
    &\longto(1+)_1(6+)_14_\infty\cr
    &\longto(7+)_14_\infty\cr
    &\longto(7+)_14_0\cr
    &\longto(11)_0
} $$
So the rules for $\beta$ we supplied seem to be sufficient for computing arithmetic expressions following the order of operations.
\qedd

\Example We can also expand our language to include parentheses.
So our alphabet becomes $\Sigma={\bb N}\cup\set{+,\cdot,(,)}\cup\set{\uline{ n+},\uline{ n\cdot},\uline{ n)}}[n\in{\bb N}]$.
We distinguish between parentheses and bold parentheses for readability.
We extend $\beta$ as follows:

\medskip
\centerline{
    \vtop{\ialign{\hfil$#$\hfil\tabskip=.25cm&\hfil$#$\hfil\cr
        \sigma^1_i,\sigma^2_j & \beta(\sigma^1_i,\sigma^2_j)\cr\noalign{\kern3pt\hrule\kern3pt}
        n,+ & \uline{ n+}_j\cr
        n,\cdot & \uline{ n\cdot}_j\cr
        \uline{ n+},m & (n+m)_j\cr
        \uline{ n\cdot},m & (n\cdot m)_j\cr
        \uline{ n\cdot},\uline{ m+} & \uline{ n\cdot m,+}_j\cr
        \uline{ n+},\uline{ m+} & \uline{ n+m,+}_j\cr
        \uline{ n\cdot},\uline{ m\cdot} & \uline{ n\cdot m,\cdot}_j\cr
        n,) & \uline{ n)}_j\cr
        \uline{ n+},\uline{ m)} & \uline{ n+m)}_j\cr
        \uline{ n\cdot},\uline{ m)} & \uline{ n\cdot m)}_j\cr
        (,\uline{ n)} & n_i\cr
    }}
}
\medskip

\noindent $(n+m)_j$ means $n+m$ with a priority of $j$, not $\uline{ n+m}_j$.
And we define the initial priorities
$$ \pi(n) = \infty,\quad \pi(+) = 1,\quad \pi(\cdot) = 2,\quad \pi({(}) = \infty,\quad \pi({)}) = 0 $$
So for example reducing $2\cdot((1+2)\cdot2)+1$,
$$ \eqalign{
    2_\infty*_2{(_\infty}{(_\infty}1_\infty+_12_\infty{)_0}*_22_\infty{)_0}+_11_\infty &\longto \uline{2*}_2{(_\infty}{(_\infty}1_\infty+_12_\infty{)_0}*_22_\infty{)_0}+_11_\infty\cr
    &\longto \uline{2*}_2{(_\infty}{(_\infty}\uline{1+}_12_\infty{)_0}*_22_\infty{)_0}+_11_\infty\cr
    &\longto \uline{2*}_2{(_\infty}{(_\infty}\uline{1+}_1\uline{2)}_0*_22_\infty{)_0}+_11_\infty\cr
    &\longto \uline{2*}_2{(_\infty}{(_\infty}\uline{3)}_0*_22_\infty{)_0}+_11_\infty\cr
    &\longto \uline{2*}_2{(_\infty}3_\infty*_22_\infty{)_0}+_11_\infty\cr
    &\longto \uline{2*}_2{(_\infty}\uline{3*}_22_\infty{)_0}+_11_\infty\cr
    &\longto \uline{2*}_2{(_\infty}\uline{3*}_2\uline{2)}_0+_11_\infty\cr
    &\longto \uline{2*}_2{(_\infty}\uline{6)}_0+_11_\infty\cr
    &\longto \uline{2*}_26_\infty+_11_\infty\cr
    &\longto \uline{2*}_2\uline{6+}_11_\infty\cr
    &\longto \uline{12+}_11_\infty\cr
    &\longto \uline{12+}_11_0\cr
    &\longto 13_0\cr
} $$
\qedd

\subsection{Stateful Reduction}

\noindent We define the following four base sets:
\benum
    \item ${\cal U}$ the universe of {\it values}, these are all the internal values an object may have.
    \item $\term_{\cal P}$ the set of {\it printable terms}, these are the tokens which a programmer may pass to the reducer.
    \item $\term_{\type}$ the set of {\it type terms}.
    \item $\term_{\cal A}$ the set of {\it abstract terms}.
\eenum
The sets $\term_{\cal P},\term_\type,\term_{\cal A}$ are all disjoint, we place no such restriction on ${\cal U}$ as the purpose it serves is different.
Let ${\cal A}$ be a set of {\it atomic abstract terms}, then the construction of abstract terms is
$$ \term_{\cal A} \ccoloneqq {\cal A}\mid{\cal A}\term_\type $$
And let $\type$ be a set of {\it atomic types}, each with an associated arity, which may be $\infty$.
Let $\type^n$ be the set of atomic types of arity $n$, then the construction of type terms is
$$ \term_\type \ccoloneqq \type^0 \mid \type^n\term_\type^1\cdots\term_\type^n\mid\type^\infty\term_\type^1\cdots\term_\type^n $$
as $n$ ranges over all ${\bb N}_{>0}$.

Define
\benum
    \item $\term\coloneqq\term_{\cal P}\cup\term_\type\cup\term_{\cal A}$ the set of {\it basic terms}.
    \item $\term_{\cal I}\coloneqq\term_\type\cup\term_{\cal A}$ the set of {\it internal terms}.
    \item $\Pi_{\cal I}\coloneqq\term_{\cal I}\times{\cal U}$ the set of {\it termed values}.
    \item $\Pi\coloneqq\Pi_{\cal I}\cup\term_{\cal P}$ the set of {\it atomic expressions}.
\eenum
Elements of $\overline\Pi$ will be written like $\sigma_n(v)$ where $\sigma$ is the term, $n$ the priority, and $v$ the value (nothing for printable terms).

We define the {\it initial priority function} as a function $\pi\colon\term_{\cal P}\longto\overline{\bb Z}$.
This can be extended canonically to a function $\pi\colon\term_{\cal P}^*\longto\overline{\Pi}^*$.

In stateful reduction, we abstract away some inputs to the initial beta-reducer in order to allow for easier implementation.
An initial beta-reducer is a partial function
$$ \varwidehat\beta\colon \term_{\cal I}\times\term^\epsilon\plongto\term_{\cal I}^\epsilon\times(\overline{\bb Z}\times\overline{\bb Z}\to\overline{\bb Z})\times
({\cal U}\times{\cal U}\times{\rm State}\pto{\cal U}\times\term_{\cal P}^*\times{\rm State}) $$

We extend this to a derived $\beta$-reducer,
$$ \beta\colon\overline\Pi^*\times{\rm State}\plongto\overline\Pi^*\times{\rm State} $$
We also define $\beta^*$ where given an input $\gen{\xi}[s]$, it runs $\beta$ on it iteratively until convergence (of $\xi$).
$\beta$ is defined with the following rules: given an input $\gen{\xi}[s]$ its image is
\benum
    \item If $\xi=\sigma_n\xi'$ for $\sigma\in\term_{\cal P}$ then
        $$ \beta\gen{\xi}[s] = \gen{s(\sigma)_n\xi'}[s] . $$
    \item If $\xi=\sigma_i(v)\xi'$ and $\hat\beta(\sigma,\epsilon)=(\alpha,\rho,f)$ is defined, then if $f(v,\_,s)=(w,\zeta,s')$ and $\rho(i)=k$ and $\beta^*\gen{\pi\zeta}[s']=\gen{\zeta'}[s'']$ then
        $$ \beta\gen{\xi}[s] = \gen{\alpha_k(w)\zeta'\xi'}[s''] . $$
    \item If $\xi=\sigma_i(v)\tau_j(u)\xi'$ and $i\geq j$ and $\hat\beta(\sigma,\tau)=(\alpha,\rho,f)$ is defined, then if $f(v,u,s)=(w,\zeta,s')$, $\rho(i,j)=k$, and
        $\beta^*\gen{\pi\zeta}[s']=\gen{\zeta'}[s'']$ then
        $$ \beta\gen{\xi}[s] = \gen{\alpha_k(w)\zeta'\xi'}[s''] . $$
    \item Otherwise, if $\xi=\sigma_i(v)\xi'$ and $\beta\gen{\xi'}[s]=\gen{\xi''}[s']$,
        $$ \beta\gen{\xi}[s] = \gen{\sigma_i(v)\xi''}[s'] . $$
\eenum

\subsubsection{States}

Similar to before, we define point-states as partial maps $\term_{\cal P}\plongto\Pi_{\cal I}$.
And if $s_1,s_2$ are two point-states and $\sigma\in\term_{\cal P}$ then
$$ s_1s_2(\sigma) = \cases{s_2(\sigma) & $\sigma\in{\rm dom}s_2$\cr s_1(\sigma) & $\sigma\in{\rm dom}s_1$} $$
We will denote finite point states as $[\sigma_1\mapsto\varkappa_1,\dots,\sigma_n\mapsto\varkappa_n]$, and this denotes the point-state which maps $\sigma_i$ to $\varkappa_i$.

A state will now have two fields: a sequence of point-states, as well as a sequence of indexes.
For a state $\bar s=\bigl[(s_1,\dots,s_n),I=(i_1,\dots,i_k)\bigr]$, let us define
\benum
    \item $\bar s+s=\bigl[(s_1,\dots,s_n,s),I\bigr]$
    \item $\bar s+_cs=\bigl[(s_1,\dots,s_n,s),(i_1,\dots,i_k,n+1)\bigr]$
    \item $\pop\bar s=\bigl[(s_1,\dots,s_{n-1}),I\bigr]$ if $i_k<n$ otherwise, $\bigl[(s_1,\dots,s_{n-1}),(i_1,\dots,i_{k-1})\bigr]$
    \item $\bar ss=\bigl[(s_1,\dots,s_ns),I\bigr]$
    \item $\bar s(\sigma)=s_1\cdots s_n(\sigma)$ for $\sigma\in\Sigma_P$
    \item $\bar s_c=s_{i_k}\cdots s_n$
\eenum
Furthermore, if $\sigma\in\term_{\cal P}$ and $\varkappa\in\Pi_{\cal I}$ let us define $\bar s\{\sigma\mapsto\varkappa\}$ as $(s_1,\dots,s_i[\sigma\mapsto\varkappa],\dots,s_n)$ where $i$ is the maximum
index such that $\sigma\in{\rm dom}s_i$.

\vfill\break

\section{The Grammar}

In this section we discuss the grammar of the language of Lang-X.
This is not naturally imposed by the parser, but it will properly parse programs of this form.

\medskip
\grammar
\noalign{\hbox{\bf Identifiers}}
    str&\ccoloneqq\(a\dots z\gbar A\dots Z\gbar\_\)\cr
    digit&\ccoloneqq\(0\dots9\)\cr
    ident&\ccoloneqq{\sl str}\ \({\sl str}\gbar{\sl digit}\)^*\cr
\noalign{\hbox{\bf Constant Expressions}}
    const&\ccoloneqq\({\sl number}\gbar{\sl product}\gbar{\sl list}\)\cr
    number&\ccoloneqq\({\sl digit}\)^*[.\({\sl digit}\)^*]\cr
    product&\coloneqq({\sl production}\ \(,{\sl production}\)^*)\cr
    production&\coloneqq\({\sl expr}\gbar{\sl product}\)\cr
\noalign{\hbox{\bf Expressions}}
    op&\ccoloneqq+\gbar*\gbar/\cr
    pop&\ccoloneqq-\cr
    expr&\ccoloneqq\hphantom{{}\gbar}{\sl ident}\cr
    &\bbar{\sl const}\cr
    &\bbar{\sl expr};\cr
    &\bbar{\sl expr}\ {\sl expr}\cr
    &\bbar{\sl primexpr}\cr
    &\bbar({\sl expr})\cr
    &\bbar{\sl expr}\ \({\sl op}\gbar{\sl pop}\)\ {\sl expr}\cr
    &\bbar{\sl pop}\ {\sl expr}\cr
    &\bbar{\sl expr}.{\sl expr}\cr
    &\bbar{\bf if}\ ({\sl expr})\ \{{\sl expr}\} \{{\sl expr}\}\cr
    &\bbar{\bf fun}\ {\sl ident}\ ({\sl pattern})\ \{{\sl expr}\}\cr
    primexpr&\ccoloneqq\hbox{\bf\_prim\_}{\sl ident}\cr
\egrammar

\vfill\break

\section{Initializing the Algorithm}

\subsection{The Initial Beta Reducer}

We now describe the initial beta reducer of Lang-X according to stateful reduction.
By convention, {\astyle atomic abstract terms} will be red, {\tstyle type terms} will be green, {\istyle internal terms} will be blue.

\noindent{\bf End}:
\blist
    \item $\isig\ \eend \longto \isig\ \minfty\ (u,\_,s\to u,\epsilon,s)$
\elist

\noindent{\bf Arithmetic}:
\blist
    \item $\tsig\ \op \longto \op\tsig\ \snd\ (u,f,s\to(u,f),\epsilon,s)$
    \item $\op\tsig\ \op\tsig \longto \op\tsig\ \snd\ \bigl((u,f),(v,g),s\to (f(u,v),g),\epsilon,s\bigr)$
    \item $\op\tsig\ \tsig \longto \tsig\ \snd\ \bigl((u,f),v,s\to f(u,v),\epsilon,s\bigr)$
    \item $\Pop\ \tsig \longto \tsig\ \snd\ \bigl((f,g),u,s\to f(u),\epsilon,s\bigr)$
    \item $\tsig\ \Pop \longto \op\tsig\ {\sf one}\ \bigl(u,(f,g),s\to(u,g),\epsilon,s\bigr)$
    \item $\tsig\ \rparen \longto \rparen\tsig\ \snd\ (u,\_,s\to u,\epsilon,s)$
    \item $\op\tsig\ \rparen\tsig \longto \rparen\tsig\ \snd\ \bigl((f,u),v,s\to f(u,v),\epsilon,s\bigr)$
    \item $\Pop\ \rparen\tsig \longto \rparen\tsig\ \snd\ \bigl((f,g),u,s\to f(u),\epsilon,s\bigr)$
    \item $\lparen\ \rparen\tsig \longto \tsig\ \fst\ (\_,u,s\to u,\epsilon,s)$
\elist

\noindent{\bf Lists}:
\blist
    \item $\lbrack\ \tsig \longto \lbrack\tsig\ \fst\ (\_,u,s\to (u),\epsilon,s)$
    \item $\lbrack\tsig\ \tsig \longto \lbrack\tsig\ \fst\ (\ell,u,s\to (\ell,u),\epsilon,s)$
    \item $\lbrack\tsig\ \rbrack \longto \list\tsig\ \pinfty\ (\ell,\_,s\to\ell,\epsilon,s)$
    \item $\period\ \num \longto \index\ \zero\ (\_,n,s\to n,\epsilon,s)$
    \item $\list\tsig\ \index \longto \tsig\ \fst\ (\ell,i,s\to \ell_i,\epsilon,s)$
\elist

\noindent{\bf Variables}:
\blist
    \item $\lett\ x \longto \letvar\ \snd\ (\_,\_,s\to(x,\varnothing),\epsilon,s)$
    \item $\letvar\ \index \longto \letvar\ \fst\ \bigl((x,\ell),n,s\to(x,(\ell,n)),\epsilon,s\bigr)$
    \item $\letvar\ \equal \longto \leteq\ \minfty\ \bigl((x,\ell),\_,s\to(x,\ell),\epsilon,s\bigr)$
    \item $\leteq\ \isig \longto \epsilon\ \varnothing\ \bigl((x,\ell),v,s\to\epsilon,\epsilon,s'\bigr)$ where $s'$ is $s[x\varmapsto\sigma(v)]$ if $\ell=\varnothing$ and otherwise let $t$ be the result of
        setting $s(x).\ell_1.\dots.\ell_n$ to $v$, then $s'=s[x\varmapsto t]$.
\elist

\noindent{\bf Scoping}:
\blist
    \item $\lbrace\ \epsilon \longto \epsilon\ \varnothing\ (\_,\_,s\to\epsilon,\epsilon,s+\varnothing)$
    \item $\rbrace\ \epsilon \longto \epsilon\ \varnothing\ (\_,\_,s\to\epsilon,\epsilon,\pop s)$
\elist

\noindent{\bf Products}:
\blist
    \item $\tsig\ \comma \longto \comma(\tsig)\ \snd\ (u,\_,s\to(u),\epsilon,s)$
    \item $\op\tsig\ \comma(\tsig) \longto \comma(\tsig)\ \snd\ ((f,u),(v)\to (f(u,v)),\epsilon,s)$
    \item $\Pop\ \comma(\tsig) \longto \comma(\tsig)\ \snd\ ((f,g),(u)\to (f(u),\epsilon,s))$
    \item $\comma\tOmeg\ \comma(\tsig) \longto \comma(\tOmeg,\tsig)\ \snd\ (\ell,\ell',s\to(\ell,\ell'),\epsilon,s)$
    \item $\comma\tOmeg\ \rparen\tsig \longto \listrparen(\tOmeg,\tsig)\ \snd\ (\ell,v\to (\ell,v),\epsilon,s)$
    \item $\lparen\ \listrparen\tOmeg \longto \product\tOmeg\ \pinfty\ (\_,\ell,s\to \ell,\epsilon,s)$
\elist

\noindent{\bf Primitives}:
\blist
    \item $\primitive\ \isig \longto \epsilon\ \varnothing\ (f,v,s\to \epsilon,w,s)$ where $f(\isig,v)=(w,s')$ (the purpose is for $f$ to have a side effect)
\elist

\noindent{\bf Code Capture}
\blist
    \item $\lbrace\a\ x\longto \lbrace\a\ \pinfty\ (\xi,\_,s\to\xi x,\epsilon,s)$ if $x\neq\{,\}$
    \item $\lbrace\a\ x\longto \code\ \pinfty\ (\xi,\_,s\to\xi,\epsilon,s)$
    \item $\lbrace\a\ \code\longto \lbrace\a\ \pinfty\ (\xi,\xi',s\to\xi\{\xi'\},\epsilon,s)$
\elist

\noindent{\bf Parameter Capture}
\blist
    \item $\lparen\a\ x \longto \lparen\a\ \fst\ (\ell,\_,s\to \ell @(x),\epsilon,s)$ for $x\neq(,)$
    \item $\lparen\a\ ) \longto \plist\ \fst\ (\ell,\_,s\to \ell,\epsilon,s)$
    \item $\lparen\a\ \plist \longto \lparen\a\ \fst\ (\ell,\ell',s\to (\ell @(\ell')),\epsilon,s)$
\elist

\noindent{\bf Function Definitions}
\blist
    \item $\fun\ x\longto \funname\ \pinfty\ (\_,\_,s\to(x,\epsilon), \epsilon, s + [\{\mapsto\lbrace\a,\}\mapsto\rbrace\a,(\mapsto\lparen\a,)\mapsto\rparen\a])$
    \item $\funname\ \plist \longto \funvars\ \pinfty\ \bigl((x,\epsilon),u,s \to (x,u),\epsilon,s\bigr)$
    \item $\funvars\ \code \longto \closure\ \fst\ \bigl((x,\ell),\xi,s\to C=\gen{\ell,\xi,s'[x\mapsto\closure(C)]},\epsilon,\pop s[x\mapsto\closure(C)]\bigr)$ where $s'=(\pop s)_c$.
\elist

\noindent{\bf Function Calls}
\blist
    \item $\closure\ \tsig \longto \epsilon\ \varnothing\ \bigl(\gen{\ell,\xi,{\it ps}},u,s\mapsto\epsilon,\xi\},s+_c{\it ps}[\ell\mapsto\tsig(u)]\bigr)$ where $\ell\mapsto\tsig(u)$ means that if
        $\ell=(x)$ then $x\mapsto \tsig(u)$.
        Otherwise $\ell=(x_1,\dots,x_n)$, $\tsig=\product\tsig_1\cdots\tsig_n$, and $u=(u_1,\dots,u_n)$ and $x_i\mapsto\tsig_i(u_i)$ (recursively).
\elist

\noindent{\bf If Statements}
\blist
    \item $\iif\ \tsig \longto \ifbool\ \fst\ (\_,n,s\to n,\epsilon,s+[\{\mapsto\lbrace\a,(\mapsto\lparen\a)]$
    \item $\ifbool\ \code \longto \ifthen\ \fst\ (n,\xi,s\to(n,\xi),\epsilon,s)$
    \item $\ifthen\ \code \longto \epsilon\ \_\ ((n,\xi_1),\xi_2,s\to\varnothing,(n=0?\ \xi_2:\xi_1),\pop s)$
\elist

\noindent{\bf Types}
\blist
    \item $\tp\tsig\ \tp\ttau \longto \tp\tsig(\ttau)\ \snd\ (\_,\_,s\to \tsig(\ttau),\epsilon,s)$
    \item $\tp\tsig\ \product(\tp\ttau_1,\dots,\tp\ttau_n) \longto \tp\tsig(\ttau_1,\dots,\ttau_n)\ \snd\ (\_,\_,s\to \tsig(\ttau_1,\dots,\ttau_n),\epsilon,s)$
    \item $\ccolon\ \tp\tsig \longto \typer\tsig\ \snd\ (\_,u,s\to u,\epsilon,s)$
    \item $\tsig\ \typer\ttau \longto \ttau\ \snd\ (u,\_,s\to u,\epsilon,s)$
\elist

\subsection{The Initial State}

The initial state is a partial state, defined as follows:

\bigskip

\hbox{\vtop{\hsize=.48\hsize
\noindent{\bf End}
\blist
    \item $;{}\mapsto(\eend,\varnothing)$
\elist

\noindent{\bf Arithmetic}
\blist
    \item $({}\mapsto(\lparen,\varnothing)$
    \item $){}\mapsto(\rparen,\varnothing)$
    \item $+{}\mapsto(\op, (n,m\to n+m))$
    \item $+{}\mapsto(\op, (n,m\to n+m))$
    \item $*{}\mapsto(\op, (n,m\to n*m))$
    \item $/{}\mapsto(\op, (n,m\to n/m))$
    \item $-{}\mapsto(\Pop, (n\to-n), (n,m\to n-m))$
    \item $@{}\mapsto(\op, (\ell_1,\ell_2\to\ell_1@\ell_2))$
    \item $!={}\mapsto(\op,(u,v\to u\neq v))$
    \item $<={}\mapsto(\op,(n,m\to n\leq m))$
    \item $>={}\mapsto(\op,(n,m\to n\geq m))$
    \item $=={}\mapsto(\op,(u,v\to u=v))$
\elist
}\hfil
\vtop{\hsize=.48\hsize
\blist
    \item $<{}\mapsto(\op,(n,m\to n<m))$
    \item $>{}\mapsto(\op,(n,m\to n>m))$
\elist

\noindent{\bf Lists}
\blist
    \item $[{}\mapsto(\lbrack,[])$
    \item $]{}\mapsto(\rbrack,\varnothing)$
    \item $.\mapsto(\period,\varnothing)$
\elist

\noindent{\bf Variables}
\blist
    \item ${\rm let}\mapsto(\lett,\varnothing)$
    \item $={}\mapsto(\equal,\varnothing)$
\elist

\noindent{\bf Scoping}
\blist
    \item $\{{}\mapsto(\lbrace,\varnothing)$
    \item $\}{}\mapsto(\rbrace,\varnothing)$
\elist

\noindent{\bf Products}
\blist
    \item $,{}\mapsto(\comma,\varnothing)$
\elist

}}
\vfill\break

\hbox{\vtop{\hsize=.48\hsize
\noindent{\bf Primitives}
\blist
    \item ${\rm\_prim\_print}\mapsto$
        \hfil\break\null\hfill$(\primitive,(a,v\to{\rm print}(v); (\varnothing,\varnothing)))$
    \item ${\rm\_prim\_len}\mapsto(\primitive,(a,\ell\to\num,\abs\ell))$
    \item ${\rm\_prim\_tail}\mapsto(\primitive,(\tsig,t::\ell\to\tsig,\ell))$
    \item ${\rm\_prim\_type}\mapsto(\primitive,(\tsig,\_\to\tp\tsig,\tsig))$
\elist

\noindent{\bf Keywords}
\blist
    \item ${\rm fun}\mapsto(\fun,\varnothing)$
    \item ${\rm if}\mapsto(\iif,\varnothing)$
\elist
}
\vtop{\hsize=.48\hsize
\noindent{\bf Types}
\blist
    \item $:{}\mapsto(\colon,\varnothing)$
    \item ${\rm Num}\mapsto(\tp\ \num,\num)$
    \item ${\rm List}\mapsto(\tp\ \list,\list)$
    \item ${\rm Closure}\mapsto(\tp\ \closure,\closure)$
    \item ${\rm Product}\mapsto(\tp\ \product,\product)$
    \item ${\rm Primitive}\mapsto(\tp\ \primitive,\primitive)$
    \item ${\rm Type}\mapsto(\tp\ \tp, \tp)$
\elist
}}

\subsection{The Initial Priorities}

The initial priority function, $\pi$, is defined as follows:

\bigskip
\hbox{\vtop{\hsize=.48\hsize
\noindent{\bf End}
\blist
    \item $;{}\mapsto-\infty$
\elist

\noindent{\bf Arithmetic}
\blist
    \item $({}\mapsto\infty$
    \item $){}\mapsto0$
    \item $+{}\mapsto1$
    \item $*{}\mapsto2$
    \item $-{}\mapsto1$
    \item $/{}\mapsto2$
    \item $@{}\mapsto1$
    \item $=={}\mapsto0$
    \item $!={}\mapsto0$
    \item $<={}\mapsto0$
    \item $>={}\mapsto0$
    \item $<{}\mapsto0$
    \item $>{}\mapsto0$
\elist
}
\vtop{\hsize=.48\hsize
\noindent{\bf Lists}
\blist
    \item $[{}\mapsto0$
    \item $]{}\mapsto0$
    \item $.{}\mapsto0$
\elist

\noindent{\bf Variables}
\blist
    \item $={}\mapsto-\infty$
\elist

\noindent{\bf Scoping}
\blist
    \item $\{{}\mapsto0$
    \item $\}{}\mapsto0$
\elist

\noindent{\bf Products}
\blist
    \item $,{}\mapsto0$
\elist

\noindent Everything else is mapped to $\infty$
}}

\vfill\break

\section{Proving Equivalence}

In this section we will prove that our algorithm interprets according to the proper order of operations.
Specifically we will define a natural method of parsing and interpreting a numerical expression, and show that given a valid expression our algorithm gives the same result.

The problem can be formulated as follows: we are given the following:
\benum
    \item A set $\s X$, which is our {\it universe}.
    \item A set $\s S$ of {\it operator symbols}.
    \item For every operator symbol $\s s\in\s S$, a function $f_{\s s}\colon\s X\times\s X\longto\s X$.
    \item For every operator symbol $\s s\in\s S$, a {\it priority} $\pi(\s s)\in{\bb N}$.
\eenum
We define the following grammar of {\it expressions}:
$$ {\sl expr} \ccoloneqq ({\sl expr}) \gbar {\sl expr}\ \s S\ {\sl expr} \gbar \s X $$
\def\unparen{\hbox{\tencsc unparen}}%
All operators are left-associative.
We define $\unparen(\xi,i)$ to mean that the $i$th character in $\xi$ is not within parentheses (this can be implemented easily: iterate backwards and count how many opening and closing parentheses there
are.
If there are more opening than closing, than it is inside parentheses.)
We now define the evaluator for expressions:

\def\eval{\hbox{\tencsc eval}}
\algorithm
\Function{eval}{$\xi$}
    \If{$\xi=\s x\in\s X$}
        \State\Return $\s x$
    \ElseIf{$\xi[0]=$`('}
        \State $\xi$ is of the form $(\rho)\xi'$
        \State\Return $\eval(\eval(\rho)\xi')$
    \Else
        \Comment Find the greatest index with the smallest operator.\cr
        Note that $\max_i{\rm argmin}_if(i)$ is the maximal $i$ which minimizes $f$.\EndComment
        \State $i\coloneqq\max_i{\rm argmin}_i\set{\pi(\xi[i])}[{\xi[i]\in\s S,\unparen(\xi,i)}]$
        \State $\s s\coloneqq\xi[i]$
        \Comment $\xi'=\xi[:i-1]$, $\xi''=\xi[i+1:]$ \EndComment
        \State $\xi$ is of the form $\xi'\s s\xi''$
        \State\Return $f_{\s s}(\eval(\xi'),\eval(\xi''))$
    \EndIf
\EndFunc
\ealgorithm

For our beta reducer, our type term will be $\s X$, the atomic abstract terms will be $\s S,\Lparen,\Rparen$, and printable terms will be elements of $\s X\cup\s S$.
We will define the following initial beta reducer, we ignore states and printable term outputs because they are held constant:
\blist
    \item $\s X\ \s S\longto\s S\s X\ \snd\ (u,f\to(u,f))$
    \item $\s S\s X\ \s S\s X \longto\s S\s X\ \snd\ ((u,f),(v,g)\to(f(u,v),g))$
    \item $\s S\s X\ \s X\longto\s X\ \snd\ ((u,f),v\to f(u,v))$
    \item $\s X\ \Rparen \longto \Rparen\s X\ \snd\ (u,\_\to u)$
    \item $\s S\s X\ \Rparen\s X \longto\Rparen\s X\ \snd\ ((u,f),v\to f(u,v))$
    \item $\Lparen\ \Rparen\s X \longto\s X\ \snd\ (\_,u\to u)$
\elist
\noindent The initial state is self-explanatory.
The initial priorities map $\s x\in\s X$ to $\infty$, $\s s\in\s S$ to $\pi(\s s)$, $($ to $\infty$, and $)$ to $0$.
Since the state never changes, we can assume that all tokens are given their value in the initial state from the outset.

Let $\beta^*$ be the total $\beta$-reducer: it iteratively applies $\beta$ until convergence
We also define $\beta_0$ and $\beta^*_0$ which include the rule that if $\xi$ is a single character $\sigma_n(u)$ then $\beta_0\xi=\sigma_0(u)$.
We write $\xi\tos\xi'$ to mean that successive applications of $\beta_0$ lead from $\xi$ to $\xi'$.

Let us define the following set of {\it reduced expressions}, as a grammar including priorities:
$$ {\sl rexpr} \ccoloneqq \s X_n\gbar \s S\s X_n\ {\sl rexpr}\gbar \s X_n\ \s S_m\ {\sl rexpr}\ (n\geq m) $$
Say that a string of internal terms $S$ is {\it closed under $\beta$-reductions} if for every string $\xi\in S$ with arbitrary priorities, $\beta\gen\xi\in S$.

\lemm
    ${\sl rexpr}$ is closed under $\beta$-reductions.
    And under $\beta_0$-reductions, every reduced expression converges to an element in $\s X$.
\elemm

\Proof We induct on the length of the string $\xi$.
We split into cases
\benum
    \item A string of the form $\s X_n$ retains its value under a $\beta$-reduction.
    \item For string of the form $\s S\s X_n\xi$, we have the following cases:
        \benum
            \item $\xi=\s X_m$, if $n\geq m$ this becomes $\s X_m\in{\sl rexpr}$.
                Otherwise, a $\beta$-reduction gives $\s S\s X_n\beta\gen{\s X_m}=\s S\s X_n\s X_m\in{\sl rexpr}$.
            \item $\xi=\s S\s X_m\xi'$, if $n\geq m$ then this becomes $\s S\s X_m\xi'\in{\sl rexpr}$.
                Otherwise, a $\beta$-reduction gives $\s S\s X_n\beta\gen{\xi}$, and $\beta\gen{\xi}\in{\sl rexpr}$ inductively.
            \item $\xi=\s X_m\ \s S_k\xi'$, if $n\geq m$ then this becomes $\s X_m\ \s S_k\xi'\in{\sl rexpr}$.
                Otherwise, a $\beta$-reduction gives $\s S\s X_n\beta\gen{\xi}\in{\sl rexpr}$ inductively.
        \eenum
    \item For a string of the form $\s X_n\s S_m\xi$, if $n\geq m$ then a $\beta$-reduction gives $\s S\s X_m\xi\in{\sl rexpr}$.
        Otherwise we get $\s X_n\beta\gen{\s S_m\xi}$, but there are no reduction rules in which $\s S$ is the first term, so this is just $\s X_n\s S_m\gen\xi\in{\sl rexpr}$ inductively.
\eenum

For the second claim, we inductively show that if $\xi\neq\s X$ has a final priority of $0$ then a $\beta_0$ reduction will decrease its length by $1$, and if its final priority is $>0$ then it either
decreases its length by $1$ or sets its final priority to $0$.
\benum
    \item If $\xi=\s S\s X_n\xi'$, then a $\beta_0$-reduction either compounds $\s S\s X_n$ with another token, or gives $\s S\s X_n\beta_0\gen{\xi'}$, which inductively will decrease $\xi'$'s length by $1$
        if $\xi'$'s final priority is zero, or will decrease it by a token or set its final priority to zero otherwise.
    \item If $\xi=\s X_n\ \s S_m\xi'$, since $n\geq m$ then this reduces to $\s S\s X_m\xi'$, which is one token less.
\eenum
So this means that every two $\beta_0$-reductions (in fact every single $\beta_0$-reduction, since we can never increase a priority: once a token's priority is $0$, it will never increase) will decrease
the length of the string.
This means that eventually the string will converge to a single token.
Since ${\sl rexpr}$ is closed under $\beta_0$-reductions, this string must be in $\s X$.
\qed

Note that ${\sl expr}\subseteq{\sl rexpr}$ since ${\sl expr}\ccoloneqq\s X_\infty\gbar\s S_n\ \s X_\infty\ {\sl expr}$.
So when $\beta$-reducing an expression, we will remain inside ${\sl rexpr}$.

Notice that if $\xi\in{\sl rexpr}$ then $\beta^*\gen\xi$ cannot contain a substring of the form $\s X_n\ \s S_m$ since this can be reduced.
And since by the above lemma, $\beta^*\gen\xi\in{\sl rexpr}$, we have $\beta^*\gen\xi=\s S\s X_{n_1}\s S\s X_{n_2}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$ with $n_1<\cdots<n_k<n_{k+1}$.

Instead of writing
\blist
    \item $\s X_n(\s x)$, we write $\s x_n$.
    \item $\s S_n(f_{\s s})$, we write $\s s_n$.
    \item $\s S\s X_n(\s x,f_{\s s})$, we write $\s s_n[\s x]$.
\elist

We begin with the case that $\xi$ does not include any parentheses.
Let us define $\pi(\xi)=\set{\pi(\xi[i])}_{0\leq i<\abs\xi}$.

\lemm
    Let $\xi\in{\sl rexpr}$ such that $\pi(\xi)\geq n$, then $\beta^*\gen{\xi\s s_n}=\s s_n[\beta^*_0\gen\xi]$.
\elemm

\Proof we know that $\xi\s s_n$ is reduced to, at some point, $\beta^*\gen\xi\s s_n$, and as noted this is equal to
$$ \s s_{n_1}^1[\s x^1]\cdots\s s^k_{n_k}[\s x^k]\s x^{k+1}_{n_{k+1}}\s s_n,\qquad n\leq n_1<\cdots<n_k<n_{k+1} $$
Then a $\beta$-reduction gives
$$ \s s^1_{n_1}[\s x^1]\cdots\s s^k_{n_k}[\s x^k]\s s_n[\s x^{k+1}] $$
and further $\beta$-reductions give (instead $f_{\s s_i}$ we'll write $f_i$):
$$ \s s^1_{n_1}[\s x^1]\cdots\s s_n[f_k(\s x^k,\s x^{k+1})] \to \cdots \to \s s_n\bigl[f_1(\s x^1,f_2(\s x^2,\cdots))\bigr] $$
And we see that $\beta_0$-reducing $\xi$ gives precisely $f_1(\s x^1,f_2(\s x^2,\cdots))$, as required.
\qed

\lemm
    If $\xi\in{\sl rexpr}$ and $n<\pi(\xi)$ then $\beta^*_0\gen{\s s_n[\s x]\xi}=f_{\s s}(\s x,\beta^*_0\gen\xi)$.
\elemm

\Proof since $n<\pi(\xi)$, $\s s_n[\s x]$ cannot reduce with the first element of $\xi$, so $\beta_0\gen{\s s_n[\s x]\xi}=\s s_n[\s x]\beta_0\gen\xi$.
If $\xi$ can be $\beta$-reduced, then we simply induct on the new string, which is one token less.
Otherwise $\xi=\s s^1_{n_1}[\s x^1]\cdots\s s^k_{n_k}[\s x^k]\s x_{n_{k+1}}^{k+1}$ with $n_1<\cdots<n_{k+1}$, as remarked before.
Then we continue with a similar process as before.
\qed

Let $\xi\in{\sl expr}$, define ${\rm sym}(\xi)$ be the operator symbols in $\xi$.
By {\it inducting on the levels of priority}: we mean induction on $\pi{\rm sym}(\xi)$.

\thrm
    Let $\xi\in{\sl expr}$, then $\beta^*_0\gen\xi=\hbox{\tencsc eval}(\xi)$.
\ethrm

\Proof we induct on the levels of priority in $\xi$.
If $\xi=\s x_n$ then this is trivial.
Otherwise let $n$ be the lowest priority of operator in $\xi$, so $\xi$ is of the form
$$ \xi = \xi_1\s s_n^1\xi_2\s s_n^2\cdots\xi_k\s s_n^k\xi_{k+1} $$
Now notice that
$$ \xi \to \beta^*\gen{\xi_1\s s_n^1\xi_2\cdots\xi_k\s s_n^k}\xi_{k+1} $$
by lemma 4.0.2 this is equal to
$$ \xi \to \s s_n^k[\beta^*_0\gen{\xi_1\s s_n^1\cdots\s s_n^{k-1}\xi_k}]\xi_{k+1} $$
and by lemma 4.0.3 reducing this gives
$$ f_k\bigl(\beta^*_0\gen{\xi_1\s s^1_n\cdots\s s^k_n\xi_k},\beta^*_0\gen{\xi_{k+1}}\bigr) $$
Continuing inductively on $k$, we get
$$ f_k\bigl(\cdots f_2(f_1(\beta^*_0\gen{\xi_1},\beta^*_0\gen{\xi_2}),\beta^*_0\gen{\xi_3})\cdots,\beta^*_0\gen{\xi_{k+1}}\bigr) $$
By induction, this is just
$$ f_k\bigl(\cdots f_2(f_1(\eval(\xi_1),\eval(\xi_2)),\eval(\xi_3))\cdots,\eval(\xi_{k+1})\bigr) \eqnum[bsres] $$
Now notice that
$$ \eval(\xi) = f_k(\eval(\xi_1\s s_n^1\cdots\s s_n^{k-1}\xi_k),\eval(\xi_{k+1})) $$
And continuing we get that this is equal to $(\bsres)$, as required.
\qed

\bye

### 2024-09-14 Docs/linear-reduction.tex
Proved equivalence with parentheses

@@
 We define the following grammar of {\it expressions}:
-$$ {\sl expr} \ccoloneqq ({\sl expr}) \gbar {\sl expr}\ \s S\ {\sl expr} \gbar \s X $$
+$$ {\sl expr} \ccoloneqq \s L\ {\sl expr}\ \s R \gbar {\sl expr}\ \s S\ {\sl expr} \gbar \s X $$
 \def\unparen{\hbox{\tencsc unparen}}%
 All operators are left-associative.
+$\s L,\s R$ represent left and right parentheses respectively.
 We define $\unparen(\xi,i)$ to mean that the $i$th character in $\xi$ is not within parentheses (this can be implemented easily: iterate backwards and count how many opening and closing parentheses there
@@
         \State\Return $\s x$
-    \ElseIf{$\xi[0]=$`('}
-        \State $\xi$ is of the form $(\rho)\xi'$
+    \ElseIf{$\xi[0]=\s L$}
+        \State $\xi$ is of the form $\s L\rho\s R\xi'$
         \State\Return $\eval(\eval(\rho)\xi')$
@@
 
-For our beta reducer, our type term will be $\s X$, the atomic abstract terms will be $\s S,\Lparen,\Rparen$, and printable terms will be elements of $\s X\cup\s S$.
+For our beta reducer, our type term will be $\s X$, the atomic abstract terms will be $\s S,\s L,\s R$ ($\s L$ for left parentheses, $\s R$ for right), and printable terms will be elements of $\s X\cup\s S$.
 We will define the following initial beta reducer, we ignore states and printable term outputs because they are held constant:
@@
     \item $\s S\s X\ \s X\longto\s X\ \snd\ ((u,f),v\to f(u,v))$
-    \item $\s X\ \Rparen \longto \Rparen\s X\ \snd\ (u,\_\to u)$
-    \item $\s S\s X\ \Rparen\s X \longto\Rparen\s X\ \snd\ ((u,f),v\to f(u,v))$
-    \item $\Lparen\ \Rparen\s X \longto\s X\ \snd\ (\_,u\to u)$
+    \item $\s X\ \s R \longto \s R\s X\ \snd\ (u,\_\to u)$
+    \item $\s S\s X\ \s R\s X \longto\s R\s X\ \snd\ ((u,f),v\to f(u,v))$
+    \item $\s L\ \s R\s X \longto\s X\ \fst\ (\_,u\to u)$
 \elist
@@
 
+\subsection{Expressions without Parentheses}
+
+First we will prove this result for expressions without parentheses, i.e. those of the grammar
+$$ {\sl expr} \ccoloneqq \s X_\infty\gbar \s X_\infty\ \s S_n\ {\sl expr} $$
 Let us define the following set of {\it reduced expressions}, as a grammar including priorities:
 $$ {\sl rexpr} \ccoloneqq \s X_n\gbar \s S\s X_n\ {\sl rexpr}\gbar \s X_n\ \s S_m\ {\sl rexpr}\ (n\geq m) $$
+Note that ${\sl expr}\subseteq{\sl rexpr}$.
 Say that a string of internal terms $S$ is {\it closed under $\beta$-reductions} if for every string $\xi\in S$ with arbitrary priorities, $\beta\gen\xi\in S$.
@@
 
-Note that ${\sl expr}\subseteq{\sl rexpr}$ since ${\sl expr}\ccoloneqq\s X_\infty\gbar\s S_n\ \s X_\infty\ {\sl expr}$.
-So when $\beta$-reducing an expression, we will remain inside ${\sl rexpr}$.
+Since ${\sl expr}\subseteq{\sl rexpr}$, when $\beta$-reducing an expression, we will remain inside ${\sl rexpr}$.
 
@@
 Now notice that
-$$ \xi \to \beta^*\gen{\xi_1\s s_n^1\xi_2\cdots\xi_k\s s_n^k}\xi_{k+1} $$
+$$ \xi \tos \beta^*\gen{\xi_1\s s_n^1\xi_2\cdots\xi_k\s s_n^k}\xi_{k+1} $$
 by lemma 4.0.2 this is equal to
-$$ \xi \to \s s_n^k[\beta^*_0\gen{\xi_1\s s_n^1\cdots\s s_n^{k-1}\xi_k}]\xi_{k+1} $$
+$$ \xi \tos \s s_n^k[\beta^*_0\gen{\xi_1\s s_n^1\cdots\s s_n^{k-1}\xi_k}]\xi_{k+1} $$
 and by lemma 4.0.3 reducing this gives
@@
 
+\subsection{Expressions with Parentheses}
+
+\def\exprp{\begingroup\sl expr\endgroup_P}
+\def\rexprp{\begingroup\sl rexpr\endgroup_P}
+Our expressions will now be of the grammar
+$$ \exprp \ccoloneqq \s X_\infty\gbar \s X_\infty\ \s S_n\ \exprp\gbar \s L_\infty\ \exprp\ \s R_0 $$
+and {\it reduced expressions} will be
+$$ \rexprp \ccoloneqq \s X_n\gbar \s S\s X_n\ \rexprp\gbar \s X_n\ \s S_m\ \rexprp\ (n\geq m)\gbar \s L_\infty\ \rexprp\ \s R_0\gbar \s L_\infty\ (\s S\s X_n)^*\ \s R\s X_0 $$
+
+\lemm
+    $\rexprp$ is closed under $\beta$-reductions.
+\elemm
+
+\Proof we will prove the following stronger result:
+
+\centerline{Let $\xi\in\rexprp$ then $\beta\gen\xi\in\rexprp$ and $\beta^*\gen\xi=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$ with $n_1<\cdots<n_{k+1}$.}
+\benum
+    \item For $\xi=\s X_n$ this is trivial.
+    \item For $\xi=\s S\s X_n\xi'$, if $\s S\s X_n$ can reduce with the first token of $\xi'$, so $\xi'=\s X_m$ or $\xi'=\s S\s X_m\xi''$, then we reduce and continue by induction.
+        Otherwise $\beta\gen\xi=\s S\s X_n\beta\gen{\xi'}$ which is inductively in $\rexprp$.
+        And if $\beta\gen{\xi'}$'s length is decreased, we induct.
+        Otherwise $\beta^*\gen{\xi'}=\xi'$ and so $\xi'=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$, so we get that $\xi=\s S\s X_n\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$ which we already
+        know from the previous section converges to the desired form.
+    \item For $\xi=\s X_n\s S_m\xi'$, we have that $\beta\gen\xi=\s S\s X_m\xi'$ and so we induct.
+    \item For $\xi=\s L_\infty\xi'\s R_0$, since $\s L_\infty$ cannot reduce with the first character of $\xi'$ (as it is not $\s R\s X$), $\beta\gen\xi=\s L_\infty\beta\gen{\xi'\s R_0}$.
+        If $\xi'$ can be reduced, then this is just $\s L_\infty\beta\gen{\xi'}\s R_0$ and we induct.
+        Otherwise, $\xi'=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$ with $n_1<\cdots<n_{k+1}$, and so this becomes $\s L_\infty\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s R\s X_0$, which is covered by the
+        final rule.
+    \item For $\xi=\s L_\infty\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s R\s X_0=\s L_\infty\xi'\s R\s X_0$, if $\xi'$ can be reduced then we simply induct.
+        Otherwise $n_1<\cdots<n_k$ and so a $\beta$-reduction gives $\s L_\infty\s S\s X_{n_1}\cdots\s S\s X_{n_{k-1}}\s R\s X_0$, and we induct.
+        The base is taken by $k=0$ in which case we get $\s L_\infty\s R\s X_0\to\s X_\infty$.
+        \qed
+\eenum
+
+From this lemma, since $\beta^*\gen\xi=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$, we get that $\beta^*_0\gen\xi\in\s X$.
+
+Further note from $(4)$ we can infer that $\s L_\infty\xi\s R_0\tos\s L_\infty\beta^*\gen\xi\s R_0$.
+From the proof of this lemma we have that $\beta^*\gen\xi=\s s_{n_1}^1[\s x^1]\cdots\s s_{n_k}^k[\s x^k]\s x_{n_{k+1}}^{k+1}$, and so we can see that
+$\s L_\infty\beta^*\gen\xi\s R_0\tos\beta^*_0\gen\xi_\infty$.
+Thus we have proven the following lemma:
+
+\lemm
+    Let $\xi\in\rexprp$, then $\beta^*\gen{\s L_\infty\xi\s R_0}=\beta^*_0\gen\xi_\infty$.
+\elemm
+
+\lemm
+    Let $\xi\in\rexprp$ and $n\leq\pi(\xi)$, then $\beta^*\gen{\xi\s s_n}=s_n[\beta^*_0\gen\xi]$.
+\elemm
+
+\Proof we know that
+$$ \xi\s s_n \tos \beta^*\gen\xi\s s_n = \s s^1_{n_1}[\s x^1]\cdots\s s^k_{n_k}[\s x^k]\s x_{n_{k+1}}^{k+1}\s s_n $$
+and so this just reduces to the same lemma as for expressions without parentheses.
+\qed
+
+\lemm
+    Let $\xi\in\rexprp$ and $n<\pi(\xi)$, then $\beta^*_0\gen{\s s_n[\s x]\xi}=f_{\s s}(\s x,\beta^*_0\gen\xi)$.
+\elemm
+
+\Proof this too, reduces to the same proof as for expressions without parentheses.
+\qed
+
+\thrm
+    For $\xi\in\exprp$, $\eval(\xi)=\beta^*_0\gen\xi$.
+\ethrm
+
+\Proof We induct on the length of $\xi$.
+If $\xi=\s x$ this is trivial.
+If $\xi=\s L_\infty\rho\s R_0\xi'$ for $\rho\in\exprp$ and so
+$$ \xi\tos\beta^*\gen{\s L_\infty\rho\s R_0}\xi' = \beta^*_0\gen\rho_\infty\xi' = \eval(\rho)\xi' $$
+where the final equality is due to induction.
+Then inductively, we have that $\beta^*_0\gen\xi=\beta^*_0\gen{\eval(\rho)\xi'}=\eval(\eval(\rho)\xi')$, which is just $\eval(\s L\rho\s R\xi')=\eval(\xi)$.
+And finally if $n$ is the lowest operator priority in $\xi$, then $\xi=\xi_1\s s_n^1\xi_2\cdots\xi_k\s s_n^k\xi_{k+1}$, this just reduces to the same proof for expressions without parentheses by the above
+two lemmas.
+\qed
+
 \bye

### 2024-10-06 Comparisons/comparer.py
Worked on Makefiles, README, comparisons

@@ -1,0 +1,25 @@
+import matplotlib.pyplot as plt
+import os
+import sys
+import time
+
+if len(sys.argv) < 2:
+    print("Must provide file name to run")
+    sys.exit(1)
+
+num_exes = 30
+
+exes = ["../llang", "../llang-classical"]
+exe_names = ["Our Algorithm", "Classical Algorithm"]
+times = []
+
+for exe in exes:
+    start = time.time()
+    for i in range(num_exes):
+        print(f"Run {i}")
+        os.system(f"{exe} {sys.argv[1]} y")
+    times.append(time.time() - start)
+
+fig = plt.figure(figsize = (10, 5))
+plt.bar(exe_names, times, color="red")
+plt.show()

### 2024-10-06 Comparisons/fib-compare.py

@@ -1,0 +1,40 @@
+import matplotlib.pyplot as plt
+import os
+import sys
+import time
+
+num_exes = 5
+exe_nums = [i for i in range(2,29,3)]
+
+exes = ["../llang", "../llang-classical"]
+
+fig = plt.figure(figsize = (10, 5))
+
+for exe in exes:
+    times = []
+    for n in exe_nums:
+        print(f"Run {n}")
+        code = f"""
+        fun fib (x) {{
+            if (x < 2) {{
+                1
+            }}{{
+                (fib (x-1)) + (fib (x-2))
+            }}
+        }}
+        let x = (fib {n});
+        _prim_print x
+        """
+        with open("fib.ml", "w") as f:
+            f.write(code)
+        start = time.time()
+        for i in range(num_exes):
+            os.system(f"{exe} fib.ml y")
+        end = time.time()
+        times.append(end - start)
+    with open("fib-res.txt", "a") as f:
+        f.write(f"{exe}: {' '.join([str(f) for f in times])}")
+    plt.plot(exe_nums, times)
+
+plt.show()
+

### 2024-10-07 Comparisons/comparer.py
Update README

@@ -7,7 +7,7 @@ if len(sys.argv) < 2:
     print("Must provide file name to run")
     sys.exit(1)
 
-num_exes = 30
+num_exes = 300
 
 exes = ["../llang", "../llang-classical"]
 exe_names = ["Our Algorithm", "Classical Algorithm"]
@@ -22,4 +22,6 @@ for exe in exes:
 
 fig = plt.figure(figsize = (10, 5))
 plt.bar(exe_names, times, color="red")
+for i in range(len(exes)):
+    print(f"{exe_names[i]}: {times[i] / num_exes}")
 plt.show()

### 2024-10-07 Comparisons/expr-compare.py

@@ -4,7 +4,7 @@ import sys
 import time
 
 num_exes = 5
-exe_nums = [i for i in range(2,26,3)]
+exe_nums = [i for i in range(2,6)]
 
 exes = ["../llang", "../llang-classical"]
 names = ["Our Algorithm", "Classical Algorithm"]
@@ -15,28 +15,17 @@ for e in range(len(exes)):
     times = []
     for n in exe_nums:
         print(f"Run {n}")
-        code = f"""
-        fun fib (x) {{
-            if (x < 2) {{
-                1
-            }}{{
-                (fib (x-1)) + (fib (x-2))
-            }}
-        }}
-        let x = (fib {n});
-        _prim_print x
-        """
-        with open("fib.ml", "w") as f:
-            f.write(code)
+        os.system(f"python3 ../Demos/create-expression.py {n} {n}")
         start = time.time()
         for i in range(num_exes):
-            os.system(f"{exes[e]} fib.ml y")
+            os.system(f"{exes[e]} expression.txt y")
         end = time.time()
         times.append(end - start)
-    with open("fib-res.txt", "a") as f:
+    with open("expr-res.txt", "a") as f:
         f.write(f"{names[e]}: {' '.join([str(f) for f in times])}")
     plt.plot(exe_nums, times, label=names[e])
 
 plt.legend()
 plt.show()
+
 

### 2024-10-08 Docs/linear-reduction.tex
Worked on Makefiles, README, comparisons
Updated documentation

@@
 \input pdfToolbox
+\input syntax-hi
 \input preamble
@@
 In this paper we will define the concept of linear reduction in the context of syntax parsing.
-We will progress through more and more complicated examples, beginning from the programming of a simple calculator until we ultimately have created an extensible programming language, which we call Lang-X.
+We will progress through more and more complicated examples, beginning from the programming of a simple calculator until we ultimately have created an extensible programming language, which we call L-Lang.
 
@@
 
-In this section we discuss the grammar of the language of Lang-X.
+In this section we discuss the grammar of the language of L-Lang.
 This is not naturally imposed by the parser, but it will properly parse programs of this form.
@@
 
-We now describe the initial beta reducer of Lang-X according to stateful reduction.
+We now describe the initial beta reducer of L-Lang according to stateful reduction.
 By convention, {\astyle atomic abstract terms} will be red, {\tstyle type terms} will be green, {\istyle internal terms} will be blue.
@@
 
+\vfill\break
+
+\section{Comparing Runtimes}
+
+For this paper, we developed an interpreter using both our algorithm and the tool menhir.
+We then compare runtimes for various different programs.
+
+\subsection{Currying}
+
+We begin with the following program:
+
+\beginhi
+fun print (x) {
+    _prim_print x
+}
+
+fun curry (f) {
+    fun curried (x) {
+        fun curriedX (y) {
+            f (x,y)
+        }
+        curriedX
+    }
+    curried
+}
+
+fun plus (x,y) {
+    x + y
+}
+
+print (plus (10, 20));
+let curry_plus = curry plus;
+print ((curry_plus 10) 20);
+\endhi
+
+We compare the time it takes our algorithm and the classical algorithm to run this program 300 times.
+
+\medskip
+\pdfximage width .5\hsize{Images/currying.png}
+\centerline{\pdfrefximage\the\pdflastximage}
+\medskip
+
+It takes our algorithm 0.0040886 seconds to run this program on average, and it takes the classical algorithm 0.0036926 seconds.
+
+\subsection{Expressions}
+
+We compare runtimes of programs of the form: {\tt let x = <expr>;}, where {\tt<expr>} is an arithmetical expression generated by the following algorithm:
+
+\medskip
+\algorithm
+\Comment Create an expression with $\ell$ subterms, with a parentheses depth of $d$, where numbers are chosen from\cr $[0,n)$.\EndComment
+\Function{Generate-Expression}{$\ell,d,n$}
+    \lIf{$d=0$} \vtop{\hsize=.8\hsize
+        \noindent \Return an expression with $\ell$ random numbers in $[0,n)$ and $\ell-1$ random operators in $\set{+,-,\times,\div}$.}
+    \State \Return $\bigl(\hbox{\tencsc Generate-Expression}(\ell,d-1,n)\bigr)\circ_1\cdots\circ_{\ell-1}\bigl(\hbox{\tencsc Generate-Expression}(\ell,d-1,n)\bigr)$
+    \nonum\State where $\circ_i\in\set{+,-,\times,\div}$ are chosen randomly
+\EndFunc
+\ealgorithm
+\medskip
+
+Running this with $n=100$, and we parameterize $\ell=d$, we get the following:
+
+\medskip
+\pdfximage width .5\hsize{Images/expr.png}
+\centerline{\pdfrefximage\the\pdflastximage}
+\medskip
+
+\subsection{Fibonacci}
+
+We compare runtimes of a fibonacci program, which computes the $n$th fibonacci number using its recursive formula, where $n$ varies.
+
+\medskip
+\pdfximage width .5\hsize{Images/fib.png}
+\centerline{\pdfrefximage\the\pdflastximage}
+\medskip
+
+Using exponential regression, we estimate that our algorithm takes $ab^n$ seconds, where $a=0.000138585$ and $b=1.64685$; and the classical algorithm takes $ab^n$ seconds where $a=0.00000362553$ and
+$b=1.5809$.
+
 \bye

### 2024-11-04 Paper/paper.tex
Added paper

\documentclass{llncs}

%\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{float}

\def\labelitemi{$\bullet$}
\def\labelenumi{(\textbf{\arabic{enumi}})}

\usepackage{tikz}
\usetikzlibrary{trees}

\newcounter{algo}

\floatstyle{ruled}
\newfloat{algo}{h}{aux}
\floatname{algo}{Algorithm}

\input preamble

\begin{document}

\title{Linear Reduction: Parsing and Interpreting Code Without Trees}
\author{Ari Feiglin\inst{1}\and Dr. Yoni Zohar\inst{1}}
\institute{University of Bar Ilan, Ramat Gan, Israel}

\maketitle

\begin{abstract}

    In this paper we discuss an alternative method to parsing and interpreting human-readable code.
    This method utilizes an iterative approach, as opposed to a recursive tree-oriented approach found in most modern compilers and interpreters.
    We will implement our algorithm in an interpreter and compare it with an interpreter created using more classical means.

    \keywords{Parsing \and Interpreting \and AST \and Parse Trees}

\end{abstract}

\section{Introduction}

Nowadays, most programming language interpreters follow the following pipeline:
\begin{enumerate}
    \item Lexing: lexically analyze the input code, separating it into atomic \textit{tokens}.
    \item Parsing: convert the parsed tokens into a \textit{parse tree}: a tree which represents the grammar of the code.
    \item Interpreting: recursively descend the tree and execute the code according to the instructions embedded in the tree.
\end{enumerate}
For example, we can parse the following expression:

\centerline{\texttt{1 + 2 * 3 + 4}}

\noindent as follows

\medskip
\centerline{\begin{tikzpicture}
\node{\texttt{+}}
child {node {\texttt{1}}}
child {node {\texttt{+}}
    child {node {\texttt{*}}
        child {node {\texttt{2}}}
        child {node {\texttt{3}}}
    }
    child {node {\texttt{4}}}
};
\end{tikzpicture}}
\medskip

One can then descend the parse tree and compute the expression.
Explicitly, if you reach a node whose operator is $\circ$, evaluate the left tree to get $x$, evaluate the right tree to get $y$, and then return $x\circ y$.

Unfortunately, constructing such a tree is not so easy.
Multiple methods for doing so exist, as well as tools for automating this process.

We will propose a novel method for the process of interpretation, which doesn't use parse trees.
Parsing code generated by context-free grammars (CFGs) is inherently a recursive process, and thus it is not surprising that our algorithm can be altered to instead produce parse trees.

\section{Notation}

\begin{enumerate}
    \item $\bN$ denotes the set of natural numbers, including $0$.
    \item $\bbN$ is defined to be $\bN\cup\set\infty$.
    \item $\bZ$ denotes the set of integer numbers.
    \item $\bbZ$ is defined to be $\bZ\cup\set{\pm\infty}$.
    \item $f\colon A\partial B$ means that $f$ is a partial function from $A$ to $B$ (i.e. $\operatorname{dom}f\subseteq A$).
    \item $\epsilon$ denotes the empty string.
    \item If $\mathcal L$ is a language (a set of strings), then $\mathcal L^\epsilon$ is equal to $\mathcal L\cup\set\epsilon$.
    \item A list is denoted by $(x_1,\dots,x_n)$.
    \item The concatenation of two lists $\ell_1$ and $\ell_2$ is denoted by $\ell_1@\ell_2$.
    \item $x::\ell$ is the list whose first element is $x$ and tail is $\ell$.
\end{enumerate}

\section{The Algorithm}

\subsection{A First Attempt -- Primitive Reduction}

The idea of the algorithm is as follows: given a string of tokens $\mb{\sigma^1}\cdots\mb{\sigma^n}$ (for readability, tokens are written in boxes), we form rules for combining two tokens.
For example $\mb{\ttt1}$ and $\mb{\ttt+}$ can be combined together to give a token $\mb{\ttt{1,+}}$.
Then if we follow a token of the form $\mb{\ttt{x,+}}$ with a token of the form $\mb{\ttt y}$, the tokens are combined to form a token $\mb{\ttt{x+y}}$.
So for example:
\[ \mbt{1} \mbt{+} \mbt{2} \mbt{+} \mbt{3} \to \mbt{1,+} \mbt2 \mbt+ \mbt3 \to \mbt{3} \mbt+ \mbt3 \to \mbt{3,+} \mbt3 \to \mbt6 \]
But what about \ttt{1+2*3}?
We'd get
\[ \mbt1 \mbt+ \mbt2 \mbt* \mbt3 \to \mbt{1,+} \mbt2 \mbt* \mbt3 \to \mbt3 \mbt* \mbt3 \to \mbt{3,*} \mbt3 \to \mbt9 \]

So to remedy this we add the concept of \textit{priorities}: for each symbol in our alphabet $\Sigma$, we assign a priority $p\in\bbN$.
This can be thought of as a \textit{priority function} $\pi\colon\Sigma\longto\bbN$.
So if the tokens $\mb{\sigma^1}\cdots\mb{\sigma^n}$ have respective priorities $p_1,\dots,p_n$ we can form the string $\mb{\sigma^1}_{p_1}\cdots\mb{\sigma^n}_{p_n}$ (where the subscript is part of the
character).
Now, we only combine two tokens if the token of the first is at least as great as the token of the second, and then the combined token gets the lower (the second) of the two priorities.
So we can assign to tokens containing numbers a priority of $\infty$, the token \ttt+ gets a priority of $1$, and $\ttt*$ gets a priority of $2$.
Then this becomes:
\[ \mbt1_\infty \mbt+_1 \mbt2_\infty \mbt*_2 \mbt3_\infty \to \mbt{1,+}_1 \mbt2_\infty \mbt*_2 \mbt3_\infty \to \mbt{1,+}_1 \mbt{2,*}_2 \mbt3_\infty \]
This cannot be combined any further, so we need another rule: we will give the final token a priority of $0$, so this becomes:
\[ \mbt{1,+}_1 \mbt{2,*}_2 \mbt3_0 \to \mbt{1,+}_1 \mbt{6}_0 \to \mbt{7}_0 \]
which is the correct result.

We can formalize this as follows: we start with an alphabet $\Sigma$.
Define $\oS=\Sigma\times\bbN$, and we define a so-called \textit{initial $\beta$-reducer}: $\hb\colon \Sigma\times\Sigma \longto \Sigma$.
This is a function which tells us how to combine tokens together.
And we extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.

Suppose $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$, then $\beta(\xi)$ is defined recursively as follows:
\begin{enumerate}
    \item if $n=1$, then $\beta(\xi)=\sigma^n_0$,
    \item if $p_1\geq p_2$, then $\beta(\xi)=\hb(\sigma^1,\sigma^2)_{p_2}\sigma^3_{p_3}\cdots\sigma^n_{p_n}$,
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}\,\beta(\sigma^2_{p_2}\cdots\sigma^n_{p_n})$.
\end{enumerate}

Notice that we need to add extra rules, like $\beta(\mbt{x,*},\mbt{y,+})=\mbt{x*y,+}$.
This is so we can handle
\begin{multline*}
    \mbt1_\infty \mbt*_2 \mbt2_\infty \mbt+_1 \mbt3_\infty \to \mbt{1,*}_2 \mbt2_\infty \mbt+_1 \mbt3_\infty \to \mbt{1,*}_2 \mbt{2,+}_1 \mbt3_\infty \to \mbt{2,+}_1 \mbt3_\infty\\
    \to \mbt{2,+}_1 \mbt3_0 \to \mbt5_0
\end{multline*}

But this has a major issue: it can't handle things like parentheses.
So we need to come up with a second attempt.

\subsection{A Second Attempt -- Partial Reduction}

In order to handle parentheses, all we need to do is to make a simple change to our algorithm.
Instead of a full function which also tells us how to compute the new priority, the initial $\beta$-reducer will be a partial function:
\[ \hb\colon \Sigma\times\Sigma\partial\Sigma\times(\bbZ\times\bbZ\to\bbZ) \]
Now we define the derived $\beta$-reducer as follows; it is a function $\beta\colon\oS^*\partial\oS^*$ where for $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$:
\begin{enumerate}
    \item if $n=1$ then $\beta(\xi)=\sigma^n_0$,
    \item if $p_1\geq p_2$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho)$ is defined, then $\beta(\xi)=\tau_{\rho(p_1,p_2)}\sigma^3_{p_3}\cdots\sigma^n_{p_n}$,
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}\,\beta(\sigma^2_{p_2}\cdots\sigma^n_{p_n})$.
\end{enumerate}

Now we define our initial $\beta$-reducer which will support parentheses.
It is an extension of our previous $\beta$-reducer:
\begin{gather*}
    \hb(\mbt x,\mb\circ) = (\mb{\ttt x,\circ},\snd),\quad \hb(\mb{\ttt x,\circ},\mbt y) = (\mb{\ttt x\circ\ttt y},\snd) \qquad \hbox{for $\circ\in\set{\ttt{+,*}}$} \\
    \hb(\mbt{x,*},\mbt{y,+}) = (\mbt{x*y,+},\snd)
\end{gather*}
where $\fst,\snd$ are the first and second projections, respectively (i.e. $(x,y)\mapsto x$ and $(x,y)\mapsto y$).
We then add the rules for parentheses:
\begin{gather*}
    \hb(\mbt x,\mbt)) = (\mbt{x,)},\snd),\quad \hb{\mb{\ttt x,\circ},\mbt{y,)}} = (\mb{\ttt x\circ\ttt y,\ttt)},\snd) \qquad \hbox{for $\circ\in\set{\ttt{+,*}}$} \\
    \hb(\mbt(,\mbt{x,)}) = (\mbt x,\fst)
\end{gather*}
\gobble) % syntax highlighting
And we define our priority function as follows:
\[ \pi(\mbt x) = \infty\hbox{ for numbers \ttt x},\quad \pi(\mbt+) = 1,\quad \pi(\mbt*) = 2,\quad \pi(\mbt() = \infty,\quad \pi(\mbt)) = 0 \]

Now take for example \ttt{2*(1+3)+4}:
\begin{align*}
    \mbt2_\oo \mbt*_2 \mbt(_\oo \mbt1_\oo \mbt+_1 \mbt3_\oo \mbt)_0 \mbt+_1 \mbt4_\oo &\to \mbt{2,*}_2 \mbt(_\oo \mbt1_\oo \mbt+_1 \mbt3_\oo \mbt)_0 \mbt+_1 \mbt4_\oo\\
        &\to \mbt{2,*}_2 \mbt(_\oo \mbt{1,+}_1 \mbt3_\oo \mbt)_0\mbt+_1 \mbt4_\oo\\
        &\to \mbt{2,*}_2 \mbt(_\oo \mbt{1,+}_1 \mbt{3,)}_0 \mbt+_1 \mbt4_\oo\\
        &\to \mbt{2,*}_2 \mbt(_\oo \mbt{4,)}_0 \mbt+_1 \mbt4_\oo\\
        &\to \mbt{2,*}_2 \mbt4_\oo \mbt+_1 \mbt4_\oo\\
        &\to \mbt{2,*}_2 \mbt{4,+}_1 \mbt4_\oo\\
        &\to \mbt{8,+}_1 \mbt4_\oo\\
        &\to \mbt{8,+}_1 \mbt4_0\\
        &\to \mbt{12}_0
\end{align*}
\gobble)\gobble)

Notice that our $\beta$-reducers don't care about the exact value of the number and operation: both $\mbt1\mbt+$ and $\mbt2\mbt*$ are mapped to something of the form $\mbt{x,$\circ$}$.
We can thus abstract away the exact value of tokens and have our initial $\beta$-reducer deal only with the form of the token.
We do such in our third attempt.

\subsection{A Third Attempt -- Valued Reduction} \label{valued-reduction}

Here we will abstract away the value of each term; the value of the new term given by the initial $\beta$-reducer will be determined by an output function.
Our alphabet of types $\Sigma$, instead of having a token for every number, will just have tokens for types (number, operator, etc.).
We will also have another set $\values$, the \textit{universe of values}.

The initial $\beta$-reducer is a function
\[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
Our alphabet of types is $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$, define $\oP=\oS\times\values$ to be the set of all possible terms (which is a type, a priority, and a value).
Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the type; $p$ is the priority; and $v$ the value.

Our rules for the derived $\beta$-reducer is similar to before: it is a function $\beta\colon\oP^* \partial \oP^*$ where for $\xi=\sigma^1_{p_1}(v_1)\cdots\sigma^n_{p_n}(v_n)$
\begin{enumerate}
    \item if $n=1$ then $\beta(\xi)=\sigma^n_0(v_n)$,
    \item if $p_1\geq p_2$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho,f)$ is defined, then
    \[ \beta(\xi)=\tau_{\rho(p_1,p_2)}(f(v_1,v_2))\sigma^3_{p_3}(v_3)\cdots\sigma^n_{p_n}(v_n) \]
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}(v_1)\beta(\sigma^2_{p_2}(v_2)\cdots\sigma^n_{p_n}(v_n))$.
\end{enumerate}

Our initial $\beta$-reducer will is defined as follows:
\begin{itemize}
    \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f\mapsto(v,f))$
    \item $\hb(\OP\NUM,\NUM)=(\NUM,\snd,((v,f),u\mapsto f(v,u)))$
    \item $\hb(\OP\NUM,\OP\NUM)=(\OP\NUM,\snd,((v,f),(u,g)\mapsto (f(v,u),g))$
    \item $\hb(\NUM,\RPAREN)=(\RPAREN\NUM,\snd,(v,\_\mapsto v))$
    \item $\hb(\OP\NUM,\RPAREN\NUM)=(\RPAREN\NUM,\snd,((v,f),u\mapsto f(v,u)))$
    \item $\hb(\LPAREN,\RPAREN\NUM)=(\NUM,\fst,(\_,v\mapsto v))$
\end{itemize}

But this of course is not capable of handling the semantics necessary for programming languages.
We need a notion of a state, and a somewhat more refined algorithm.

\subsection{A Fourth Attempt -- Stateful Reduction}

We define the following four base sets:
\begin{enumerate}
    \item $\values$ the \textit{universe of values}; these are all the internal values an object may have.
    \item $\termp$ the set of \textit{printable terms}; this is the set of all printable tokens a programmer may type.
    \item $\terms$ the set of \textit{type terms}; these are terms which correspond to types in the classical sense.
    \item $\terma$ the set of \textit{abstract terms}; these are intermediate terms (like $\OP\NUM$ from the previous section).
\end{enumerate}
The sets $\termp,\terms,\terma$ are disjoint, and we place no such restriction on $\values$.

While we don't need to place any further restrictions on $\termp,\terms,\terma$, we found it useful to explicitly construct them.
To construct $\terms$, we begin with a set of \textit{base types} $\Sigma$.
Each base type has an associated arity in $\bbN$; let $\Sigma^n$ denote the set of all base types of arity $n$.
Then we define $\terms$ as follows:
\[ \terms \ccoloneqq \Sigma^0 \mid \Sigma^1\terms \mid \Sigma^2\terms\terms \mid \Sigma^3\terms\terms\terms \mid \cdots \mid \Sigma^\oo(\terms\cdots\terms) \]

To construct $\terma$ we start with $\mathcal A$, a set of \textit{atomic abstract terms}, again each with an arity in $\bbN$.
$\mathcal A^n$ is the set of atomic abstract terms of a set arity $n$.
Then $\terma$ is defined by
\[ \terma \ccoloneqq \mathcal A^0\mid\mathcal A^1\terms^\epsilon\mid\mathcal A^2\terms^\epsilon\terms^\epsilon\mid\mathcal A^3\terms^\epsilon\terms^\epsilon\terms^\epsilon\mid\cdots\mid
\mathcal A^\infty\terms^\epsilon\cdots\terms^\epsilon \]
Here we allow inputs to atomic abstract terms to be empty (i.e. $\terms^\epsilon$ instead of just $\terms$).

So if $\Sigma=\set{\NUM_0,\LIST_1,\PRODUCT_\infty}$ (the subscripts are the arities of each base type) and $\mathcal A=\set{\OP}$, then
\[ \NUM,\LIST\NUM,\PRODUCT(\LIST\NUM\NUM)\in\terms,\qquad \OP\LIST\NUM\in\terma \]
For readability we can put parentheses and commas where necessary, but importantly these sets have the \textit{unique reconstruction property}: it is possible to compute their construction from the
string itself (i.e. a string cannot be constructed from two different patterns).

Now we define the following sets:
\begin{enumerate}
    \item $\term=\termp\cup\terms\cup\terms$, the set of \textit{terms},
    \item $\termi=\terms\cup\terma$, the set of \textit{internal terms},
    \item $\expi=\termi\times\values$, the set of \textit{internal expressions},
    \item $\exp=\termi\cup\termp$, the set of \textit{expressions}.
\end{enumerate}
Finally we define $\oe=\exp\times\bbZ$.
Elements of $\oe$ will be written like $\sigma_n(v)$ as before; for printable terms we act as if the value is empty.

The \textit{initial priority function} is a function $\pi\colon\termp\longto\bbZ$.
This can be naturally extended to $\pi\colon\termp^*\longto\oe^*$.

Finally, our \textit{initial $\beta$-reducer} is a function ($\state$ is a set of functions $\termp\partial\expi$ which we will define later)
\[ \hb\colon\termi\times\term^\epsilon \partial \termi^\epsilon\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\times\state\partial\values\times\termp^*\times\state) \]
and we extend this to a \textit{derived $\beta$-reducer}, a function
\[ \beta\colon\oe^*\times\state \partial \oe^*\times\state \]
inputs in $\oe^*\times\state$ will be written as $\gen\xi[\bar s]$ for $\xi\in\oe^*$ and $\bar s\in\state$.
We define $\beta^*\gen\xi[\bar s]$ to be the value obtained by composing $\beta$ with itself of $\gen\xi[\bar s]$ until the convergence of $\xi$.
Meaning
\begin{multline*}
    \beta^* = \beta^n\gen\xi[\bar s]\\
    \hbox{where $n=\min\set{n\geq1}[{\beta^n\gen\xi[\bar s]=\gen{\xi'}[\bar s'],\beta\gen{\xi'}[\bar s']=\gen{\xi'}[\bar s'']}]$}
\end{multline*}
We define $\beta\gen\xi[\bar s]$ as follows, where $\xi=\sigma^1_{p_1}(v_1)\cdots\sigma^n_{p_n}(v_n)$.
\begin{enumerate}
    \item if $\sigma^1\in\termp$ then
    \[ \beta\gen\xi[\bar s] = \gen{s(\sigma^1)_{p_1}\sigma^2_{p_2}(v_2)\cdots\sigma^n_{p_n}(v_n)}[\bar s] \]
    \item if $\hb(\sigma^1,\epsilon)=(\tau,\rho,f)$ is defined, then compute $f(v_1,\_,s)=(u,\xi',s')$ and $\beta^*\gen{\pi\xi'}[\bar s']=\gen{\xi''}[\bar s'']$, and
    \[ \beta\gen\xi[\bar s] = \gen{\tau_{\rho(p_1,\_)}\xi''\sigma^2_{p_2}(v_2)\cdots\sigma^n_{p_n}(v_n)}[\bar s''] \]
    \item if $p_1\geq p_2$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho,f)$ is defined, then compute $f(v_1,v_2,s)=(u,\xi',s')$ and $\beta^*\gen{\pi\xi'}[\bar s']=\gen{\xi''}[\bar s'']$, and
    \[ \beta\gen\xi[\bar s] = \gen{\tau_{\rho(p_1,p_2)}\xi''\sigma^3_{p_3}(v_2)\cdots\sigma^n_{p_n}(v_n)}[\bar s''] \]
    \item otherwise compute $\beta\gen{\sigma^2_{p_2}(v_2)\cdots\sigma^n_{p_n}(v_n)}[\bar s]=\gen{\xi'}[\bar s']$ and
    \[ \beta\gen\xi[\bar s] = \gen{\sigma^1_{p_1}(v_1)\xi'}[\bar s'] \]
\end{enumerate}

\subsubsection{States}

We begin by defining \textit{point-states} to be partial maps $\termp\partial\expi$.
If $s_1,s_2$ are two point-states then we define their composition to be
\[ s_1s_2(\sigma) = \begin{cases} s_2(\sigma) & \sigma\in\dom s_2\\ s_1(\sigma) & \sigma\in\dom s_1-\dom s_2\end{cases} \]
Point-states with finite domain will be denoted by $[\sigma_i\mapsto\varkappa_i]_{i=1}^n$ where $\sigma_1,\dots,\sigma_n$ is the domain and $\varkappa_1,\dots,\varkappa_n$ are their respective images.

A \textit{state} will be a pair of a sequence of point-states and a set of indices.
This will be denoted by $\bar s=[(s_1,\dots,s_n),\;I=(i_1,\dots,i_k)]$ (we require that $k\leq n$).
We define
\begin{enumerate}
    \item $\bar s+s=\bigl[(s_1,\dots,s_n,s),\;I\bigr]$, for $s$ a point-state,
    \item $\bar s+_cs=\bigl[(s_1,\dots,s_n,s),\;(i_1,\dots,i_k,n+1)\bigr]$, for $s$ a point-state,
    \item $\pop\bar s=\bigl[(s_1,\dots,s_{n-1}),\;I\bigr]$ if $i_k<n$, otherwise it is defined to be equal to $\bigl[(s_1,\dots,s_{n-1}),\;(i_1,\dots,i_{k-1})\bigr]$,
    \item $\bar ss=\bigl[(s_1,\dots,s_{n-1},s_ns),\;I\bigr]$ for $s$ a point-state,
    \item $\bar s(\sigma)=s_1\cdots s_n(\sigma)$ for $\sigma\in\termp$,
    \item $\bar s_c=s_{i_k}\cdots s_n$,
    \item if $\sigma\in\termp$ and $\varkappa\in\expi$ then $\bar s\{\sigma\mapsto\varkappa\}=\bigl[(s_1,\dots,s_k[\sigma\mapsto\varkappa],\dots,s_n),\;I\bigr]$ where $k$ is the maximum index such that
    $\sigma\in\dom s_k$.
\end{enumerate}

\section{Implementing the Algorithm}

We will implement our algorithm in an interpreter for a toy language we dubbed ``LLang''.
We will also create an interpreter for LLang using menhir\NEEDCITE{}.
In this section we will detail the implementation of our algorithm and compare and contrast the two implementations.

\subsection{The Grammar}

The grammar for our toy language is pretty straight-forward:

\medskip
\grammar
\noalign{\hbox{\bf Identifiers}}
    letter&\ccoloneqq\(a\dots z\gbar A\dots Z\gbar\_\)\cr
    digit&\ccoloneqq\(0\dots9\)\cr
    non\_digit&\ccoloneqq\(\_\gbar{\it letter}\)\cr
    ident&\ccoloneqq{\it letter}\ \({\it digit}\gbar{\it non\_digit}\)^*\cr
    pident&\ccoloneqq {\bf\_prim\_} \({\it digit}\gbar{\it non\_digit}\)^*\cr
\noalign{\hbox{\bf Constant Expressions}}
    const&\ccoloneqq\({\it number}\gbar{\it product}\)\cr
    number&\ccoloneqq\({\it digit}\)^*[.\({\it digit}\)^*]\cr
\noalign{\hbox{\bf Operators}}
    op&\ccoloneqq\(@\gbar+\gbar-\gbar*\gbar/\gbar{<=}\gbar{<}\gbar{>=}\gbar{>}\gbar{==}\gbar{!}{=}\)\cr
\noalign{\hbox{\bf Expressions}}
    expr&\ccoloneqq\epsilon\cr
        &\bbar{\bf let}\ {\it ident}\ =\ {\it value}\ ;\ {\it expr}\cr
        &\bbar{\bf fun}\ {\it ident}\ (\ {\it plist}\ )\ \{\ {\it expr}\ \}\ {\it expr}\cr
        &\bbar{\bf if}\ (\ {\it value}\ )\ \{\ {\it expr}\ \}\ \{\ {\it expr}\ \}\ {\it expr}\cr
        &\bbar{\it funcall}\ ;\ {\it expr}\cr
        &\bbar{\it pident}\ {\it value}\ {\it expr}\cr
        &\bbar{\it value}\cr
\noalign{\hbox{\bf Values}}
    value&\ccoloneqq{\it const}\cr
        &\bbar{\it ident}\cr
        &\bbar{\it value}\ {\it op}\ {\it value}\cr
        &\bbar{\it funcall}\cr
        &\bbar(\ {\it product}\ )\cr
    product&\ccoloneqq{\it value}\cr
        &\bbar{\it value}\ ,\ {\it product}\cr
\noalign{\hbox{\bf Function Calls}}
    plist&\ccoloneqq{\it ident}\cr
        &\bbar{\it ident}\ ,\ {\it plist}\cr
    funcall&\ccoloneqq{\it value}\ {\it value}\cr
\egrammar

This basic grammar will suffice for our classical implementation.
For our algorithm, we will add more features (such as lists and nested products).

\subsection{The Initial $\beta$-Reducer}

Our algorithm is based on an initial $\beta$-reducer instead of a grammar.
Thus in order to implement our algorithm we must first form an initial $\beta$-reducer.
So we must translate our previous section to an initial $\beta$-reducer.

Instead of explicitly writing the sets $\mathcal A$ and $\Sigma$, base types will be underlined, atomic abstract terms will be written normally, and internal terms will be boxed.
General terms will be denoted by $\sigma,\tau$, and lists of terms will be denoted by $\Omega$.

\noindent{\bf End}:
\blist
    \item $\isig\ \eend \longto \isig\ \minfty\ (u,\_,s\to u,\epsilon,s)$
\elist

\noindent{\bf Arithmetic}:
\blist
    \item $\tsig\ \op \longto \op\tsig\ \snd\ (u,f,s\to(u,f),\epsilon,s)$
    \item $\op\tsig\ \op\tsig \longto \op\tsig\ \snd\ \bigl((u,f),(v,g),s\to (f(u,v),g),\epsilon,s\bigr)$
    \item $\op\tsig\ \tsig \longto \tsig\ \snd\ \bigl((u,f),v,s\to f(u,v),\epsilon,s\bigr)$
    \item $\Pop\ \tsig \longto \tsig\ \snd\ \bigl((f,g),u,s\to f(u),\epsilon,s\bigr)$
    \item $\tsig\ \Pop \longto \op\tsig\ {\sf one}\ \bigl(u,(f,g),s\to(u,g),\epsilon,s\bigr)$
    \item $\tsig\ \rparen \longto \rparen\tsig\ \snd\ (u,\_,s\to u,\epsilon,s)$
    \item $\op\tsig\ \rparen\tsig \longto \rparen\tsig\ \snd\ \bigl((f,u),v,s\to f(u,v),\epsilon,s\bigr)$
    \item $\Pop\ \rparen\tsig \longto \rparen\tsig\ \snd\ \bigl((f,g),u,s\to f(u),\epsilon,s\bigr)$
    \item $\lparen\ \rparen\tsig \longto \tsig\ \fst\ (\_,u,s\to u,\epsilon,s)$
\elist

\noindent{\bf Lists}:
\blist
    \item $\Lbrack\ \tsig \longto \Lbrack\tsig\ \fst\ (\_,u,s\to (u),\epsilon,s)$
    \item $\Lbrack\tsig\ \tsig \longto \Lbrack\tsig\ \fst\ (\ell,u,s\to (\ell,u),\epsilon,s)$
    \item $\Lbrack\tsig\ \Rbrack \longto \List\tsig\ \pinfty\ (\ell,\_,s\to\ell,\epsilon,s)$
    \item $\period\ \num \longto \index\ \zero\ (\_,n,s\to n,\epsilon,s)$
    \item $\List\tsig\ \index \longto \tsig\ \fst\ (\ell,i,s\to \ell_i,\epsilon,s)$
\elist

\noindent{\bf Variables}:
\blist
    \item $\lett\ x \longto \letvar\ \snd\ (\_,\_,s\to(x,\varnothing),\epsilon,s)$
    \item $\letvar\ \index \longto \letvar\ \fst\ \bigl((x,\ell),n,s\to(x,(\ell,n)),\epsilon,s\bigr)$
    \item $\letvar\ \equal \longto \leteq\ \minfty\ \bigl((x,\ell),\_,s\to(x,\ell),\epsilon,s\bigr)$
    \item $\leteq\ \isig \longto \epsilon\ \varnothing\ \bigl((x,\ell),v,s\to\epsilon,\epsilon,s'\bigr)$ where $s'$ is $s[x\mapsto\sigma(v)]$ if $\ell=\varnothing$ and otherwise let $t$ be the result of
        setting $s(x).\ell_1.\dots.\ell_n$ to $v$, then $s'=s[x\mapsto t]$.
\elist

\noindent{\bf Scoping}:
\blist
    \item $\Lbrace\ \epsilon \longto \epsilon\ \varnothing\ (\_,\_,s\to\epsilon,\epsilon,s+\varnothing)$
    \item $\Rbrace\ \epsilon \longto \epsilon\ \varnothing\ (\_,\_,s\to\epsilon,\epsilon,\pop s)$
\elist

\noindent{\bf Products}:
\blist
    \item $\tsig\ \comma \longto \comma(\tsig)\ \snd\ (u,\_,s\to(u),\epsilon,s)$
    \item $\op\tsig\ \comma(\tsig) \longto \comma(\tsig)\ \snd\ ((f,u),(v)\to (f(u,v)),\epsilon,s)$
    \item $\Pop\ \comma(\tsig) \longto \comma(\tsig)\ \snd\ ((f,g),(u)\to (f(u),\epsilon,s))$
    \item $\comma\tOmeg\ \comma(\tsig) \longto \comma(\tOmeg,\tsig)\ \snd\ (\ell,\ell',s\to(\ell,\ell'),\epsilon,s)$
    \item $\comma\tOmeg\ \rparen\tsig \longto \Listrparen(\tOmeg,\tsig)\ \snd\ (\ell,v\to (\ell,v),\epsilon,s)$
    \item $\lparen\ \Listrparen\tOmeg \longto \product\tOmeg\ \pinfty\ (\_,\ell,s\to \ell,\epsilon,s)$
\elist

\noindent{\bf Primitives}:
\blist
    \item $\primitive\ \isig \longto \epsilon\ \varnothing\ (f,v,s\to \epsilon,w,s)$ where $f(\isig,v)=(w,s')$ (the purpose is for $f$ to have a side effect)
\elist

\noindent{\bf Code Capture}
\blist
    \item $\Lbrace\a\ x\longto \Lbrace\a\ \pinfty\ (\xi,\_,s\to\xi x,\epsilon,s)$ if $x\neq\{,\}$
    \item $\Lbrace\a\ x\longto \code\ \pinfty\ (\xi,\_,s\to\xi,\epsilon,s)$
    \item $\Lbrace\a\ \code\longto \Lbrace\a\ \pinfty\ (\xi,\xi',s\to\xi\{\xi'\},\epsilon,s)$
\elist

\noindent{\bf Parameter Capture}
\blist
    \item $\lparen\a\ x \longto \lparen\a\ \fst\ (\ell,\_,s\to \ell @(x),\epsilon,s)$ for $x\neq(,)$
    \item $\lparen\a\ ) \longto \plist\ \fst\ (\ell,\_,s\to \ell,\epsilon,s)$
    \item $\lparen\a\ \plist \longto \lparen\a\ \fst\ (\ell,\ell',s\to (\ell @(\ell')),\epsilon,s)$
\elist

\noindent{\bf Function Definitions}
\blist
    \item $\fun\ x\longto \funname\ \pinfty\ (\_,\_,s\to(x,\epsilon), \epsilon, s + [\{\mapsto\Lbrace\a,\}\mapsto\Rbrace\a,(\mapsto\lparen\a,)\mapsto\rparen\a])$
    \item $\funname\ \plist \longto \funvars\ \pinfty\ \bigl((x,\epsilon),u,s \to (x,u),\epsilon,s\bigr)$
    \item $\funvars\ \code \longto$\hfil\break\null\hfill$\closure\ \fst\ \bigl((x,\ell),\xi,s\to C=\gen{\ell,\xi,s'[x\mapsto\closure(C)]},\epsilon,\pop s[x\mapsto\closure(C)]\bigr)$ where $s'=(\pop s)_c$.
\elist

\noindent{\bf Function Calls}
\blist
    \item $\closure\ \tsig \longto \epsilon\ \varnothing\ \bigl(\gen{\ell,\xi,{\it ps}},u,s\mapsto\epsilon,\xi\},s+_c{\it ps}[\ell\mapsto\tsig(u)]\bigr)$ where $\ell\mapsto\tsig(u)$ means that if
        $\ell=(x)$ then $x\mapsto \tsig(u)$.
        Otherwise $\ell=(x_1,\dots,x_n)$, $\tsig=\product\tsig_1\cdots\tsig_n$, and $u=(u_1,\dots,u_n)$ and $x_i\mapsto\tsig_i(u_i)$ (recursively).
\elist

\noindent{\bf If Statements}
\blist
    \item $\iif\ \tsig \longto \ifbool\ \fst\ (\_,n,s\to n,\epsilon,s+[\{\mapsto\Lbrace\a,(\mapsto\lparen\a)]$
    \item $\ifbool\ \code \longto \ifthen\ \fst\ (n,\xi,s\to(n,\xi),\epsilon,s)$
    \item $\ifthen\ \code \longto \epsilon\ \_\ ((n,\xi_1),\xi_2,s\to\varnothing,(n=0?\ \xi_2:\xi_1),\pop s)$
\elist

\noindent{\bf Types}
\blist
    \item $\tp\tsig\ \tp\ttau \longto \tp\tsig(\ttau)\ \snd\ (\_,\_,s\to \tsig(\ttau),\epsilon,s)$
    \item $\tp\tsig\ \product(\tp\ttau_1,\dots,\tp\ttau_n) \longto$\hfil\break\null\hfill$\tp\tsig(\ttau_1,\dots,\ttau_n)\ \snd\ (\_,\_,s\to \tsig(\ttau_1,\dots,\ttau_n),\epsilon,s)$
    \item $\ccolon\ \tp\tsig \longto \typer\tsig\ \snd\ (\_,u,s\to u,\epsilon,s)$
    \item $\tsig\ \typer\ttau \longto \ttau\ \snd\ (u,\_,s\to u,\epsilon,s)$
\elist

\subsection{The Initial State}

The initial state is a partial state, defined as follows:

\bigskip

\hbox{\vtop{\hsize=.48\hsize
\noindent{\bf End}
\blist
    \item $;{}\mapsto(\eend,\varnothing)$
\elist

\noindent{\bf Arithmetic}
\blist
    \item $({}\mapsto(\lparen,\varnothing)$
    \item $){}\mapsto(\rparen,\varnothing)$
    \item $+{}\mapsto(\op, (n,m\to n+m))$
    \item $+{}\mapsto(\op, (n,m\to n+m))$
    \item $*{}\mapsto(\op, (n,m\to n*m))$
\elist
}\hfil
\vtop{\hsize=.48\hsize
\blist
    \item $/{}\mapsto(\op, (n,m\to n/m))$
    \item $-{}\mapsto$\hfil\break\null\hfill$(\Pop, (n\to-n), (n,m\to n-m))$
    \item $@{}\mapsto(\op, (\ell_1,\ell_2\to\ell_1@\ell_2))$
    \item $!={}\mapsto(\op,(u,v\to u\neq v))$
    \item $<={}\mapsto(\op,(n,m\to n\leq m))$
    \item $>={}\mapsto(\op,(n,m\to n\geq m))$
    \item $=={}\mapsto(\op,(u,v\to u=v))$
\elist}}
\vfill\break

{\vtop{\vsize=.48\hsize
}}

\hbox{\vtop{\hsize=.48\hsize
\blist
    \item $<{}\mapsto(\op,(n,m\to n<m))$
    \item $>{}\mapsto(\op,(n,m\to n>m))$
\elist

\noindent{\bf Lists}
\blist
    \item $[{}\mapsto(\Lbrack,[])$
    \item $]{}\mapsto(\Rbrack,\varnothing)$
    \item $.\mapsto(\period,\varnothing)$
\elist

\noindent{\bf Variables}
\blist
    \item ${\rm let}\mapsto(\lett,\varnothing)$
    \item $={}\mapsto(\equal,\varnothing)$
\elist

\noindent{\bf Scoping}
\blist
    \item $\{{}\mapsto(\Lbrace,\varnothing)$
    \item $\}{}\mapsto(\Rbrace,\varnothing)$
\elist

\noindent{\bf Products}
\blist
    \item $,{}\mapsto(\comma,\varnothing)$
\elist}\hfil

\vtop{\hsize=.48\hsize
\noindent{\bf Primitives}
\blist
    \item ${\rm\_prim\_print}\mapsto$\hfil\break
        \null\kern-1.6pt$(\primitive,(a,v\to{\rm print}(v); (\varnothing,\varnothing)))$
    \item ${\rm\_prim\_len}\mapsto$\hfil\break\null\hfill$(\primitive,(a,\ell\to\num,\abs\ell))$
    \item ${\rm\_prim\_tail}\mapsto$\hfil\break\null\hfill$(\primitive,(\tsig,t::\ell\to\tsig,\ell))$
    \item ${\rm\_prim\_type}\mapsto$\hfil\break\null\hfill$(\primitive,(\tsig,\_\to\tp\tsig,\tsig))$
\elist

\noindent{\bf Keywords}
\blist
    \item ${\rm fun}\mapsto(\fun,\varnothing)$
    \item ${\rm if}\mapsto(\iif,\varnothing)$
\elist

\noindent{\bf Types}
\blist
    \item $:{}\mapsto(\colon,\varnothing)$
    \item ${\rm Num}\mapsto(\tp\ \num,\num)$
    \item ${\rm List}\mapsto(\tp\ \List,\List)$
    \item ${\rm Closure}\mapsto(\tp\ \closure,\closure)$
    \item ${\rm Product}\mapsto(\tp\ \product,\product)$
    \item ${\rm Primitive}\mapsto$\hfil\break\null\hfill$(\tp\ \primitive,\primitive)$
    \item ${\rm Type}\mapsto(\tp\ \tp, \tp)$
\elist
}}

\section{Proving Correctness}

\def\eval{\hbox{\tencsc eval}}
\def\unparen{\hbox{\tencsc unparen}}

In this section we will show that valued reduction (\S\ref{valued-reduction}) computes arithmetical expressions correctly.
To do so, we will show that valued reduction computes expressions and gives a result equal to a na\"\i ve algorithm.

But first, we must define the problem.
We are given the following:
\begin{enumerate}
    \item a set $\s X$ which is our \textit{universe} (e.g. numbers),
    \item a set $\s S$ of \textit{operator symbols},
    \item for every $\s s\in\s S$ a \textit{priority} $\pi(\s s)\in\bbN$,
    \item for every $\s s\in\s S$ a function $f_{\s s}\colon\s X\times\s X\longto\s X$.
\end{enumerate}
We also take two symbols $\s L$ and $\s R$ for the left and right parentheses.
$\s X,\s S,\s L,\s R$ will also be our symbols for the $\beta$-reduction.

Expressions are defined by the following grammar:
\[ {\it expr} \ccoloneqq \s L\ {\it expr}\ \s R  \gbar {\it expr}\ \s S\ {\it expr} \gbar \s X \]
Operators are assumed to be left-associative.

Let us define $\unparen(\xi,i)$ to mean that the $i$th character in $\xi$ is not within parentheses (this can be implemented easily).
We define the \eval{} algorithm (see algorithm \ref{alg:eval}) to evaluate strings in ${\it expr}$.

\begin{algo}\label{alg:eval}
\caption{\eval uating arithmetical expressions}
\algorithm
\Function{eval}{$\xi$}
    \If{$\xi=\s x\in\s X$}
        \State\Return $\s x$
    \ElseIf{$\xi[0]=\s L$}
        \State $\xi$ is of the form $\s L\rho\s R\xi'$
        \State\Return $\eval(\eval(\rho)\xi')$
    \Else
        \Comment Find the greatest index with the smallest operator.\cr
        Note that $\max_i{\rm argmin}_if(i)$ is the maximal $i$ which minimizes $f$.\EndComment
        \State $i\coloneqq\max_i{\rm argmin}_i\set{\pi(\xi[i])}[{\xi[i]\in\s S,\unparen(\xi,i)}]$
        \State $\s s\coloneqq\xi[i]$
        \Comment $\xi'=\xi[:i-1]$, $\xi''=\xi[i+1:]$ \EndComment
        \State $\xi$ is of the form $\xi'\s s\xi''$
        \State\Return $f_{\s s}(\eval(\xi'),\eval(\xi''))$
    \EndIf
\EndFunc
\ealgorithm
\end{algo}

As opposed to pure valued (stateless) reduction, we will adopt some conventions from stateful reduction.
Since our string is a string in $(\set{\s L,\s R}\cup\s X\cup\s S)^*$ and not $\set{\s L,\s R,\s X,\s S}^*$ (note the difference!), we must first convert the string into one admissable for
$\beta$-reduction.
To do so, we use stateful reduction and use the initial state and priority map which maps elements $\s x\in\s X$ to $\s X_\infty(\s x)$, $\s S\in\s S$ to $\s S_{\pi(\s s)}(f_{\s s})$, $\s L$ to
$\s L_\infty$, and $\s R$ to $\s R_0$.
Since our state doesn't change (our initial $\beta$-reducer will not affect the state) we can assume that our string has been mapped to its associated $\oe$-string from the start.
That is, instead of strings like \ttt{1+2}, we can assume that our input strings are of the form $\s X_\infty(1)\ \s S_1(+)\ \s X_\infty(2)$.

Secondly, we will adopt the convention of having type terms, abstract terms, etc.
Our base type will be $\Sigma=\set{\s X}$, and our atomic abstract terms $\mathcal A=\set{\s L_0,\s R_1,\s S_1}$ (printable terms are $\set{\s L,\s R}\cup\s X\cup\s S$, but as said before we assumed these
are already altered by the state).

Our initial $\beta$-reducer will be
\begin{itemize}
    \item $\s X\ \s S\longto\s S\s X\ \snd\ (u,f\to(u,f))$
    \item $\s S\s X\ \s S\s X \longto\s S\s X\ \snd\ ((u,f),(v,g)\to(f(u,v),g))$
    \item $\s S\s X\ \s X\longto\s X\ \snd\ ((u,f),v\to f(u,v))$
    \item $\s X\ \s R \longto \s R\s X\ \snd\ (u,\_\to u)$
    \item $\s S\s X\ \s R\s X \longto\s R\s X\ \snd\ ((u,f),v\to f(u,v))$
    \item $\s L\ \s R\s X \longto\s X\ \fst\ (\_,u\to u)$
\end{itemize}

\subsection{The Proof}

We can rewrite the grammar ${\it expr}$ with priorities as follows:
\[ {\it expr} \ccoloneqq \s X_\infty \gbar \s X_\infty\ \s S_n\ {\it expr} \gbar \s L_\infty\ {\it expr}\ \s R_0 \gbar \s L_\infty\ {\it expr}\ \s R_0\ \s S_n\ {\it expr} \]
when considering an input for $\eval$, we will of course neglect the priorities in the string.
And when considering an input for $\beta$-reduction, we will act as if each character is of the term indicated with the value in the string (i.e. $1_\infty+_12_\infty\in{\it expr}$ is
$\s X_\infty(1)\s S_1(+)\s X_\infty(2)$).

Let us define the set of {\it intermediate expressions} as
\begin{multline*}
    {\it intexpr} \ccoloneqq \s X_n\gbar \s S\s X_n\ {\it intexpr} \gbar \s X_n\ \s S_m\ {\it intexpr}\gbar \s L_\infty\ {\it intexpr}\ \s R_0\ {\it opintexpr}\\
    \gbar \s L_\infty\ (\s S\s X_n)^*\ \s R\s X_0\ {\it opintexpr}
\end{multline*}
\[ {\it opintexpr} \ccoloneqq \epsilon \gbar \s S_n\ {\it intexpr} \]
where $m\leq n\in\bbN$.

Let $\beta_0$ be the normal derived $\beta$-reducer.
And let $\beta$ be the derived $\beta$-reducer without the rule that a single character's priority is changed to zero (instead it just keeps the character's priority instead).
We define $\beta^*(\xi)$ to be the value of the full reduction of $\xi$ under $\beta$ (i.e. the string obtained by repeated applications of $\beta$ until convergence).
And $\beta^*_0(\xi)$ is defined similarly.

\begin{definition}

    A set of strings $S$ is \emph{closed under $\beta$-reductions} if when $\xi\in S$, then $\beta(\xi)\in S$
    And $S$ is \emph{closed under $\beta_0$-reductions} if when $\xi\in S$, then $\beta_0(\xi)\in S$.

\end{definition}

Obviously if $S$ is closed under $\beta_0$-reductions, it is closed under $\beta$-reductions.

\begin{lemma}

    ${\it intexpr}$ is closed under $\beta_0$-reductions, and for $\xi\in{\it intexpr}$, $\beta^*(\xi)=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$ where $k\geq0$ and $n_1<\cdots<n_{k+1}$.

\end{lemma}

\begin{proof}

    We prove this by induction on the length and structure of $\xi$.
    \begin{enumerate}
        \item For $\xi=\s X_n$, this is trivial as $\xi$ is a fixed-point of $\beta$.
        \item For $\xi=\s S\s X_n\xi'$, if $\s S\s X_n$ can be reduced with the first token in $\xi'$ then we have the following cases:
        \begin{enumerate}
            \item $\xi'=\s X_m$, then $\beta(\xi)=\s X_m$ which is an intermediate expression and of the desired form.
            \item $\xi'=\s S\s X_m\xi''$, then $\beta(\xi)=\s S\s X_m\xi''$ which is an intermediate expression, and since the length of this string is less than $\xi$, we can prove the second part of
            the lemma by induction.
            \item $\xi'=\s X_m\ \s S_k\xi''$ with $\xi''\in{\it intexpr}$, in which case $\beta(\xi)=\s X_m\s S_k\xi''$, which is in ${\it intexpr}$ (since $m\geq k$).
            And since the length of this string is one less than $\xi$, we prove the second part of the lemma by induction.
        \end{enumerate}
        Otherwise $\beta(\xi)=\s S\s X_n\beta(\xi')$, and since by induction $\beta(\xi')\in{\it intexpr}$ we have the first part of the lemma.
        For the second part, if $\xi'$ is not a fixed point of $\beta$, $\beta(\xi')$'s length is less than $\xi'$'s, and thus since $\beta^*(\xi)=\beta^*(\s S\s X_n\beta(\xi'))$ we can proceed by
        induction.
        If $\xi'$ is a fixed point of $\beta$, by our inductive assumption (the second part of the lemma), $\xi'=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$.
        And so $\xi$ is also a fixed point of $\beta$ and has the desired form (since $n<n_1$ as otherwise we could reduce $\xi$).
        \item For $\xi=\s X_n\s S_m\xi'$, this can be reduced to $\beta(\xi)=\s S\s X_m\xi'$ and we simply induct.
        \item For $\xi=\s L_\infty\xi'\s R_0\zeta$, note that since $\xi'$ doesn't start with $\s R\s X$, $\s L$ cannot be reduced with $\xi'$.
        Thus $\beta(\xi)=\s L_\infty\beta(\xi'\s R_0\zeta)$.
        If $\xi'$ can be reduced, this is just $\s L_\infty\beta(\xi')\s R_0\zeta$ and so we simply induct.

        Otherwise it is a fixed point and so $\xi'=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$ with $n_1<\cdots<n_{k+1}$.
        Then $\beta(\xi)=\s L_\infty\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s R\s X_0\zeta$ which is an intermediate expression whose length is shorter than $\xi$ so we induct.
        \item For $\xi=\s L_\infty\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s R\s X_0\zeta=\s L_\infty\xi'\s R\s X_0\zeta$, if $\xi'$ can be reduced or is empty
        (because $\s L_\infty\s R\s X_0\zeta\to\s X_\infty\zeta$), we simply induct.
        Otherwise $n_1<\cdots<n_k$ and so $\beta(\xi)=\s L_\infty\s S\s X_{n_1}\cdots\s S\s X_{n_{k-1}}\s R\s X_0\zeta$, which is an intermediate expression whose length is shorter.
        \qed
    \end{enumerate}

\end{proof}

Notice that $\beta^*_0(\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}})\in\s X$ (this can be proved by simple induction), and so by the above lemma we get that for any intermediate expression $\xi$,
$\beta^*_0(\xi)\in\s X$ (since $\beta^*_0(\xi)=\beta^*_0(\beta^*(\xi))$).

Furthermore, from $({\bf4})$ we see that $\beta^*(\s L_\infty\xi\s R_0)=\s L_\infty\beta^*(\xi)\s R_0$.
Since $\beta^*(\xi)=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$, it is trivial to then see that $\beta^*(\s L_\infty\xi\s R_0)=\beta^*_0(\xi)_\infty$ (we can assign $\beta^*_0(\xi)$ a priority
since as we said, it is in $\s X$).
So we have proven:

\begin{lemma}

    Let $\xi\in{\it intexpr}$, then $\beta^*(\s L_\infty\xi\s R_0)=\beta^*_0(\xi)_\infty$.

\end{lemma}

We introduce some more notation.
Let us write $\xi\to^*\xi'$ to mean that $\beta^n(\xi)=\xi'$ for some $n$ and $\xi\to^*_0\xi'$ to mean that $\beta^n_0(\xi)=\xi'$ for some $n$.
We write $\s s_n$ for $\s S_n(f_{\s s})$, $\s s_n[\s x]$ for $\s S\s X_n(\s x,f_{\s s})$, and $\s x_n$ for $\s X_n(\s x)$.
And if $\xi\in{\it intexpr}$ we write $\pi(\xi)$ for the set of priorities in $\xi$, and $\pi_u(\xi)$ for the set of priorities in $\xi$ of operators which are unparenthesized.

\begin{lemma}

    If $N\leq\pi_u(\xi)$ then $N\leq\pi_u(\beta(\xi))$ for $\xi\in{\it intexpr}$.
    If the first inequality is strict, so is the second.

\end{lemma}

\begin{proof}

    We prove this by induction on the length of $\xi$.
    \begin{enumerate}
        \item For $\xi=\s X_n$ this is trivial.
        \item For $\xi=\s S\s X_n\xi'$, if $\s S\s X_n$ can be reduced with the first token of $\xi'$, we simply note that it takes the second token's priority, which $N$ is already less than.
        Otherwise $\beta(\xi)=\s S\s X_n\beta(\xi')$ and so $\pi_u(\beta(\xi))=\set n\cup\pi_u(\beta(\xi'))\geq N$ as required.
        \item For $\xi=\s X_n\s S_m\xi'$, $\beta(\xi)=\s S\s X_m\xi'$ and this is trivial.
        \item For $\xi=\s L_\infty\xi'\s R_0\zeta$, a $\beta$-reduction gives a string of the form $\s L_\infty\xi''\s R\s X_0\zeta$ and so $\pi_u(\beta(\xi))=\pi_u(\zeta)\geq N$.
        \item For $\xi=\s L_\infty\xi'\s R\s X_0\zeta$, a $\beta$-reduction either gives a string of the form $\s L_\infty\xi''\s R\s X_0\zeta$ as before, or $\s X_\infty\zeta$.
        The first case was covered in the last point, the second case is simple: $N\leq\pi_u(\zeta)$ and certainly $N\leq\infty$.
    \end{enumerate}
    Note that if we have a strict inequality, all of our arguments still hold.
    \qed

\end{proof}

From this and the previous lemma we can infer that if $n\leq\pi_u(\xi)$ then $n\leq\pi_u(\beta^*(\xi))=\pi(\beta^*(\xi))$ (since $\beta^*(\xi)$ cannot have parentheses).
So $\beta^*(\xi)$ is of the form
\[ \s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}},\qquad n\leq n_1<\cdots<n_{k+1} \]

\begin{lemma}

    Let $\xi$ be an intermediate expression, and $n\leq\pi_u(\xi)$.
    Then $\beta^*(\xi\s s_n)=\s s_n[\beta^*_0(\xi)]$.

\end{lemma}

\begin{proof}

    As explained above,
    \[ \xi\s s_n \to^* \beta^*(\xi)\s s_n = \s s^1_{n_1}[\s x^1]\cdots\s s^k_{n_k}[\s x^k]\s x^{k+1}_{n_{k+1}}\s s_n \]
    for $n\leq n_1<\cdots<n_k<n_{k+1}$.
    A $\beta$-reduction gives
    \[ \s s^1_{n_1}[\s x^1]\cdots\s s^k_{n_k}[\s x^k]\s s_n[\s x^{k+1}] \]
    further $\beta$-reductions give (instead of $f_{\s s_i}$ we write $f_i$)
    \[ \s s^1_{n_1}[\s x^1]\cdots\s s^{k-1}_{n_{k-1}}[\s x^{k-1}]\s s_n[f_k(\s x^k,\s x^{k+1})] \to \cdots \to \s s_n\bigl[f_1(\s x^1,f_2(\s x^2,\dots))\bigr] \]
    And it isn't hard to show that the $\beta^*_0$-reduction of $\s s^1_{n_1}[\s x^1]\cdots\s s^k_{n_k}[\s x^k]\s x^{k+1}_{n_{k+1}}$ is precisely $f_1(\s x^1,f_2(\s x^2,\dots))$ as required.
    \qed

\end{proof}

\begin{lemma}

    If $\xi\in{\it intexpr}$, $n<\pi_u(\xi)$.
    Then $\beta^*_0(\s s_n[\s x]\xi)=f_{\s s}(\s x,\beta^*_0(\xi))$.

\end{lemma}

\begin{proof}

    Since $n<\pi(\xi)$, $\s s_n[\s x]$ cannot be reduced with $\xi$, and since $n<\pi_u(\beta^k(\xi))$ for all $k$.
    Since $\s s_n[\s x]$ can only reduce with unparenthesized expressions (that is, there is no rule reducing $\s S\s X$ and $\s L$), this means that $\s s_n[\s x]$ cannot be reduced until $\xi$ is
    reduced fully.
    So
    \[ \s s_n[\s x]\xi \to^* \s s_n[\s x]\beta^*(\xi) = \s s_n[\s x]\s s_{n_1}[\s x^1]\cdots\s s_{n_k}[\s x^k]\s x^{k+1}_{n_{k+1}} \]
    where $n<n_1<\cdots<n_{k+1}$.
    A $\beta_0$-reduction just changes the priority of $\s x^{k+1}$ to $0$.
    Then $\beta^*_0$-reducing gives as before
    \[ f_{\s s}(\s x,f_{\s s_1}(\s x^1,f_{\s s_2}(\s x^2,\dots))) = f_{\s s}(\s x,\beta^*_0(\xi)) \qed \]

\end{proof}

\begin{theorem}

    For $\xi\in{\it expr}$, $\eval(\xi)=\beta^*_0(\xi)$.

\end{theorem}

\begin{proof}

    We induct on the length of $\xi$.
    \begin{enumerate}
        \item If $\xi=\s X_m$ this is trivial.
        \item If $\xi=\s L_\infty\xi'\s R_0\zeta$ then $\xi\to^*=\beta^*_0(\xi')\zeta$ which inductively is equal to $\eval(\xi')\zeta$.
        And inductively $\beta^*_0(\eval(\xi')\zeta)=\eval(\eval(\xi')\zeta)$ which is equal to $\eval(\xi)$ by definition.
        \item Otherwise let $n$ be the lowest priority of operators in $\xi$ which are unparenthesized, i.e. $n=\min\pi_u(\xi)$.
        Then $\xi=\xi'\s s_n\xi''$ where $\s s$ is the last unparenthesized operator with a priority of $n$, so $\xi',\xi''\in{\it intexpr}$ and $n\leq\pi_u(\xi')$ and $n<\pi_u(\xi'')$.
        This means that
        \[ \xi \to^* \s s_n[\beta^*_0(\xi')]\xi'' \to^*_0 f_{\s s}(\beta^*_0(\xi'),\beta^*_0(\xi'')) \]
        Inductively $\beta^*_0(\xi')=\eval(\xi')$ and $\beta^*_0(\xi'')=\eval(\xi'')$, meaning this is equal to $f_{\s s}(\eval(\xi'),\eval(\xi''))$.
        By definition this is $\eval(\xi)$ as required.
        \qed
    \end{enumerate}

\end{proof}

\end{document}

### 2024-11-05 Paper/paper.tex

@@
 \usepackage{enumitem}
+\usepackage{float}
 \usepackage{hyperref}
-\usepackage{float}
+\usepackage{listings}
+\usepackage{graphicx}
 
@@
 
-\begin{algo}\label{alg:eval}
-\caption{\eval uating arithmetical expressions}
+\begin{algo}
 \algorithm
@@
     \ElseIf{$\xi[0]=\s L$}
-        \State $\xi$ is of the form $\s L\rho\s R\xi'$
-        \State\Return $\eval(\eval(\rho)\xi')$
+        \State $\xi$ is of the form $\s L\xi'\s R\zeta$
+        \State\Return $\eval(\eval(\xi')\zeta)$
     \Else
@@
 \ealgorithm
+\caption{\eval uating arithmetical expressions}
+\label{alg:eval}
 \end{algo}
@@
 
+\section{Comparing Implementations}
+
+In this section we will comapre the runtimes of both implementations (one using linear reduction and one using menhir\NEEDCITE{}).
+A few notes beforehand:
+\begin{itemize}
+    \item The linear reduction implementation has more features than the menhir implementation, this should not severely impact the comparisons and we will of course not be comparing the implementations
+            on code which the menhir implementation does not support.
+    \item LLang supports left composition.
+            Meaning code of the form \ttt{f x y} is read as \ttt{(f x) y}.
+            We found no way of supporting this with menhir.
+            This is because our grammar for the menhir implementation contains the following fragment:
+            \[ {\it value} \ccoloneqq {\it const}\gbar{\it ident}\gbar{\it funcall},\qquad {\it funcall}\ccoloneqq{\it value}\ {\it value} \]
+            So \ttt{f x y} can be parsed as \ttt{f (x y)} instead:
+
+            \centerline{\begin{tikzpicture}[
+                    level 1/.style = {sibling distance=10em},
+                    level 2/.style = {sibling distance=10em},
+                    level 3/.style = {sibling distance=10em},
+                    level 4/.style = {sibling distance=10em},
+                    level 5/.style = {sibling distance=10em},
+                    level 6/.style = {sibling distance=10em},
+                    ]
+                \node {\it value}
+                    child {node {\it funcall}
+                        child {node {\it value}
+                            child {node {\it ident}
+                                child {node {\ttt f}}
+                            }
+                        }
+                        child {node {\it value}
+                            child {node {\it funcall}
+                                child {node {\it value}
+                                    child {node {\it ident}
+                                        child {node {\ttt x}}
+                                    }
+                                }
+                            }
+                            child {node {\it funcall}
+                                child {node {\it value}
+                                    child {node {\it ident}
+                                        child {node {\ttt y}}
+                                    }
+                                }
+                            }
+                        }
+                    };
+            \end{tikzpicture}
+            }
+\end{itemize}
+
+\subsection{Currying}
+
+We begin with a simple program to demonstrate currying.
+Currying is the action of taking a function $f\colon X\times Y\longto Z$ and converting it to a function $c(f)\colon X\longto(Y\longto Z)$ where $c(f)(x)\colon y\mapsto f(x,y)$.
+Our LLang code is as follows:
+
+\lstinputlisting[language=Caml, frame=single]{../Demos/currying.ml}
+
+Comparing the time it takes both implementations to run this 300 times, we see that it takes linear reduction 0.0040886 seconds, opposed to the classical implementation which takes 0.0036926 seconds.
+
+\subsection{Arithmetical Expressions}
+
+We can generate random arithmetical expressions using the following algorithm:
+
+\medskip
+\algorithm
+\Comment Create an expression with $\ell$ subterms, with a parentheses depth of $d$,\cr where numbers are chosen from $[0,n)$.\EndComment
+\Function{Generate-Expression}{$\ell,d,n$}
+    \lIf{$d=0$} \vtop{\hsize=.8\hsize
+        \noindent \Return an expression with $\ell$ randomly chosen numbers in\hfil\break $[0,n)$ and $\ell-1$ random operators in $\set{+,-,\times,\div}$.}
+    \State \Return $\bigl(\hbox{\tencsc Generate-Expression}(\ell,d-1,n)\bigr)\circ_1\cdots\circ_{\ell-1}$
+    \nonum\State \hfill$\bigl(\hbox{\tencsc Generate-Expression}(\ell,d-1,n)\bigr)$
+    \nonum\State where $\circ_i\in\set{+,-,\times,\div}$ are chosen randomly
+\EndFunc
+\ealgorithm
+\medskip
+
+\noindent We then run LLang code of the form \ttt{let x = <expr>;} where \ttt{<expr>} is generated by the above algorithm.
+We measure how long it takes to run this code five times with expressions generated by $\hbox{\tencsc Generate-Expression}(n,n,100)$ as $n$ varies.
+Doing so produces the following graph in figure \ref{fig:expr}.
+
+\begin{figure}
+
+    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr.png}}
+    \caption{Time to compute $\hbox{\tencsc Generate-Expression}(n,n)$}
+    \label{fig:expr}
+
+\end{figure}
+
+\subsection{Fibonacci}
+
+The fibonacci sequence is a linear recurrence defined by $F_n=F_{n-1}+F_{n-2}$ for $n>1$ and $F_1=F_0=1$.
+We wrote a script to compute the $n$th fibonacci number recursively in LLang:
+
+\begin{lstlisting}[language=Caml, frame=single]
+fun fib (x) {
+    if (x < 2) {
+        1
+    }{
+        (fib (x-1)) + (fib (x-2))
+    }
+}
+let x = (fib n);
+_prim_print x
+\end{lstlisting}
+
 \end{document}

### 2024-11-06 Paper/paper.tex
Updated paper

@@
 \noindent We then run LLang code of the form \ttt{let x = <expr>;} where \ttt{<expr>} is generated by the above algorithm.
-We measure how long it takes to run this code five times with expressions generated by $\hbox{\tencsc Generate-Expression}(n,n,100)$ as $n$ varies.
-Doing so produces the following graph in figure \ref{fig:expr}.
+We measure how long it takes to run this code five times with expressions generated by $\hbox{\tencsc Generate-Expression}(n,n,100)$ as $n$ varies (figure \ref{fig:expr1}) and
+expressions generated by $\hbox{\tencsc Generate-Expression}(n,1,100)$ as $n$ varies (figure \ref{fig:expr2}).
 
@@
     \caption{Time to compute $\hbox{\tencsc Generate-Expression}(n,n)$}
-    \label{fig:expr}
+    \label{fig:expr1}
+    \medskip
+    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr1.png}}
+    \caption{Time to compute $\hbox{\tencsc Generate-Expression}(n,1)$}
+    \label{fig:expr2}
 
@@
 
+\noindent Running each implementation 5 times for each $x$, we get the following graph in figure \ref{fig:fib} (as $x$ varies):
+
+\begin{figure}
+
+    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr.png}}
+    \caption{Time to compute $\hbox{\tt fib}(x)$}
+    \label{fig:fib}
+
+\end{figure}
+
+Our implementation takes approximately $\sim ab^x$ seconds to run where $a=0.000138585, b=1.64685$, and the menhir-based interpreter takes $\sim ab^x$ seconds where $a=0.00000362553, b=1.5809$.
+
+\subsection{Conclusion}
+
+We can see that linear reduction is significantly slower than the menhir-based interpreter.
+This can be for various reasons:
+\begin{enumerate}
+    \item Linear reduction is an $O(n^2)$-time algorithm (when no reduction produces a string of printable tokens, and every reduction reduces two tokens to one).
+        This is because a string of the form $\s s^1_1\s s^2_2\cdots\s s^n_n$ will take $n$ applications of the $\beta$-reducer, each application taking $n-k$ iterations (since $\beta$ is recursive).
+        This can be made more efficient: instead of starting the next $\beta$-reduction at the beginning of a string, it can be started at the token before the one just reduced.
+        Meaning once $\s s^{n-1}$ and $\s s^n$ are reduced to give $\s x^{n-1}$, we can begin the next iteration of the $\beta$-reducer at $\s s^{n-2}\s x^{n-1}$.
+        Unfortunately, implementing this did not lead to substantial increases in efficiency.
+
+    \item When handling function calls, linear reduction simply re-reduces the string in the definition of the function.
+        This requires it to parse and interpret the function again.
+        But classically, since the function definition has already been parsed and a parse tree created, further parsing of the function is not created.
+        This means that classical methods handle function calls better than linear reduction, but on the other hand linear reduction can handle macro-like programming better.
+        For example, if one were to change the state to map \ttt( to $\Lbrace$, the menhir-based interpreter would not be able to reparse the function to reinterpret \ttt(.
+        But this naturally happens with linear reduction, for better or worse.
+        \gobble)\gobble)
+
+    \item The standard methods of parsing code are well-studied and optimized.
+        Menhir does not naively parse code, it is an optimized tool.
+        Thus comparing a naive implementation of linear reduction to menhir is not a fair comparison.
+\end{enumerate}
+
+\section{Producing Parse Trees}
+
+Linear reduction can be used to just parse, instead of parsing and interpreting simultaneously.
+This has its benefits, as separating parsing and interpretation allows for the inspection of code before running it, which can help alert the programmer to syntax errors.
+Furthermore, this means linear reduction can be used in the classical interpreter and compiler pipeline, as it can be used as a parser.
+
+To do so, we define $\tree$, the set of trees, as follows:
+\[ \tree \ccoloneqq \Sigma \gbar \Sigma(\tree) \gbar \Sigma(\tree)(\tree) \gbar \Sigma(\tree)(\tree)(\tree) \gbar \cdots \]
+So a tree is simply a multi-branched tree whose nodes are elements of our alphabet $\Sigma$.
+Then we define an \textit{initial $\beta$-reducer} as
+\[ \hb\colon \Sigma\times\Sigma \partial \Sigma \times (\bbZ\times\bbZ\to\bbZ) \times (\tree\times\tree\to\tree) \]
+This extends to a \textit{derived $\beta$-reducer}
+\[ \beta\colon (\oS\times\tree)^* \longto (\oS\times\tree)^* \]
+where $\beta(\xi)$ is defined as
+\begin{enumerate}
+    \item If $\xi=\sigma^1_n(t)\sigma^2_m(s)\xi'$ where $n\geq m$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho,f)$ is defined then $\beta(\xi)=\tau_{\rho(n,m)}(f(t,s))\xi'$.
+    \item Otherwise $\beta(\xi)=\sigma^1_n(t)\beta(\sigma^2_m(s)\xi')$.
+\end{enumerate}
+We can also add the case that if $\xi$ is a single token then $\beta(\xi)=\xi_0$.
+
+For example, for arithmetic we can define
+\blist
+    \item $\NUM\ \OP\to\OP\NUM,\snd,(t,f\mapsto f(t))$
+    \item $\OP\NUM\ \NUM\to \NUM,\snd,(f(t),s\mapsto f(t)(s))$
+    \item $\OP\NUM\ \OP\NUM\to \OP\NUM,\snd,\bigl(f(t),g(s)\mapsto g(f(t)(s))\bigr)$
+    \item $\NUM\ \RPAREN\to \RPAREN\NUM,\snd,(t,\_\mapsto t)$
+    \item $\OP\NUM\ \RPAREN\NUM\to \RPAREN\NUM,\snd,(f(t),s\mapsto f(t)(s))$
+    \item $\LPAREN\ \RPAREN\NUM\to \NUM,\fst,(\_,t\mapsto t)$
+\elist
+
+So for example, \ttt{1 + 2 * (3 + 4)} is then parsed as
+\begin{align*}
+    \mb1_\infty \mb+_1 \mb2_\infty \mb*_2 \mb(_\infty \mb3_\infty \mb+_1 \mb4_\infty \mb)_0 &\to^* \mb{+(1)}_1 \mb{*(2)}_2 \mb(_\infty \mb{+(3)}_1 \mb{4}_0\\
+    &\to \mb{+(1)}_1 \mb{*(2)}_2 \mb(_\infty \mb{+(3)(4)}_0\\
+    &\to \mb{+(1)}_1 \mb{*(2)}_2 \mb{+(3)(4)}_\infty\\
+    &\to^* \mb{+(1)}_1 \mb{*(2)(+(3)(4))}_0\\
+    &\to \mb{+(1)(*(2)(+(3)(4)))}_0
+\end{align*}
+\gobble)\gobble)
+
 \end{document}

### 2024-11-12 Comparisons/comparer.py
Fixed dockerfile

@@ -24,4 +24,6 @@ fig = plt.figure(figsize = (10, 5))
 plt.bar(exe_names, times, color="red")
 for i in range(len(exes)):
     print(f"{exe_names[i]}: {times[i] / num_exes}")
-plt.show()
+
+print("Saving image to comparer.png")
+plt.savefig("comparer.png")

### 2024-11-12 Comparisons/expr-compare.py

@@ -4,7 +4,7 @@ import sys
 import time
 
 num_exes = 5
-exe_nums = [i for i in range(2,6)]
+exe_nums = [i for i in range(2,60)]
 
 exes = ["../llang", "../llang-classical"]
 names = ["Our Algorithm", "Classical Algorithm"]
@@ -15,7 +15,7 @@ for e in range(len(exes)):
     times = []
     for n in exe_nums:
         print(f"Run {n}")
-        os.system(f"python3 ../Demos/create-expression.py {n} {n}")
+        os.system(f"python3 ../Demos/create-expression.py {n} {1}")
         start = time.time()
         for i in range(num_exes):
             os.system(f"{exes[e]} expression.txt y")
@@ -26,6 +26,6 @@ for e in range(len(exes)):
     plt.plot(exe_nums, times, label=names[e])
 
 plt.legend()
-plt.show()
-
+print("Saving image to expr.png")
+plt.savefig("expr.png")
 

### 2024-11-12 Comparisons/fib-compare.py

@@ -20,7 +20,8 @@ for e in range(len(exes)):
             if (x < 2) {{
                 1
             }}{{
-                (fib (x-1)) + (fib (x-2))
+                let y = (fib (x-1)) + (fib (x-2));
+                y
             }}
         }}
         let x = (fib {n});
@@ -38,5 +39,6 @@ for e in range(len(exes)):
     plt.plot(exe_nums, times, label=names[e])
 
 plt.legend()
-plt.show()
+print("Saving image to fib.png")
+plt.savefig("fib.png")
 

### 2024-11-21 Paper/paper.tex
Fixed dockerfile
attempt to push

@@
 \title{Linear Reduction: Parsing and Interpreting Code Without Trees}
-\author{Ari Feiglin\inst{1}\and Dr. Yoni Zohar\inst{1}}
-\institute{University of Bar Ilan, Ramat Gan, Israel}
+\author{Ari Feiglin\inst{1}\and Yoni Zohar\inst{1}}
+\institute{Bar-Ilan University, Ramat Gan, Israel}
 
@@
 \begin{abstract}
-
+test
     In this paper we discuss an alternative method to parsing and interpreting human-readable code.
     This method utilizes an iterative approach, as opposed to a recursive tree-oriented approach found in most modern compilers and interpreters.
-    We will implement our algorithm in an interpreter and compare it with an interpreter created using more classical means.
+    We implement our algorithm in an interpreter and compare it with an interpreter created using more classical means.
 
@@
 
-We will propose a novel method for the process of interpretation, which doesn't use parse trees.
+We propose a novel method for the process of interpretation, which doesn't use parse trees.
 Parsing code generated by context-free grammars (CFGs) is inherently a recursive process, and thus it is not surprising that our algorithm can be altered to instead produce parse trees.
 
+The idea of the algorithm is as follows: the input is a stream of tokens, which are indivisible characters in any language.
+Given a string of tokens $\mb{\sigma^1}\cdots\mb{\sigma^n}$ (for readability, tokens are written in boxes), we form rules for combining two tokens.
+For example $\mb{\ttt1}$ and $\mb{\ttt+}$ can be combined together to give a token $\mb{\ttt{1,+}}$.
+Then if we follow a token of the form $\mb{\ttt{x,+}}$ with a token of the form $\mb{\ttt y}$, the tokens are combined to form a token $\mb{\ttt{x+y}}$.
+So for example:
+\[ \mbt{1} \mbt{+} \mbt{2} \mbt{+} \mbt{3} \to \mbt{1,+} \mbt2 \mbt+ \mbt3 \to \mbt{3} \mbt+ \mbt3 \to \mbt{3,+} \mbt3 \to \mbt6 \]
+But what about \ttt{1+2*3}?
+We'd get
+\[ \mbt1 \mbt+ \mbt2 \mbt* \mbt3 \to \mbt{1,+} \mbt2 \mbt* \mbt3 \to \mbt3 \mbt* \mbt3 \to \mbt{3,*} \mbt3 \to \mbt9 \]
+
+So to remedy this we add the concept of \textit{priorities}: for each token in our alphabet $\Sigma$, we assign a priority $p\in\bbN$.
+This can be thought of as a \textit{priority function} $\pi\colon\Sigma\longto\bbN$.
+So if the tokens $\mb{\sigma^1}\cdots\mb{\sigma^n}$ have respective priorities $p_1,\dots,p_n$ we can form the string $\mb{\sigma^1}_{p_1}\cdots\mb{\sigma^n}_{p_n}$ (where the subscript is part of the
+character).
+Now, we only combine two tokens if the token of the first is at least as great as the token of the second, and then the combined token gets the lower (the second) of the two priorities.
+So we can assign to tokens containing numbers a priority of $\infty$, the token \ttt+ gets a priority of $1$, and $\ttt*$ gets a priority of $2$.
+Then this becomes:
+\[ \mbt1_\infty \mbt+_1 \mbt2_\infty \mbt*_2 \mbt3_\infty \to \mbt{1,+}_1 \mbt2_\infty \mbt*_2 \mbt3_\infty \to \mbt{1,+}_1 \mbt{2,*}_2 \mbt3_\infty \]
+This cannot be combined any further, so we need another rule: we give the final token a priority of $0$, so this becomes:
+\[ \mbt{1,+}_1 \mbt{2,*}_2 \mbt3_0 \to \mbt{1,+}_1 \mbt{6}_0 \to \mbt{7}_0 \]
+which is the correct result.
+
 \section{Notation}
@@
     \item $\bN$ denotes the set of natural numbers, including $0$.
-    \item $\bbN$ is defined to be $\bN\cup\set\infty$.
+    \item $\bbN$ is defined to be $\bN\cup\set{+\infty}$.
     \item $\bZ$ denotes the set of integer numbers.
@@
 
-The idea of the algorithm is as follows: given a string of tokens $\mb{\sigma^1}\cdots\mb{\sigma^n}$ (for readability, tokens are written in boxes), we form rules for combining two tokens.
-For example $\mb{\ttt1}$ and $\mb{\ttt+}$ can be combined together to give a token $\mb{\ttt{1,+}}$.
-Then if we follow a token of the form $\mb{\ttt{x,+}}$ with a token of the form $\mb{\ttt y}$, the tokens are combined to form a token $\mb{\ttt{x+y}}$.
-So for example:
-\[ \mbt{1} \mbt{+} \mbt{2} \mbt{+} \mbt{3} \to \mbt{1,+} \mbt2 \mbt+ \mbt3 \to \mbt{3} \mbt+ \mbt3 \to \mbt{3,+} \mbt3 \to \mbt6 \]
-But what about \ttt{1+2*3}?
-We'd get
-\[ \mbt1 \mbt+ \mbt2 \mbt* \mbt3 \to \mbt{1,+} \mbt2 \mbt* \mbt3 \to \mbt3 \mbt* \mbt3 \to \mbt{3,*} \mbt3 \to \mbt9 \]
-
-So to remedy this we add the concept of \textit{priorities}: for each symbol in our alphabet $\Sigma$, we assign a priority $p\in\bbN$.
-This can be thought of as a \textit{priority function} $\pi\colon\Sigma\longto\bbN$.
-So if the tokens $\mb{\sigma^1}\cdots\mb{\sigma^n}$ have respective priorities $p_1,\dots,p_n$ we can form the string $\mb{\sigma^1}_{p_1}\cdots\mb{\sigma^n}_{p_n}$ (where the subscript is part of the
-character).
-Now, we only combine two tokens if the token of the first is at least as great as the token of the second, and then the combined token gets the lower (the second) of the two priorities.
-So we can assign to tokens containing numbers a priority of $\infty$, the token \ttt+ gets a priority of $1$, and $\ttt*$ gets a priority of $2$.
-Then this becomes:
-\[ \mbt1_\infty \mbt+_1 \mbt2_\infty \mbt*_2 \mbt3_\infty \to \mbt{1,+}_1 \mbt2_\infty \mbt*_2 \mbt3_\infty \to \mbt{1,+}_1 \mbt{2,*}_2 \mbt3_\infty \]
-This cannot be combined any further, so we need another rule: we will give the final token a priority of $0$, so this becomes:
-\[ \mbt{1,+}_1 \mbt{2,*}_2 \mbt3_0 \to \mbt{1,+}_1 \mbt{6}_0 \to \mbt{7}_0 \]
-which is the correct result.
-
-We can formalize this as follows: we start with an alphabet $\Sigma$.
+We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
 Define $\oS=\Sigma\times\bbN$, and we define a so-called \textit{initial $\beta$-reducer}: $\hb\colon \Sigma\times\Sigma \longto \Sigma$.
@@
 
-Notice that we need to add extra rules, like $\beta(\mbt{x,*},\mbt{y,+})=\mbt{x*y,+}$.
-This is so we can handle
+Notice that since in our reduction we get pairs of tokens in the form $\mbt{x,*},\mbt{y,+}$, we need to add extra rules to handle these cases.
+Specifically here we have $\hb(\mbt{x,*},\mbt{y,+})=\mbt{x*y,+}$.
+Namely the initial beta-reducer is the function:
+\begin{enumerate}
+    \item $\hb(\mbt x,\mb\circ)=\mbt{x,$\circ$}$ for an operator symbol $\circ$,
+    \item $\hb(\mbt{x,$\circ$},\mbt y)=\mbt{x$\circ$y}$,
+    \item $\hb(\mbt{x,$\circ$},\mbt{y,$\bullet$})=\mbt{x$\circ$y,$\bullet$}$,
+    \item and any other inputs can be given an arbitrary output.
+\end{enumerate}
+So for example we can now handle expressions of the form
 \begin{multline*}
@@
 But this has a major issue: it can't handle things like parentheses.
-So we need to come up with a second attempt.
+We could of course handle parentheses by changing how we assign priorities: every time we see an open parentheses we could increase
+the priority of each token and then beta-reduce.
+But this complicates how we allocate priorities, and does not generalize well.
+Furthermore the notion introduced in the next section, partial reduction, generalizes well and is used as a basepoint in the
+remainder of this paper.
 
@@
 
-Here we will abstract away the value of each term; the value of the new term given by the initial $\beta$-reducer will be determined by an output function.
+Here we abstract away the value of each term; the value of the new term given by the initial $\beta$-reducer will be determined by an output function.
 Our alphabet of types $\Sigma$, instead of having a token for every number, will just have tokens for types (number, operator, etc.).
-We will also have another set $\values$, the \textit{universe of values}.
+We also have another set $\values$, the \textit{universe of values}.
 
@@
 
-Finally, our \textit{initial $\beta$-reducer} is a function ($\state$ is a set of functions $\termp\partial\expi$ which we will define later)
+Finally, our \textit{initial $\beta$-reducer} is a function ($\state$ is a set of functions $\termp\partial\expi$ which we define later)
 \[ \hb\colon\termi\times\term^\epsilon \partial \termi^\epsilon\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\times\state\partial\values\times\termp^*\times\state) \]
@@
 
-We will implement our algorithm in an interpreter for a toy language we dubbed ``LLang''.
-We will also create an interpreter for LLang using menhir\NEEDCITE{}.
-In this section we will detail the implementation of our algorithm and compare and contrast the two implementations.
+We implement our algorithm in an interpreter for a toy language we dub ``LLang''.
+We also create an interpreter for LLang using menhir\NEEDCITE{}.
+In this section we detail the implementation of our algorithm and compare and contrast the two implementations.
 
@@
 
-In this section we will show that valued reduction (\S\ref{valued-reduction}) computes arithmetical expressions correctly.
-To do so, we will show that valued reduction computes expressions and gives a result equal to a na\"\i ve algorithm.
+In this section we show that valued reduction (\S\ref{valued-reduction}) computes arithmetical expressions correctly.
+To do so, we show that valued reduction computes expressions and gives a result equal to a na\"\i ve algorithm.
 
@@
 
-As opposed to pure valued (stateless) reduction, we will adopt some conventions from stateful reduction.
+As opposed to pure valued (stateless) reduction, we adopt some conventions from stateful reduction.
 Since our string is a string in $(\set{\s L,\s R}\cup\s X\cup\s S)^*$ and not $\set{\s L,\s R,\s X,\s S}^*$ (note the difference!), we must first convert the string into one admissable for
@@
 
-Secondly, we will adopt the convention of having type terms, abstract terms, etc.
+Secondly, we adopt the convention of having type terms, abstract terms, etc.
 Our base type will be $\Sigma=\set{\s X}$, and our atomic abstract terms $\mathcal A=\set{\s L_0,\s R_1,\s S_1}$ (printable terms are $\set{\s L,\s R}\cup\s X\cup\s S$, but as said before we assumed these
@@
 when considering an input for $\eval$, we will of course neglect the priorities in the string.
-And when considering an input for $\beta$-reduction, we will act as if each character is of the term indicated with the value in the string (i.e. $1_\infty+_12_\infty\in{\it expr}$ is
+And when considering an input for $\beta$-reduction, we act as if each character is of the term indicated with the value in the string (i.e. $1_\infty+_12_\infty\in{\it expr}$ is
 $\s X_\infty(1)\s S_1(+)\s X_\infty(2)$).
@@
 
+\section{Producing Parse Trees}
+
+Linear reduction can be used to just parse, instead of parsing and interpreting simultaneously.
+This has its benefits, as separating parsing and interpretation allows for the inspection of code before running it, which can help alert the programmer to syntax errors.
+Furthermore, this means linear reduction can be used in the classical interpreter and compiler pipeline, as it can be used as a parser.
+
+To do so, we define $\tree$, the set of trees, as follows:
+\[ \tree \ccoloneqq \Sigma \gbar \Sigma(\tree) \gbar \Sigma(\tree)(\tree) \gbar \Sigma(\tree)(\tree)(\tree) \gbar \cdots \]
+So a tree is simply a multi-branched tree whose nodes are elements of our alphabet $\Sigma$.
+Then we define an \textit{initial $\beta$-reducer} as
+\[ \hb\colon \Sigma\times\Sigma \partial \Sigma \times (\bbZ\times\bbZ\to\bbZ) \times (\tree\times\tree\to\tree) \]
+This extends to a \textit{derived $\beta$-reducer}
+\[ \beta\colon (\oS\times\tree)^* \longto (\oS\times\tree)^* \]
+where $\beta(\xi)$ is defined as
+\begin{enumerate}
+    \item If $\xi=\sigma^1_n(t)\sigma^2_m(s)\xi'$ where $n\geq m$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho,f)$ is defined then $\beta(\xi)=\tau_{\rho(n,m)}(f(t,s))\xi'$.
+    \item Otherwise $\beta(\xi)=\sigma^1_n(t)\beta(\sigma^2_m(s)\xi')$.
+\end{enumerate}
+We can also add the case that if $\xi$ is a single token then $\beta(\xi)=\xi_0$.
+
+For example, for arithmetic we can define
+\blist
+    \item $\NUM\ \OP\to\OP\NUM,\snd,(t,f\mapsto f(t))$
+    \item $\OP\NUM\ \NUM\to \NUM,\snd,(f(t),s\mapsto f(t)(s))$
+    \item $\OP\NUM\ \OP\NUM\to \OP\NUM,\snd,\bigl(f(t),g(s)\mapsto g(f(t)(s))\bigr)$
+    \item $\NUM\ \RPAREN\to \RPAREN\NUM,\snd,(t,\_\mapsto t)$
+    \item $\OP\NUM\ \RPAREN\NUM\to \RPAREN\NUM,\snd,(f(t),s\mapsto f(t)(s))$
+    \item $\LPAREN\ \RPAREN\NUM\to \NUM,\fst,(\_,t\mapsto t)$
+\elist
+
+So for example, \ttt{1 + 2 * (3 + 4)} is then parsed as
+\begin{align*}
+    \mb1_\infty \mb+_1 \mb2_\infty \mb*_2 \mb(_\infty \mb3_\infty \mb+_1 \mb4_\infty \mb)_0 &\to^* \mb{+(1)}_1 \mb{*(2)}_2 \mb(_\infty \mb{+(3)}_1 \mb{4}_0\\
+    &\to \mb{+(1)}_1 \mb{*(2)}_2 \mb(_\infty \mb{+(3)(4)}_0\\
+    &\to \mb{+(1)}_1 \mb{*(2)}_2 \mb{+(3)(4)}_\infty\\
+    &\to^* \mb{+(1)}_1 \mb{*(2)(+(3)(4))}_0\\
+    &\to \mb{+(1)(*(2)(+(3)(4)))}_0
+\end{align*}
+\gobble)\gobble)
+
 \section{Comparing Implementations}
 
-In this section we will comapre the runtimes of both implementations (one using linear reduction and one using menhir\NEEDCITE{}).
+In this section we compare the runtimes of both implementations (one using linear reduction and one using menhir\NEEDCITE{}).
 A few notes beforehand:
@@
 
-\subsection{Conclusion}
+\section{Conclusion}
 
@@
 
-\section{Producing Parse Trees}
-
-Linear reduction can be used to just parse, instead of parsing and interpreting simultaneously.
-This has its benefits, as separating parsing and interpretation allows for the inspection of code before running it, which can help alert the programmer to syntax errors.
-Furthermore, this means linear reduction can be used in the classical interpreter and compiler pipeline, as it can be used as a parser.
-
-To do so, we define $\tree$, the set of trees, as follows:
-\[ \tree \ccoloneqq \Sigma \gbar \Sigma(\tree) \gbar \Sigma(\tree)(\tree) \gbar \Sigma(\tree)(\tree)(\tree) \gbar \cdots \]
-So a tree is simply a multi-branched tree whose nodes are elements of our alphabet $\Sigma$.
-Then we define an \textit{initial $\beta$-reducer} as
-\[ \hb\colon \Sigma\times\Sigma \partial \Sigma \times (\bbZ\times\bbZ\to\bbZ) \times (\tree\times\tree\to\tree) \]
-This extends to a \textit{derived $\beta$-reducer}
-\[ \beta\colon (\oS\times\tree)^* \longto (\oS\times\tree)^* \]
-where $\beta(\xi)$ is defined as
-\begin{enumerate}
-    \item If $\xi=\sigma^1_n(t)\sigma^2_m(s)\xi'$ where $n\geq m$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho,f)$ is defined then $\beta(\xi)=\tau_{\rho(n,m)}(f(t,s))\xi'$.
-    \item Otherwise $\beta(\xi)=\sigma^1_n(t)\beta(\sigma^2_m(s)\xi')$.
-\end{enumerate}
-We can also add the case that if $\xi$ is a single token then $\beta(\xi)=\xi_0$.
-
-For example, for arithmetic we can define
-\blist
-    \item $\NUM\ \OP\to\OP\NUM,\snd,(t,f\mapsto f(t))$
-    \item $\OP\NUM\ \NUM\to \NUM,\snd,(f(t),s\mapsto f(t)(s))$
-    \item $\OP\NUM\ \OP\NUM\to \OP\NUM,\snd,\bigl(f(t),g(s)\mapsto g(f(t)(s))\bigr)$
-    \item $\NUM\ \RPAREN\to \RPAREN\NUM,\snd,(t,\_\mapsto t)$
-    \item $\OP\NUM\ \RPAREN\NUM\to \RPAREN\NUM,\snd,(f(t),s\mapsto f(t)(s))$
-    \item $\LPAREN\ \RPAREN\NUM\to \NUM,\fst,(\_,t\mapsto t)$
-\elist
-
-So for example, \ttt{1 + 2 * (3 + 4)} is then parsed as
-\begin{align*}
-    \mb1_\infty \mb+_1 \mb2_\infty \mb*_2 \mb(_\infty \mb3_\infty \mb+_1 \mb4_\infty \mb)_0 &\to^* \mb{+(1)}_1 \mb{*(2)}_2 \mb(_\infty \mb{+(3)}_1 \mb{4}_0\\
-    &\to \mb{+(1)}_1 \mb{*(2)}_2 \mb(_\infty \mb{+(3)(4)}_0\\
-    &\to \mb{+(1)}_1 \mb{*(2)}_2 \mb{+(3)(4)}_\infty\\
-    &\to^* \mb{+(1)}_1 \mb{*(2)(+(3)(4))}_0\\
-    &\to \mb{+(1)(*(2)(+(3)(4)))}_0
-\end{align*}
-\gobble)\gobble)
-
 \end{document}

### 2024-12-04 Paper/paper.tex
Updated paper
Added bibliography

@@
 \usepackage{graphicx}
+\usepackage[backend=biber]{biblatex}
+
+\addbibresource{paper-bib.bib}
 
@@
 \begin{abstract}
-test
     In this paper we discuss an alternative method to parsing and interpreting human-readable code.
@@
 
-Nowadays, most programming language interpreters follow the following pipeline:
+Nowadays, most programming language interpreters follow the following pipeline (see \cite{dragon})
 \begin{enumerate}
@@
 Unfortunately, constructing such a tree is not so easy.
-Multiple methods for doing so exist, as well as tools for automating this process.
+Multiple methods for doing so exist (some are listed in \cite{dragon}), as well as tools for automating this process.
 
 We propose a novel method for the process of interpretation, which doesn't use parse trees.
-Parsing code generated by context-free grammars (CFGs) is inherently a recursive process, and thus it is not surprising that our algorithm can be altered to instead produce parse trees.
+But our algorithm is not limited to this: parsing code generated by context-free grammars (CFGs) is inherently a recursive process, and thus it is not surprising that our algorithm can be altered to
+instead produce parse trees.
 
@@
 We implement our algorithm in an interpreter for a toy language we dub ``LLang''.
-We also create an interpreter for LLang using menhir\NEEDCITE{}.
+We also create an interpreter for LLang using menhir\cite{menhir}.
 In this section we detail the implementation of our algorithm and compare and contrast the two implementations.
@@
 \blist
-    \item $\iif\ \tsig \longto \ifbool\ \fst\ (\_,n,s\to n,\epsilon,s+[\{\mapsto\Lbrace\a,(\mapsto\lparen\a)]$
-    \item $\ifbool\ \code \longto \ifthen\ \fst\ (n,\xi,s\to(n,\xi),\epsilon,s)$
+    \item $\iif\ \tsig \longto \ifbooll\ \fst\ (\_,n,s\to n,\epsilon,s+[\{\mapsto\Lbrace\a,(\mapsto\lparen\a)]$
+    \item $\ifbooll\ \code \longto \ifthen\ \fst\ (n,\xi,s\to(n,\xi),\epsilon,s)$
     \item $\ifthen\ \code \longto \epsilon\ \_\ ((n,\xi_1),\xi_2,s\to\varnothing,(n=0?\ \xi_2:\xi_1),\pop s)$
@@
 
-In this section we compare the runtimes of both implementations (one using linear reduction and one using menhir\NEEDCITE{}).
+In this section we compare the runtimes of both implementations (one using linear reduction and one using menhir\cite{menhir}).
 A few notes beforehand:
@@
 
+\bigskip
+
+\printbibliography
+
 \end{document}

### 2024-12-05 Paper/paper.tex
Updated paper

@@
     \item Lexing: lexically analyze the input code, separating it into atomic \textit{tokens}.
-    \item Parsing: convert the parsed tokens into a \textit{parse tree}: a tree which represents the grammar of the code.
+    \item Parsing: convert the parsed tokens into a \textit{parse tree}: a tree which represents the code according to the language's grammar.
     \item Interpreting: recursively descend the tree and execute the code according to the instructions embedded in the tree.
@@
 We propose a novel method for the process of interpretation, which doesn't use parse trees.
-But our algorithm is not limited to this: parsing code generated by context-free grammars (CFGs) is inherently a recursive process, and thus it is not surprising that our algorithm can be altered to
-instead produce parse trees.
+But our algorithm is not limited to this, and can be altered to produce parse trees as well (see \S\ref{producing-parse-trees}).
+This means that our algorithm can be utilized as part of the pipeline for compilers and traditional interpreters.
 
@@
 Given a string of tokens $\mb{\sigma^1}\cdots\mb{\sigma^n}$ (for readability, tokens are written in boxes), we form rules for combining two tokens.
-For example $\mb{\ttt1}$ and $\mb{\ttt+}$ can be combined together to give a token $\mb{\ttt{1,+}}$.
+For example, $\mb{\ttt1}$ and $\mb{\ttt+}$ can be combined together to give a token $\mb{\ttt{1,+}}$.
 Then if we follow a token of the form $\mb{\ttt{x,+}}$ with a token of the form $\mb{\ttt y}$, the tokens are combined to form a token $\mb{\ttt{x+y}}$.
@@
 We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
-Define $\oS=\Sigma\times\bbN$, and we define a so-called \textit{initial $\beta$-reducer}: $\hb\colon \Sigma\times\Sigma \longto \Sigma$.
-This is a function which tells us how to combine tokens together.
-And we extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.
+Define $\oS=\Sigma\times\bbN$, and we assume the existence of a so-called \textit{initial $\beta$-reducer}:
+$\hb\colon \Sigma\times\Sigma \longto \Sigma$, which is defined by the implementer of the language being interpreted, and is the function which
+tells us how to combine tokens together.
+We extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.
 
@@
 Specifically here we have $\hb(\mbt{x,*},\mbt{y,+})=\mbt{x*y,+}$.
-Namely the initial beta-reducer is the function:
-\begin{enumerate}
-    \item $\hb(\mbt x,\mb\circ)=\mbt{x,$\circ$}$ for an operator symbol $\circ$,
+
+So let us define $\Sigma$ to consist of tokens of the form $\mbt{x}$ for numbers $\ttt x$, $\mb\circ$ for operator symbols $\circ$ (addition,
+multiplication, etc.), and $\mbt{x,$\circ$}$.
+Then we define the initial $\beta$-reducer by
+\begin{enumerate}
+    \item $\hb(\mbt x,\mb\circ)=\mbt{x,$\circ$}$ for operator symbols $\circ$,
     \item $\hb(\mbt{x,$\circ$},\mbt y)=\mbt{x$\circ$y}$,
@@
 \[ \hb\colon \Sigma\times\Sigma\partial\Sigma\times(\bbZ\times\bbZ\to\bbZ) \]
+We add a function to compute the new priority of the token to the output of the initial $\beta$-reducer instead of taking the priorities as inputs.
+This is because the new token computed should not be dependent on the priorities of the input tokens, as this just adds extra unnecessary
+complexity.
+
 Now we define the derived $\beta$-reducer as follows; it is a function $\beta\colon\oS^*\partial\oS^*$ where for $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$:
@@
 \end{enumerate}
+
+We define our alphabet $\Sigma$ to be the same as in the previous section.
 
@@
 \[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
-Our alphabet of types is $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$, define $\oP=\oS\times\values$ to be the set of all possible terms (which is a type, a priority, and a value).
+Our alphabet of types is $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$, define $\oP=\oS\times\values$ to be the set of all possible terms (a term consists of a type, a priority, and a value).
 Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the type; $p$ is the priority; and $v$ the value.
 
-Our rules for the derived $\beta$-reducer is similar to before: it is a function $\beta\colon\oP^* \partial \oP^*$ where for $\xi=\sigma^1_{p_1}(v_1)\cdots\sigma^n_{p_n}(v_n)$
+Our rules for the derived $\beta$-reducer are similar to before: it is a function $\beta\colon\oP^* \partial \oP^*$ where for $\xi=\sigma^1_{p_1}(v_1)\cdots\sigma^n_{p_n}(v_n)$
 \begin{enumerate}
@@
 \end{itemize}
+
+So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
+So it is converted to
+\[ \NUM_\infty(2)\ \OP_2(*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \]
+The derivation is then
+\begin{gather*}
+    \OP\NUM_2(2,*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \\
+    \OP\NUM_2(2,*)\ \LPAREN_\infty\ \OP\NUM_1(1,+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \\
+    \OP\NUM_2(2,*)\ \LPAREN_\infty\ \OP\NUM_1(1,+)\ \RPAREN\NUM_0(3)\ \OP_1(+)\ \NUM_\infty(4) \\
+    \OP\NUM_2(2,*)\ \LPAREN_\infty\ \RPAREN\NUM_0(4)\ \OP_1(+)\ \NUM_\infty(4) \\
+    \OP\NUM_2(2,*)\ \NUM_\infty(4)\ \OP_1(+)\ \NUM_\infty(4) \\
+    \OP\NUM_2(2,*)\ \OP\NUM_1(4,+)\ \NUM_\infty(4) \\
+    \OP\NUM_2(8,+)\ \NUM_\infty(4) \\
+    \OP\NUM_2(8,+)\ \NUM_0(4) \\
+    \NUM_0(12)
+\end{gather*}
 
@@
 
-\section{Producing Parse Trees}
+\section{Producing Parse Trees} \label{producing-parse-trees}
 

### 2024-12-14 Paper/paper.tex
Changed name to iterative reduction

@@
 
-\title{Linear Reduction: Parsing and Interpreting Code Without Trees}
+\title{Iterative Reduction: Parsing and Interpreting Code Without Trees}
 \author{Ari Feiglin\inst{1}\and Yoni Zohar\inst{1}}
@@
 
-We implement our algorithm in an interpreter for a toy language we dub ``LLang''.
-We also create an interpreter for LLang using menhir\cite{menhir}.
+We implement our algorithm in an interpreter for a toy language we dub ``ILang''.
+We also create an interpreter for ILang using menhir\cite{menhir}.
 In this section we detail the implementation of our algorithm and compare and contrast the two implementations.
@@
 
-Linear reduction can be used to just parse, instead of parsing and interpreting simultaneously.
+Iterative reduction can be used to just parse, instead of parsing and interpreting simultaneously.
 This has its benefits, as separating parsing and interpretation allows for the inspection of code before running it, which can help alert the programmer to syntax errors.
-Furthermore, this means linear reduction can be used in the classical interpreter and compiler pipeline, as it can be used as a parser.
+Furthermore, this means iterative reduction can be used in the classical interpreter and compiler pipeline, as it can be used as a parser.
 
@@
 
-In this section we compare the runtimes of both implementations (one using linear reduction and one using menhir\cite{menhir}).
+In this section we compare the runtimes of both implementations (one using iterative reduction and one using menhir\cite{menhir}).
 A few notes beforehand:
 \begin{itemize}
-    \item The linear reduction implementation has more features than the menhir implementation, this should not severely impact the comparisons and we will of course not be comparing the implementations
+    \item The iterative reduction implementation has more features than the menhir implementation, this should not severely impact the comparisons and we will of course not be comparing the implementations
             on code which the menhir implementation does not support.
-    \item LLang supports left composition.
+    \item ILang supports left composition.
             Meaning code of the form \ttt{f x y} is read as \ttt{(f x) y}.
@@
 Currying is the action of taking a function $f\colon X\times Y\longto Z$ and converting it to a function $c(f)\colon X\longto(Y\longto Z)$ where $c(f)(x)\colon y\mapsto f(x,y)$.
-Our LLang code is as follows:
+Our ILang code is as follows:
 
@@
 
-Comparing the time it takes both implementations to run this 300 times, we see that it takes linear reduction 0.0040886 seconds, opposed to the classical implementation which takes 0.0036926 seconds.
+Comparing the time it takes both implementations to run this 300 times, we see that it takes iterative reduction 0.0040886 seconds, opposed to the classical implementation which takes 0.0036926 seconds.
 
@@
 
-\noindent We then run LLang code of the form \ttt{let x = <expr>;} where \ttt{<expr>} is generated by the above algorithm.
+\noindent We then run ILang code of the form \ttt{let x = <expr>;} where \ttt{<expr>} is generated by the above algorithm.
 We measure how long it takes to run this code five times with expressions generated by $\hbox{\tencsc Generate-Expression}(n,n,100)$ as $n$ varies (figure \ref{fig:expr1}) and
@@
 The fibonacci sequence is a linear recurrence defined by $F_n=F_{n-1}+F_{n-2}$ for $n>1$ and $F_1=F_0=1$.
-We wrote a script to compute the $n$th fibonacci number recursively in LLang:
+We wrote a script to compute the $n$th fibonacci number recursively in ILang:
 
@@
 
-We can see that linear reduction is significantly slower than the menhir-based interpreter.
+We can see that iterative reduction is significantly slower than the menhir-based interpreter.
 This can be for various reasons:
 \begin{enumerate}
-    \item Linear reduction is an $O(n^2)$-time algorithm (when no reduction produces a string of printable tokens, and every reduction reduces two tokens to one).
+    \item Iterative reduction is an $O(n^2)$-time algorithm (when no reduction produces a string of printable tokens, and every reduction reduces two tokens to one).
         This is because a string of the form $\s s^1_1\s s^2_2\cdots\s s^n_n$ will take $n$ applications of the $\beta$-reducer, each application taking $n-k$ iterations (since $\beta$ is recursive).
@@
 
-    \item When handling function calls, linear reduction simply re-reduces the string in the definition of the function.
+    \item When handling function calls, iterative reduction simply re-reduces the string in the definition of the function.
         This requires it to parse and interpret the function again.
         But classically, since the function definition has already been parsed and a parse tree created, further parsing of the function is not created.
-        This means that classical methods handle function calls better than linear reduction, but on the other hand linear reduction can handle macro-like programming better.
+        This means that classical methods handle function calls better than iterative reduction, but on the other hand iterative reduction can handle macro-like programming better.
         For example, if one were to change the state to map \ttt( to $\Lbrace$, the menhir-based interpreter would not be able to reparse the function to reinterpret \ttt(.
-        But this naturally happens with linear reduction, for better or worse.
+        But this naturally happens with iterative reduction, for better or worse.
         \gobble)\gobble)
@@
         Menhir does not naively parse code, it is an optimized tool.
-        Thus comparing a naive implementation of linear reduction to menhir is not a fair comparison.
+        Thus comparing a naive implementation of iterative reduction to menhir is not a fair comparison.
 \end{enumerate}

### 2024-12-16 Paper/paper.tex
compiling like i like
adding comments infrastructure

@@
 \usepackage{graphicx}
-\usepackage[backend=biber]{biblatex}
-
-\addbibresource{paper-bib.bib}
+\usepackage{mdframed}
 
@@
 
+\newcommand{\yoni}[1]{{\begin{mdframed}[linecolor=red]{\color{red}#1}\end{mdframed}}}
+\newcommand{\ari}[1]{{\begin{mdframed}[linecolor=blue]{\color{blue}#1}\end{mdframed}}}
 \begin{document}
@@
 \section{Introduction}
-
+\yoni{comment}
 Nowadays, most programming language interpreters follow the following pipeline (see \cite{dragon})
@@
     \item Parsing: convert the parsed tokens into a \textit{parse tree}: a tree which represents the code according to the language's grammar.
+\yoni{comment}
     \item Interpreting: recursively descend the tree and execute the code according to the instructions embedded in the tree.
@@
 For example, $\mb{\ttt1}$ and $\mb{\ttt+}$ can be combined together to give a token $\mb{\ttt{1,+}}$.
+\yoni{comment}
 Then if we follow a token of the form $\mb{\ttt{x,+}}$ with a token of the form $\mb{\ttt y}$, the tokens are combined to form a token $\mb{\ttt{x+y}}$.
@@
 
-\bigskip
-
-\printbibliography
+\bibliographystyle{abbrv}
+\bibliography{paper-bib}
 

### 2024-12-16 Paper/paper.tex
an example for an example + minor stuff

@@
 \usepackage{mdframed}
+\usepackage{cleveref}
 
@@
 \subsection{A First Attempt -- Primitive Reduction}
-
+\label{sec:firstattempt}
 We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
@@
 
-Notice that since in our reduction we get pairs of tokens in the form $\mbt{x,*},\mbt{y,+}$, we need to add extra rules to handle these cases.
-Specifically here we have $\hb(\mbt{x,*},\mbt{y,+})=\mbt{x*y,+}$.
-
-So let us define $\Sigma$ to consist of tokens of the form $\mbt{x}$ for numbers $\ttt x$, $\mb\circ$ for operator symbols $\circ$ (addition,
-multiplication, etc.), and $\mbt{x,$\circ$}$.
-Then we define the initial $\beta$-reducer by
-\begin{enumerate}
-    \item $\hb(\mbt x,\mb\circ)=\mbt{x,$\circ$}$ for operator symbols $\circ$,
-    \item $\hb(\mbt{x,$\circ$},\mbt y)=\mbt{x$\circ$y}$,
-    \item $\hb(\mbt{x,$\circ$},\mbt{y,$\bullet$})=\mbt{x$\circ$y,$\bullet$}$,
-    \item and any other inputs can be given an arbitrary output.
-\end{enumerate}
-So for example we can now handle expressions of the form
+% Notice that since in our reduction we get pairs of tokens in the form $\mbt{x,*},\mbt{y,+}$, we need to add extra rules to handle these cases.
+% Specifically here we have $\hb(\mbt{x,*},\mbt{y,+})=\mbt{x*y,+}$.
+
+\begin{example}
+Define $\Sigma$ to consist of tokens of the form $\mbt{x}$ 
+for every $x\in\mathbb{N}$, 
+$\mb\circ$ for $\circ\in\{+,*\}$, and $\mbt{x,$\circ$}$
+for every $x\in\mathbb{N}$
+and $\circ\in\{+,*\}$.
+
+Then we define the initial $\beta$-reducer $\hb$ by
+\begin{enumerate}
+    \item $\hb(\mbt x,\mb\circ)=\mbt{x,$\circ$}$ for $\circ\in\{+,*\}$
+    \item $\hb(\mbt{x,$\circ$},\mbt y)=\mbt{x$\circ$y}$ for 
+    $x,y\in\mathbb{N}$ and $\circ\in\{+,*\}$
+    \item $\hb(\mbt{x,$\circ$},\mbt{y,$\bullet$})=\mbt{x$\circ$y,$\bullet$}$ 
+    for $x,y\in\mathbb{N}$ and $\circ,\bullet\in\{+,*\}$
+    \item $\hb$ is defined arbitrarily on any other input.
+\end{enumerate}
+
+Now, consider the expression $1*2+3$. 
+After assigning priorities, we obtain
+the following expression:
+\yoni{The last sentence is an example of how the assignment
+of priorities feels a bit out of context here. This could improve
+if they become a formal part of the framework, from the very beginning.}
+$\mbt1_\infty \mbt*_2 \mbt2_\infty \mbt+_1 \mbt3_\infty$.
+We now get:
 \begin{multline*}
@@
 \end{multline*}
+\end{example}
 
@@
 the priority of each token and then beta-reduce.
-But this complicates how we allocate priorities, and does not generalize well.
+But this complicates how we allocate priorities, 
+\yoni{would u say that currently we can set priorities statically, but with
+the idea of increasing priorities we would have to set them dynamically?
+If so, I would mention that.}
+and does not generalize well.
 Furthermore the notion introduced in the next section, partial reduction, generalizes well and is used as a basepoint in the
@@
 
-We define our alphabet $\Sigma$ to be the same as in the previous section.
-
+We define our alphabet $\Sigma$ to be the same as in~\Cref{sec:firstattempt}.
 Now we define our initial $\beta$-reducer which will support parentheses.

### 2024-12-16 Paper/paper.tex
comments to the second exmaple

@@
 
+\begin{example}
 We define our alphabet $\Sigma$ to be the same as in~\Cref{sec:firstattempt}.
@@
 And we define our priority function as follows:
+\yoni{feels out of context, the notion of a priority function
+was not defined. Also, notice that some tokens get priorities and some don't, which is a bit confusing.}
 \[ \pi(\mbt x) = \infty\hbox{ for numbers \ttt x},\quad \pi(\mbt+) = 1,\quad \pi(\mbt*) = 2,\quad \pi(\mbt() = \infty,\quad \pi(\mbt)) = 0 \]
@@
 \gobble)\gobble)
+
+\end{example}
 

### 2024-12-16 Paper/paper.tex
comments on the third example

@@
 
+\yoni{I would consider using the word "kinds" instead of "types". "kinds" is not great either, but types are stuff like int, real, int->int, which i am not sure is appropriate here.}
+
 The initial $\beta$-reducer is a function
 \[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
-Our alphabet of types is $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$, define $\oP=\oS\times\values$ to be the set of all possible terms (a term consists of a type, a priority, and a value).
+Define $\oP=\oS\times\values$ to be the set of all possible terms,
+that is, 
+a term consists of a type, a priority, and a value.
 Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the type; $p$ is the priority; and $v$ the value.
@@
 
+\begin{example}
+Define $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$.
 Our initial $\beta$-reducer will is defined as follows:
@@
 
+\yoni{I don't really understand it. For example, what is 
+$(v,f\mapsto(v,f))$? Perhaps introducing some more definitions
+and details can help clarify?
+}
+\yoni{What happened to the squares around the tokens in this example?}
+
 So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
@@
 \[ \NUM_\infty(2)\ \OP_2(*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \]
+
+\yoni{Here we are missing a formal assignment of priorities.}
+
 The derivation is then
@@
 
+\end{example}
 But this of course is not capable of handling the semantics necessary for programming languages.

### 2024-12-16 Paper/paper.tex
comments for states

@@
 The sets $\termp,\terms,\terma$ are disjoint, and we place no such restriction on $\values$.
+\yoni{The comment about no restrictions for $\values$ is unclear.}
 
@@
 \[ \terms \ccoloneqq \Sigma^0 \mid \Sigma^1\terms \mid \Sigma^2\terms\terms \mid \Sigma^3\terms\terms\terms \mid \cdots \mid \Sigma^\oo(\terms\cdots\terms) \]
+
+\yoni{I don't think this is very standard. Perhaps you can define
+ it inductively: Every element of $\Sigma^0$ is in $T_{\Sigma}$,
+  and for eac $n$, if $t_i\in T_{\Sigma}$ for each $1\leq i \leq n$
+  then $f t_1 \ldots t_n \in T_{\Sigma}$ for each $f\in\Sigma^{n}$,
+  and then for each $g\in\Sigma^{\infty}$ and any $n$, similarly...
+Do we actually need $\Sigma^{\infty}$ BTW? }
 
@@
 
-So if $\Sigma=\set{\NUM_0,\LIST_1,\PRODUCT_\infty}$ (the subscripts are the arities of each base type) and $\mathcal A=\set{\OP}$, then
+\begin{example}
+If $\Sigma^{0}=\{\NUM\}$,
+$\Sigma^{1}=\{\LIST\}$, and $\Sigma^{\infty}=\{\PRODUCT\}$
+and $\Sigma=\Sigma^0\cup\Sigma^1\cup\Sigma^\infty$,
+ and $\mathcal A=\set{\OP}$, 
+\yoni{plz specify the "arity" of $\OP$.}
+ then
 \[ \NUM,\LIST\NUM,\PRODUCT(\LIST\NUM\NUM)\in\terms,\qquad \OP\LIST\NUM\in\terma \]
-For readability we can put parentheses and commas where necessary, but importantly these sets have the \textit{unique reconstruction property}: it is possible to compute their construction from the
+\yoni{I think it is better to put a space between these, especially
+in $\LIST\NUM\NUM$.}
+\end{example}
+For readability we can put parentheses and commas where necessary, but importantly these sets have the \textit{unique reconstruction property}:
+\yoni{If I understand correctly you are assuming this property, right? If so, please say so.}
+it is possible to compute their construction from the
 string itself (i.e. a string cannot be constructed from two different patterns).
@@
 \begin{enumerate}
-    \item $\term=\termp\cup\terms\cup\terms$, the set of \textit{terms},
+    \item $\term=\termp\cup\terms\cup\terms$, 
+\yoni{Should the last $\terms$ be $\terma$?}
+    the set of \textit{terms},
     \item $\termi=\terms\cup\terma$, the set of \textit{internal terms},
@@
 Finally we define $\oe=\exp\times\bbZ$.
-Elements of $\oe$ will be written like $\sigma_n(v)$ as before; for printable terms we act as if the value is empty.
+Elements of $\oe$ will be written like $\sigma_n(v)$ as before; 
+\yoni{But elements of $\oe$ seems like pairs to me, not triplets.}
+for printable terms we act as if the value is empty.
+\yoni{Please clarify what does acting as if... means.}
 
@@
 This can be naturally extended to $\pi\colon\termp^*\longto\oe^*$.
+\yoni{please show how.}
 
 Finally, our \textit{initial $\beta$-reducer} is a function ($\state$ is a set of functions $\termp\partial\expi$ which we define later)
+\yoni{Using something and defining it later should be avoided...}
+
+\yoni{Generally, this section is hard to follow. I suggest adding plenty of examples, after every new definition / concept, please add an appropriate small example.}
+
 \[ \hb\colon\termi\times\term^\epsilon \partial \termi^\epsilon\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\times\state\partial\values\times\termp^*\times\state) \]

### 2024-12-17 Paper/paper.tex
Responded to comments, added example

@@
 \usepackage{mdframed}
-\usepackage{cleveref}
+%\usepackage{cleveref}
 
@@
 
-\subsection{A First Attempt -- Primitive Reduction}
-\label{sec:firstattempt}
+\subsection{A First Attempt -- Primitive Reduction} \label{sec:firstattempt}
+
 We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
@@
 
+To assign priorities to an input stream of tokens, we assume the existence of a \textit{priority function} $\pi\colon\Sigma\partial\bbN$.
+We can then extend this to a function $\pi\colon\Sigma^*\longto\oS^*$ by $\pi(\sigma^1\cdots\sigma^n)=\sigma^1_{\rho_1}\cdots\sigma^n_{\rho_n}$ where $\rho_i=\pi(\sigma^i)$.
+Notice that not every token is assigned an initial priority; such tokens thus cannot be in the initial input stream to the $\beta$-reduction.
+
 % Notice that since in our reduction we get pairs of tokens in the form $\mbt{x,*},\mbt{y,+}$, we need to add extra rules to handle these cases.
@@
 \begin{example}
-Define $\Sigma$ to consist of tokens of the form $\mbt{x}$ 
-for every $x\in\mathbb{N}$, 
-$\mb\circ$ for $\circ\in\{+,*\}$, and $\mbt{x,$\circ$}$
-for every $x\in\mathbb{N}$
+
+    Define $\Sigma$ to consist of tokens of the form $\mbt{x}$ for every $x\in\mathbb{N}$, $\mb\circ$ for $\circ\in\{+,*\}$, and $\mbt{x,$\circ$}$ for every $x\in\mathbb{N}$
     and $\circ\in\{+,*\}$.
@@
 
+    We also define the initial priority function to give numbers a priority of $\infty$, and operations a priority corresponding to their precedence in the order of operations,
+    where smaller priorities have less precedence.
+    Namely, we give $+$ a priority of $1$ and $*$ a priority of $2$.
+
     Now, consider the expression $1*2+3$. 
-After assigning priorities, we obtain
-the following expression:
+    After assigning priorities, we obtain the following expression:
+    $\mbt1_\infty \mbt*_2 \mbt2_\infty \mbt+_1 \mbt3_\infty$.
     \yoni{The last sentence is an example of how the assignment
@@
     if they become a formal part of the framework, from the very beginning.}
-$\mbt1_\infty \mbt*_2 \mbt2_\infty \mbt+_1 \mbt3_\infty$.
+    \ari{Done, thoughts?}
     We now get:
@@
     \end{multline*}
+
 \end{example}
@@
 the priority of each token and then beta-reduce.
-But this complicates how we allocate priorities, 
+But this complicates how we allocate priorities, as we can no longer statically assign priorities using a simple priority function $\pi$ but must instead design a more complicated algorithm in order
+to allow for generalization.
 \yoni{would u say that currently we can set priorities statically, but with
@@
 If so, I would mention that.}
-and does not generalize well.
+\ari{Done, thoughts?}
 Furthermore the notion introduced in the next section, partial reduction, generalizes well and is used as a basepoint in the
@@
 \begin{example}
-We define our alphabet $\Sigma$ to be the same as in~\Cref{sec:firstattempt}.
+
+    We define our alphabet $\Sigma$ to be the same as in~\ref{sec:firstattempt}.
     Now we define our initial $\beta$-reducer which will support parentheses.
@@
     was not defined. Also, notice that some tokens get priorities and some don't, which is a bit confusing.}
+    \ari{Both points are now explained in the previous section, is this good?}
     \[ \pi(\mbt x) = \infty\hbox{ for numbers \ttt x},\quad \pi(\mbt+) = 1,\quad \pi(\mbt*) = 2,\quad \pi(\mbt() = \infty,\quad \pi(\mbt)) = 0 \]
@@
 Here we abstract away the value of each term; the value of the new term given by the initial $\beta$-reducer will be determined by an output function.
-Our alphabet of types $\Sigma$, instead of having a token for every number, will just have tokens for types (number, operator, etc.).
+Our alphabet of tokens $\Sigma$, instead of having a token for every number, will just have tokens for types (number, operator, etc.).
 We also have another set $\values$, the \textit{universe of values}.
@@
 \yoni{I would consider using the word "kinds" instead of "types". "kinds" is not great either, but types are stuff like int, real, int->int, which i am not sure is appropriate here.}
+\ari{I understand this, I'm not such a fan of ``kinds'' but I guess using ``types'' isn't really faithful to its understanding in CS.
+I'd like to maybe discuss this with you more when we meet.}
 
@@
 \begin{example}
+
     Define $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$.
@@
     \end{itemize}
+    In the first point $(v,f\mapsto(v,f))$ means that the function to compute the value of the new token is one which takes two values $v,f$ and outputs a new value which is the pair $(v,f)$.
 
@@
     }
+    \ari{Explained, is this good?}
     \yoni{What happened to the squares around the tokens in this example?}
@@
     So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
-So it is converted to
+    In this example allocating initial priorities is not so simple: the priorities depend not on the type of the token in $\Sigma$, but the token in the stream given to the lexer.
+    So we will assume the lexer allocates priorities similar to the previous example in the previous section.
+    This issue will be resolved in the next section.
+    So the stream is converted to
     \[ \NUM_\infty(2)\ \OP_2(*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \]
+    We no longer need boxes to distinguish between tokens because the individual tokens and so we will no longer use them.
 
     \yoni{Here we are missing a formal assignment of priorities.}
+    \ari{I added a few sentences explaining this.}
 
@@
 \end{enumerate}
-The sets $\termp,\terms,\terma$ are disjoint, and we place no such restriction on $\values$.
+The sets $\termp,\terms,\terma$ must be disjoint, and we place no such restriction on $\values$ (it can include elements from $\termp,\terms,\terma$).
 \yoni{The comment about no restrictions for $\values$ is unclear.}
@@
 Each base type has an associated arity in $\bbN$; let $\Sigma^n$ denote the set of all base types of arity $n$.
-Then we define $\terms$ as follows:
+Then we define $\terms$ inductively as follows:
+\begin{enumerate}
+    \item for $\sigma\in\Sigma^0$; $\sigma$ is a type term.
+    \item for $\sigma\in\Sigma^n$ for $n>0$, and $\tau_1,\dots,\tau_n$ type term; $\sigma\tau_1\cdots\tau_n$ is a type term.
+    \item finally for $\sigma\in\Sigma^\infty$ and any $n\geq0$, if $\tau_1,\dots,\tau_n$ are type terms so is $\sigma(\tau_1\cdots\tau_n)$.
+\end{enumerate}
+We can write this definition as:
 \[ \terms \ccoloneqq \Sigma^0 \mid \Sigma^1\terms \mid \Sigma^2\terms\terms \mid \Sigma^3\terms\terms\terms \mid \cdots \mid \Sigma^\oo(\terms\cdots\terms) \]
@@
 Do we actually need $\Sigma^{\infty}$ BTW? }
+\ari{Done.
+And yeah, $\Sigma^\infty$ is important since products have a variable number of parameters.}
 
@@
 $\mathcal A^n$ is the set of atomic abstract terms of a set arity $n$.
-Then $\terma$ is defined by
+Then $\terma$ is defined as follows:
+\begin{enumerate}
+    \item for $\sigma\in\c A^0$; $\sigma$ is an abstract term.
+    \item for $\sigma\in\c A^n$ for $n>0$, and $\tau_1,\dots,\tau_n\in\terms^\epsilon$; $\sigma(\tau_1,\dots,\tau_n)$ is an abstract term.
+        When writing such an abstract term, we may omit parentheses and commas if there is no confusion.
+    \item for $\sigma\in\c A^\infty$ and any $n\geq0$, for $\tau_1,\dots,\tau_n\in\terms^\epsilon$; $\sigma(\tau_1,\dots,\tau_n)$ is an abstract term.
+\end{enumerate}
+Alternatively:
 \[ \terma \ccoloneqq \mathcal A^0\mid\mathcal A^1\terms^\epsilon\mid\mathcal A^2\terms^\epsilon\terms^\epsilon\mid\mathcal A^3\terms^\epsilon\terms^\epsilon\terms^\epsilon\mid\cdots\mid
 \mathcal A^\infty\terms^\epsilon\cdots\terms^\epsilon \]
-Here we allow inputs to atomic abstract terms to be empty (i.e. $\terms^\epsilon$ instead of just $\terms$).
+Here we allow parameters to atomic abstract terms to be empty (i.e. $\terms^\epsilon$ instead of just $\terms$).
+
+For readability we can put and omit parentheses and commas where convenient, but importantly these sets have the \textit{unique reconstruction property}:
+it is possible to compute their construction from the string itself (i.e. a string cannot be constructed from two different patterns).
+This property is not hard to prove.
+\yoni{If I understand correctly you are assuming this property, right? If so, please say so.}
+\ari{No, by the definition of these sets, they have this property.}
 
 \begin{example}
-If $\Sigma^{0}=\{\NUM\}$,
-$\Sigma^{1}=\{\LIST\}$, and $\Sigma^{\infty}=\{\PRODUCT\}$
-and $\Sigma=\Sigma^0\cup\Sigma^1\cup\Sigma^\infty$,
- and $\mathcal A=\set{\OP}$, 
+
+    If $\Sigma^{0}=\{\NUM\}$, $\Sigma^{1}=\{\LIST\}$, and $\Sigma^{\infty}=\{\PRODUCT\}$, and $\Sigma=\Sigma^0\cup\Sigma^1\cup\Sigma^\infty$, and $\c A=\mathcal A^1=\set{\OP}$, 
     \yoni{plz specify the "arity" of $\OP$.}
+    \ari{Done}
     then
-\[ \NUM,\LIST\NUM,\PRODUCT(\LIST\NUM\NUM)\in\terms,\qquad \OP\LIST\NUM\in\terma \]
+    \[ \NUM,\LIST\ \NUM,\PRODUCT(\LIST\ \NUM\ \NUM)\in\terms,\qquad \OP\ \LIST\ \NUM\in\terma \]
     \yoni{I think it is better to put a space between these, especially
     in $\LIST\NUM\NUM$.}
+    \ari{Done}
+
 \end{example}
-For readability we can put parentheses and commas where necessary, but importantly these sets have the \textit{unique reconstruction property}:
-\yoni{If I understand correctly you are assuming this property, right? If so, please say so.}
-it is possible to compute their construction from the
-string itself (i.e. a string cannot be constructed from two different patterns).
 
@@
 \begin{enumerate}
-    \item $\term=\termp\cup\terms\cup\terms$, 
+    \item $\term=\termp\cup\terms\cup\terma$, 
         \yoni{Should the last $\terms$ be $\terma$?}
+        \ari{Yeah}
     the set of \textit{terms},
@@
     \item $\expi=\termi\times\values$, the set of \textit{internal expressions},
-    \item $\exp=\termi\cup\termp$, the set of \textit{expressions}.
+    \item $\exp=\expi\cup\termp$, the set of \textit{expressions}.
 \end{enumerate}
 Finally we define $\oe=\exp\times\bbZ$.
-Elements of $\oe$ will be written like $\sigma_n(v)$ as before; 
+Elements of $\oe$ will be written like $\sigma_n(v)$ as before: $\sigma$ is the term (printable or internal), $n$ the priority, and $v$ the value.
+In the case that $\sigma$ is a printable term, we still use this notation even though the value does not exist: so we just take it to be empty.
 \yoni{But elements of $\oe$ seems like pairs to me, not triplets.}
-for printable terms we act as if the value is empty.
+\ari{There was a slight mistake in the definition of $\exp$.}
+% for printable terms we act as if the value is empty.
 \yoni{Please clarify what does acting as if... means.}
+\ari{Rewritten}
 
 The \textit{initial priority function} is a function $\pi\colon\termp\longto\bbZ$.
-This can be naturally extended to $\pi\colon\termp^*\longto\oe^*$.
+This can be naturally extended to $\pi\colon\termp^*\longto\oe^*$ as before: $\pi(\sigma^1\cdots\sigma^n)=\sigma^1_{\rho_1}\cdots\sigma^n_{\rho_n}$ where $\rho_i=\pi(\sigma^i)$.
 \yoni{please show how.}
+\ari{Done}
 
@@
 \yoni{Using something and defining it later should be avoided...}
+\ari{We can move the definition up, but I felt like it messed with the flow.}
 
 \yoni{Generally, this section is hard to follow. I suggest adding plenty of examples, after every new definition / concept, please add an appropriate small example.}
+\ari{I'm not really sure what kinds of examples to add; I added a big example at the end that we can potentially split up?
+But I feel like we need all of the definitions to give a good example.}
 
@@
 \end{enumerate}
+
+\ari{Adding the following example:}
+
+\begin{example}
+
+    Let $\Sigma^0=\set\NUM$, $\Sigma^1=\set\LIST$, $\Sigma=\Sigma^0\cup\Sigma^1$; and finally we have $\c A^0=\set{\COLON,\LBRACK,\RBRACK}$, $\c A^1=\set\OP$, $\c A=\c A^0\cup\c A^1$.
+    Let our set of printable terms consists of numbers as well as $\set{\mbt+,\mbt*,\mbt;,\mbt[,\mbt]}$.
+    For readability, we again return to putting boxes around printable terms (but only \textit{printable} terms).
+    Our initial priority function is
+    \[ \pi(\mbt n) = \infty,\ \pi(\mbt+) = 1,\ \pi(\mbt*) = 2,\ \pi(\mbt;) = 0,\ \pi(\mbt[) = 0,\ \pi(\mbt]) = 0 \]
+    We can transform the following stream \ttt{[1 + 2; 3 * 4;]} (which should be interpreted as the list $(3,12)$) into the stream of expressions:
+    \[ \mbt[_0\ \mbt1_\infty\ \mbt+_1\ \mbt2_\infty\ \mbt;_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0 \]
+
+    Our initial state will map numbers $\mbt n$ to $\NUM(n)$; operations $\mbt{$\circ$}$ to $\OP(\circ)$; $\mbt;$ to $\COLON$; $\mbt[$ to $\LBRACK$; and $\mbt]$ to $\RBRACK$.
+
+    Our initial $\beta$-reducer is defined as follows: (we will neglect the input state and output state and string in the output of the $\beta$-reducer as they are held constant and empty, respectively.
+    That is, instead of $\hb(\dots)=(\dots,(v,u,\bar s\mapsto\dots,\epsilon,\bar s))$ we will just write $\hb(\dots)=(\dots,(v,u\mapsto\dots))$.)
+    \begin{itemize}
+        \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f,\mapsto(v,f)))$
+        \item $\hb(\OP\NUM,\NUM)=(\NUM,\snd,((v,f),u,\bar s\mapsto f(v,u)))$
+        \item $\hb(\OP\NUM,\OP\NUM)=(\OP\NUM,\snd,((v,f),(u,g)\mapsto (f(v,u),g)))$
+        \item $\hb(\sigma,\COLON)=(\sigma,\zero,(v,\_\mapsto v))$
+        \item $\hb(\LBRACK,\sigma)=(\LBRACK\sigma,\fst,(\_,v\mapsto(v)))$
+        \item $\hb(\LBRACK\sigma,\sigma)=(\LBRACK\sigma,\fst,(\ell,v\mapsto\ell::(v)))$
+        \item $\hb(\LBRACK\sigma,\RBRACK)=(\LIST\sigma,\pinfty,(\ell,\_\mapsto\ell))$
+    \end{itemize}
+
+    We now begin the $\beta$-reduction: 
+    \ialign to\hsize{\rlap{$#$}\tabskip=0pt plus 1fil&$\to#$\hfil\tabskip=\z@\cr
+    \mbt[_0\ \mbt1_\infty\ \mbt+_1\ \mbt2_\infty\ \mbt;_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK_0\ \mbt1_\infty\ \mbt+_1\ \mbt2_\infty\ \mbt;_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK_0\ \NUM_\infty(1)\ \mbt+_1\ \mbt2_\infty\ \mbt;_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK_0\ \NUM_\infty(1)\ \OP_1(+)\ \mbt2_\infty\ \mbt;_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK_0\ \OP\NUM_1(1,+)\ \mbt2_\infty\ \mbt;_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK_0\ \OP\NUM_1(1,+)\ \NUM_\infty(2)\ \mbt;_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK_0\ \OP\NUM_1(1,+)\ \NUM_\infty(2)\ \COLON_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK_0\ \OP\NUM_1(1,+)\ \NUM_0(2)\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK_0\ \NUM_0(3)\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3))\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3))\ \NUM_\infty(3)\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3))\ \NUM_\infty(3)\ \OP_2(*)\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3))\ \OP\NUM_2(3,*)\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3))\ \OP\NUM_2(3,*)\ \NUM_\infty(4)\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3))\ \OP\NUM_2(3,*)\ \NUM_\infty(4)\ \COLON_0\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3))\ \OP\NUM_2(3,*)\ \NUM_0(4)\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3))\ \NUM_0(12)\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3,12))\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3,12))\ \RBRACK_0\cr
+        &\LIST\NUM_0((3,12))\cr
+    }
+
+\end{example}
 

### 2024-12-17 Paper/paper.tex
Cleaned up comments

@@
     $\mbt1_\infty \mbt*_2 \mbt2_\infty \mbt+_1 \mbt3_\infty$.
-    \yoni{The last sentence is an example of how the assignment
-    of priorities feels a bit out of context here. This could improve
-    if they become a formal part of the framework, from the very beginning.}
-    \ari{Done, thoughts?}
     We now get:
@@
 to allow for generalization.
-\yoni{would u say that currently we can set priorities statically, but with
-the idea of increasing priorities we would have to set them dynamically?
-If so, I would mention that.}
-\ari{Done, thoughts?}
 Furthermore the notion introduced in the next section, partial reduction, generalizes well and is used as a basepoint in the
@@
 
-    \yoni{I don't really understand it. For example, what is 
-    $(v,f\mapsto(v,f))$? Perhaps introducing some more definitions
-    and details can help clarify?
-    }
-    \ari{Explained, is this good?}
-    \yoni{What happened to the squares around the tokens in this example?}
-
     So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
@@
     We no longer need boxes to distinguish between tokens because the individual tokens and so we will no longer use them.
-
-    \yoni{Here we are missing a formal assignment of priorities.}
-    \ari{I added a few sentences explaining this.}
 
@@
 
-\yoni{I don't think this is very standard. Perhaps you can define
- it inductively: Every element of $\Sigma^0$ is in $T_{\Sigma}$,
-  and for eac $n$, if $t_i\in T_{\Sigma}$ for each $1\leq i \leq n$
-  then $f t_1 \ldots t_n \in T_{\Sigma}$ for each $f\in\Sigma^{n}$,
-  and then for each $g\in\Sigma^{\infty}$ and any $n$, similarly...
-Do we actually need $\Sigma^{\infty}$ BTW? }
-\ari{Done.
-And yeah, $\Sigma^\infty$ is important since products have a variable number of parameters.}
-
 To construct $\terma$ we start with $\mathcal A$, a set of \textit{atomic abstract terms}, again each with an arity in $\bbN$.
@@
 This property is not hard to prove.
-\yoni{If I understand correctly you are assuming this property, right? If so, please say so.}
-\ari{No, by the definition of these sets, they have this property.}
 
@@
     If $\Sigma^{0}=\{\NUM\}$, $\Sigma^{1}=\{\LIST\}$, and $\Sigma^{\infty}=\{\PRODUCT\}$, and $\Sigma=\Sigma^0\cup\Sigma^1\cup\Sigma^\infty$, and $\c A=\mathcal A^1=\set{\OP}$, 
-    \yoni{plz specify the "arity" of $\OP$.}
-    \ari{Done}
     then
     \[ \NUM,\LIST\ \NUM,\PRODUCT(\LIST\ \NUM\ \NUM)\in\terms,\qquad \OP\ \LIST\ \NUM\in\terma \]
-    \yoni{I think it is better to put a space between these, especially
-    in $\LIST\NUM\NUM$.}
-    \ari{Done}
 
@@
 \begin{enumerate}
-    \item $\term=\termp\cup\terms\cup\terma$, 
-        \yoni{Should the last $\terms$ be $\terma$?}
-        \ari{Yeah}
-    the set of \textit{terms},
+    \item $\term=\termp\cup\terms\cup\terma$, the set of \textit{terms},
     \item $\termi=\terms\cup\terma$, the set of \textit{internal terms},
@@
 In the case that $\sigma$ is a printable term, we still use this notation even though the value does not exist: so we just take it to be empty.
-\yoni{But elements of $\oe$ seems like pairs to me, not triplets.}
-\ari{There was a slight mistake in the definition of $\exp$.}
-% for printable terms we act as if the value is empty.
-\yoni{Please clarify what does acting as if... means.}
-\ari{Rewritten}
 
@@
 This can be naturally extended to $\pi\colon\termp^*\longto\oe^*$ as before: $\pi(\sigma^1\cdots\sigma^n)=\sigma^1_{\rho_1}\cdots\sigma^n_{\rho_n}$ where $\rho_i=\pi(\sigma^i)$.
-\yoni{please show how.}
-\ari{Done}
 

### 2024-12-28 Paper/paper.tex
Updated paper for optimization

@@
 
+\section{Optimizing the Algorithm}
+
+Iterative reduction is an $\Theta(n^2)$-time algorithm under the assumption that every reduction reduces the number of tokens by one: (which is the case in the language provided), and no reduction yields a string of printable terms (e.g.
+when creating parse trees).
+This is because a string of the form $\s s^1_1\s s^2_2\cdots\s s^n_n$ will take $n$ applications of the $\beta$-reducer, each application taking $n-k$ iterations (since $\beta$ is recursive).
+
+But we can make this more efficient; indeed, iterative reduction can be made into a $\Theta(n)$-time algorithm.
+Currently, the implementation of iterative reduction is as follows (assuming the existence of $\hb$):
+
+\def\bhelp{\hbox{\tencsc$\hb$-helper}}
+\begin{algo}[h]
+\algorithm
+    \Function{$\beta$-helper}{$\xi,\bar s$}
+        \State $\xi = \sigma^1_n(u)\sigma^2_m(v)\xi'$
+        \lIf{$\sigma^1\in\termp$} \Return $\gen{\bar s(\sigma^1)_n\sigma^2_m(v)\xi'}[\bar s']$
+        \If{$\hb(\sigma^1,\epsilon)=(\tau,\rho,f)$ is defined}
+            \State $(w,\zeta,\bar s')\gets f(u,\_,\bar s)$
+            \State $\gen{\xi''}[\bar s'']\gets\beta^*\gen{\pi(\zeta)}[\bar s']$
+            \State $n'\gets \rho(n,\_)$
+            \State\Return $\gen{\tau_{n'}(w)\xi''\xi'}[\bar s'']$
+        \ElseIf{$n\geq m$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho,f)$ is defined}
+            \State $(w,\zeta,\bar s')\gets f(u,v,\bar s)$
+            \State $\gen{\xi''}[\bar s'']\gets\beta^*\gen{\pi(\zeta)}[\bar s']$
+            \State $n'\gets \rho(n,m)$
+            \State\Return $\gen{\tau_{n'}(w)\xi''\xi'}[\bar s'']$
+        \Else
+            \State\Return undefined
+        \EndIf
+    \EndFunc
+    \nonum\State
+    \Function{$\beta$}{$\xi,\bar s$}
+        \If{$\xi=\epsilon$} \Return $\gen\xi[\bar s]$
+        \ElseIf{$\bhelp\gen\xi[\bar s]=\gen{\xi'}[\bar s']$ is defined}
+            \Return $\gen{\xi'}[\bar s']$
+        \Else
+            \State $\xi=\sigma_n(u)\xi'$
+            \State $\gen{\xi''}[\bar s']\gets\beta\gen{\xi'}[\bar s]$
+            \State\Return $\gen{\sigma_n(u)\xi''}[\bar s']$
+        \EndIf
+    \EndFunc
+    \nonum\State
+    \Function{$\beta^*$}{$\xi,\bar s$}
+        \State $\gen{\xi'}[\bar s']\gets\beta\gen\xi[\bar s]$
+        \If{$\xi==\xi'$} \Return $\gen\xi[\bar s]$
+        \Else\ \Return $\beta^*\gen{\xi'}[\bar s']$
+        \EndIf
+    \EndFunc
+\ealgorithm
+\end{algo}
+
+All \bhelp{} does is check if $\hb$ can be applied to the current token(s), it cannot be optimized in a significant manner.
+What we will optimize is the manner in which we apply $\beta$.
+We can think of our algorithm as having a pointer to the current token in the stream.
+If \bhelp{} can be applied at this pointer, then we alter the stream accordingly, and move the pointer to the beginning, and
+repeat.
+
+But notice that if we have a stream of tokens $\s s^1\cdots\s s^k\cdots\s s^n$ (the current token is $\s s^k$), and we can apply \bhelp{}, we get $\s s^1\cdots\s s^{k-1}\s t\cdots\s s^n$.
+Now, the fact that the pointer was at $\s s^k$ means that it could not have reduced any of the tokens $\s s^1,\dots,\s s^{k-1}$.
+So going back to the beginning will waste time: we don't need to check indices $1,\dots,k-2$.
+Instead what we will do is move the pointer back one (as now it is possible that $\s s^{k-1}$ and $\s t$ can be reduced).
+
+And if $\hb$ cannot be applied at the current pointer, we increment the pointer to the next token (since this is the final rule
+in iterative reduction).
+
+\begin{algo}
+\algorithm
+    \Function{$\hb^*$}{$\xi,\bar s$}
+        \If{$\xi=\epsilon$} \Return $(\gen\xi[\bar s],\bot)$
+        \ElseIf{$\bhelp\gen\xi[\bar s]=\gen{\xi'}[\bar s']$ is defined} \Return $(\gen{\xi'}[\bar s'],\top)$
+        \Else
+            \State $\xi=\sigma^1_n(u)\xi'$
+            \State $(\gen{\xi''}[\bar s'],\s X)\gets\hb^*\gen{\xi'}[\bar s]$
+            \lIf{$\s X$} \Return $\hb^*\gen{\sigma^1_n(u)\xi''}[\bar s']$
+            \lElse \Return $(\gen{\sigma^1_n(u)}[\bar s'],\bot)$
+        \EndIf
+    \EndFunc
+    \nonum\State
+    \Function{$\beta^*$}{$\xi,\bar s$}
+        \State $(\gen{\xi'}[\bar s'],\s X)\gets\hb^*\gen\xi[\bar s]$
+        \lIf{$\s X$} \Return $\beta^*\gen{\xi'}[\bar s']$
+        \lElse \Return $\gen{\xi'}[\bar s']$
+    \EndFunc
+\ealgorithm
+\end{algo}
+
+The algorithm was explained intuitively above, we now focus on how this implementation is faithful to the explanation.
+Firstly, $\beta^*$ calls $\hb^*$, and $\hb^*$ checks if $\bhelp$ can be applied at the current pointer.
+If it can, then it applies it and returns (essentially moving the pointer back one).
+It also returns $\top$, meaning it successfully applied a reduction.
+
+If it cannot apply $\bhelp$, then it moves forward one and applies $\hb^*$.
+Now, if this returns $\bot$, this means that no reductions we performed.
+This means that all we know is that the next reduction is before the current token, so we have to go back to the start.
+Thus we continue returning $\bot$ to the end.
+If it returned $\top$, then there was a reduction performed at the subsequent token, so we try to continue reducing at the current
+token (since we know we cannot reduce previous tokens).
+
+\begin{theorem}
+
+    Assuming $\hb$ is $O(1)$, no reductions produce a stream of printable tokens (as is the case with creating parse trees), and
+    every application of $\hb$ reduces the number of tokens by one, then this algorithm is $O(n)$.
+
+\end{theorem}
+
+\begin{proof}
+
+    Without loss of generality, we will assume that we always reduce a single token (i.e. we never perform
+    $\hb(\sigma^1,\sigma^2)$, instead always performing $\hb(\sigma^1,\epsilon)$).
+    And we will ignore states, since their change does not affect time complexity (since $\hb$ is $O(1)$).
+
+    We will alter our algorithm slightly: for every token $\sigma$, define $\#\sigma=2$.
+    Now, when we call $\hb^*(\sigma^1\cdots\sigma^n)$, we decrement $\#\sigma^1$ by $1$.
+    If $\hb(\sigma^1,\epsilon)$ is defined, then we increment $\#\sigma^0$ by $1$ ($\sigma^0$ is the token
+    before $\sigma^1$, or $\sigma^2$ if $\sigma^1$ was the first token).
+
+    We claim that $\sigma$ can only be reduced if $\#\sigma=1$, and this means that $\#\xi=\sum_{\sigma\in\xi}\#\sigma$ is
+    decremented by $1$ at each call to $\hb^*$ and never incremented.
+    Thus if we can also show that $\#\sigma$ is always non-negative, then we have shown that we call $\hb^*$ at most $2n$ times,
+    and so the algorithm takes $O(n)$ time.
+
+    Notice that if we go passed $\sigma$, then it will only be revisited if the subsequent token is reduced.
+    Suppose that the subsequent token is not reduced, then we only work backward if the algorithm reaches the end of the string,
+    but then $\hb^*$ will return $\bot$, and it will not be called again.
+    This means that if $\#\sigma=1$ and $\sigma$ is not reduced, then $\#\sigma$ can never become $0$ since it is revisited only if
+    the next token is reduced, which would increment $\#\sigma$ by $1$.
+    \qed
+
+\end{proof}
+
 \section{Comparing Implementations}
@@
 
-Comparing the time it takes both implementations to run this 300 times, we see that it takes iterative reduction 0.0040886 seconds, opposed to the classical implementation which takes 0.0036926 seconds.
+Comparing the time it takes both implementations to run this 300 times, we see that it takes iterative reduction (unoptimized) 0.001055 seconds, optimized iterative reduction 0.000998 seconds, and the classical implementation (menhir) 0.000866 seconds.
+
+\begin{figure}
+
+    \centerline{\includegraphics[width=\hsize]{../Docs/Images/comparer.png}}
+    \caption{Time to compute \ttt{currying.ml}}
+    \label{fig:curry}
+
+\end{figure}
 
@@
 
-    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr.png}}
+    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr-n-n.png}}
     \caption{Time to compute $\hbox{\tencsc Generate-Expression}(n,n)$}
@@
     \medskip
-    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr1.png}}
+    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr-n-1.png}}
     \caption{Time to compute $\hbox{\tencsc Generate-Expression}(n,1)$}
@@
 
-    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr.png}}
+    \centerline{\includegraphics[width=\hsize]{../Docs/Images/fib.png}}
     \caption{Time to compute $\hbox{\tt fib}(x)$}
@@
 \begin{enumerate}
-    \item Iterative reduction is an $O(n^2)$-time algorithm (when no reduction produces a string of printable tokens, and every reduction reduces two tokens to one).
-        This is because a string of the form $\s s^1_1\s s^2_2\cdots\s s^n_n$ will take $n$ applications of the $\beta$-reducer, each application taking $n-k$ iterations (since $\beta$ is recursive).
-        This can be made more efficient: instead of starting the next $\beta$-reduction at the beginning of a string, it can be started at the token before the one just reduced.
-        Meaning once $\s s^{n-1}$ and $\s s^n$ are reduced to give $\s x^{n-1}$, we can begin the next iteration of the $\beta$-reducer at $\s s^{n-2}\s x^{n-1}$.
-        Unfortunately, implementing this did not lead to substantial increases in efficiency.
 

### 2025-01-11 Paper/paper.tex
Added some formal languages to the paper

@@
 
-\begin{theorem}
+\begin{theorem} \label{thrm:complexity}
 
@@
 
+\section{$\beta$-Reducible Languages}
+
+In this section we will try to understand better the class of languages which can be parsed using $\beta$-reduction.
+The definition of $\beta$-reduction in this section will differ slightly from the remainder of this paper, but core ideas will be preserved.
+We will construct a hierarchy of $\beta$-reducible languages, which we will compare with the Chomsky Hierarchy.
+
+\subsection{$\beta$-Reducible Languages}
+
+The simplest class in our hierarchy will be $\beta$-reducible languages, the class of which we will denote by \BR.
+Define a \BR-schema to be a tuple $B=(\Sigma,\Gamma,s,\hb)$ where
+\begin{enumerate}
+    \item $\Sigma$ is a finite alphabet of terminals,
+    \item $\Gamma$ is a finite alphabet of non-terminals (so $\Sigma$ and $\Gamma$ are disjoint),
+    \item $s$ is a state, which is a map $\Sigma\to\Gamma$,
+    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon\Gamma\times(\Gamma\cup\set{\epsilon,\square})\partial\Gamma\cup\set{\epsilon,\bullet}$.
+\end{enumerate}
+Given a \BR-schema, we define a $\beta$-reducer to be a function $\beta\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given a string $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta(\xi)$ as follows:
+\begin{enumerate}
+    \item If $\xi=\bullet$, then $\beta(\xi)=\xi$
+    \item If $\xi=\sigma\in\Gamma$, then $\beta(\xi)=\hb(\sigma,\square)$ if it is defined, and $\sigma$ otherwise,
+    \item If $\xi=\sigma_1\sigma_2\xi'$, then
+        \begin{enumerate}
+            \item $\beta(\xi)=\hb(\sigma_1,\sigma_2)\xi'$ if it is defined, otherwise
+            \item $\beta(\xi)=\hb(\sigma_1,\epsilon)\sigma_2\xi'$ if it is defined, otherwise
+            \item $\beta(\xi)=\sigma_1\beta(\sigma_2\xi')$.
+        \end{enumerate}
+\end{enumerate}
+Then we define $\beta^*(\xi)$ to be the fixed point of successive applications of $\beta$ to $\xi$, i.e. $\beta^*(\xi)=\beta^n(\xi)$ where $\beta^n(\xi)=\beta^{n+1}(\xi)$.
+For a word $w=\sigma_1\cdots\sigma_n=\in\Sigma^*$, we define $s(w)=s(\sigma_1)\cdots s(\sigma_n)\in\Gamma^*$.
+The language \textit{generated} (or \textit{accepted}) by a \BR-schema $B$ is defined to be all the words $w\in\Sigma^*$ such that $\beta^*(s(w))=\bullet$.
+
+Finally, let $\sigma\in\Gamma$: define the set $C(\sigma)=\bigcup_{n\geq1}C_n(\sigma)$ where $C_0(\sigma)=\set\sigma$ and $C_{n+1}(\sigma)=\set{\hb(\tau,\epsilon),\hb(\tau,\square)}[\tau\in C_n(\sigma)]$.
+Call a \BR-schema \textit{valid} if $\sigma\notin C(\sigma)$ for all $\sigma\in\Gamma$.
+A $\beta$-reducible language is one generated by a valid \BR-schema; the class of all $\beta$-reducible languages is denoted \BR.
+The reason for requiring a \BR-schema to be valid is to ensure that the $\beta$-reducer halts.
+Otherwise, suppose $\sigma\in C(\sigma)$, so we could have that $\hb(\sigma,\square)=\sigma$ for example, and then $\beta$ would never halt on the string $\sigma$.
+Validity is not a strong requirement.
+
+According to theorem \ref{thrm:complexity}, all $\beta$-reducible languages can be reduced by their $\beta$-reducer in $O(n)$ time.
+
+Note that the empty word cannot be accepted by a \BR-schema, but we don't really lose any expressive power with this: we can always add another field to the \BR-schema saying whether or not it should accept the empty word.
+
+For example, $\set{a^nb^n}[n\geq1]$ is in \BR.
+We define $B=(\Sigma,\Gamma,s,\hb)$ as follows:
+\begin{enumerate}
+    \item $\Sigma=\set{a,b}$,
+    \item $\Gamma=\set{S_a,S_b,S_{ab},S_{aab}}$,
+    \item $s(a)=S_a$ and $s(b)=S_b$,
+    \item $\hb$ is defined by the rules:
+        \[ S_a,S_b\to S_{ab},\quad S_a,S_{ab}\to S_{aab},\quad S_{aab},S_b=S_{ab},\quad S_{ab},\square\to\bullet \]
+\end{enumerate}
+Indeed, a word of the form $a\cdots ab\cdots b$ will be mapped to $S_a\cdots S_aS_b\cdots S_b$, which will then be reduced to $S_a\cdots S_aS_{ab}S_b\cdots S_b\to S_a\cdots S_{aab}S_b\cdots S_b\to S_a\cdots S_{ab}\cdots S_b$, and so on until $S_{ab}$, which is then reduced
+to $\bullet$.
+
+Now, we claim that every regular language is in \BR{} (we will not consider languages containing the empty string).
+Recall that regular languages are precisely the languages generated by regular expressions.
+A regular expression over $\Sigma$ is a string defined by
+\[ \c R \ccoloneqq \Sigma \gbar (\c R+\c R) \gbar (\c R\cdot\c R) \gbar \c R^+ \]
+The language generated by $r\in\c R$ is $\c L(r)$, defined recursively:
+\begin{enumerate}
+    \item If $r=\sigma\in\Sigma$, then $\c L(r)=\set\sigma$,
+    \item If $r=(r_1+r_2)$, then $\c L(r)=\c L(r_1)\cup\c L(r_2)$,
+    \item If $r=(r_1\cdot r_2)$, then $\c L(r)=\c L(r_1)\cdot\c L(r_2)$,
+    \item If $r=r_1^+$, then $\c L(r)=\c L(r_1)^+$ (where $L^+=\bigcup_{n\geq1}L^n$).
+\end{enumerate}
+(Importantly, these generate all the regular languages not containing the empty string.)
+
+\begin{definition}
+
+    A \BR-schema $B=(\Sigma,\Gamma,s,\hb)$ is in \emph{regular form} if
+    \begin{enumerate}
+        \item $\Gamma$ consists only of symbols of the form $S_\omega$ where $\omega$ is some string (in some alphabet).
+        \item $s$ maps $\sigma$ to $S_\sigma$ for all $\sigma\in\Sigma$.
+        \item $\hb$ has reduction rules only of the form $S_\omega,S_\sigma\to S_{\omega'}$ and $S_\omega,\square\to\bullet$.
+    \end{enumerate}
+
+\end{definition}
+
+\begin{theorem}
+
+    Every regular language can be generated by a \BR-schema in regular form.
+
+\end{theorem}
+
+\begin{proof}
+
+    We proceed by induction on the structure of a regular expression.
+    For singletons, this is trivial: $\set\sigma$ can be generated by the reduction rule $S_\sigma,\square\to\bullet$.
+
+    Now if the regular expression is of the form $r_1+r_2$, by induction there exists \BR-schemas $B_1=(\Sigma,\Gamma_1,s,\hb_1)$ and $B_2=(\Sigma,\Gamma_2,s,\hb_2)$ for $\c L(r_1),\c L(r_2)$ respectively.
+    We define $B=(\Sigma,\Gamma,s,\hb)$ as follows ($\Gamma$ is defined implicitly as all the symbols occurring in the following definition of $\hb$): we start by setting $\hb$ to be the union of $\hb_1$ and $\hb_2$.
+    Furthermore, for every reduction rule $S_\omega,S\to S_{\omega'}$ in $\hb_i$, we add $S_{\mbox\omega_i},S\to S_{\mbox{\omega'}_i}$ to $\hb$.
+    We are assuming that no such symbols are already in use; if they are we can just change the number in the subscript of $\mbox{\omega}_i$.
+
+    If there exists symbols $S_\omega,S_\sigma$ such that $\hb_1(S_\omega,S_\sigma)=S_{\omega_1}$ and $\hb_2(S_\omega,S_\sigma)=S_{\omega_2}$ where $S_{\omega_1}\neq S_{\omega_2}$ then we will set $\hb(S_\omega,S_\sigma)=S_{\omega_1+\omega_2}$.
+    Now, for any symbol of the form $S_{\omega_1+\omega_2}$ that we have just added, we add the following reduction rules: for any $\sigma\in\Sigma$,
+    \begin{enumerate}
+        \item if $\hb_i(S_{\omega_i},S_\sigma)=S_{\omega_i'}$ is defined for only one $i\in\set{1,2}$, then $\hb(S_{\omega_1+\omega_2},S_\sigma)=S_{\mbox{\omega_i'}_i}$,
+        \item if $\hb_i(S_{\omega_i},S_\sigma)=S_{\omega_i'}$ is defined for both $i$s, and $\omega_1'=\omega_2'$, then $\hb(S_{\omega_1+\omega_2},S_\sigma)=S_{\omega_1'}=S_{\omega_2'}$,
+        \item otherwise $\hb(S_{\omega_1+\omega_2},S_\sigma)=S_{\omega_1'+\omega_2'}$.
+    \end{enumerate}
+
+    We continue to regular expressions of the form $r_1\cdot r_2$.
+    Again, we have \BR-schemas $B_1$ and $B_2$ which generate the languages for $r_1,r_2$ respectively.
+    We define a new \BR-schema $B$ as follows: first add all the rules of $\hb_1$ to $\hb$ except those of the form $S_\omega,\square\to\bullet$.
+    For every rule $S_\omega,\square\to\bullet$ in $\hb_1$, and $S_\sigma,S\to S_{\omega'}$ in $\hb_2$, we define new rules
+    \[ S_\omega,S_\sigma \to S_{\omega\sigma},\quad S_{\omega\sigma},S\to S_{\omega\omega'} \]
+    and for every other rule of the form $S_{\omega'},S_\sigma\to S_{\omega''}$ in $\hb_2$ we add the new rule $S_{\omega\omega'},S_\sigma\to S_{\omega\omega''}$.
+    (We assume that $S_{\omega\sigma}$ and $S_{\omega\omega'}$ have not been used yet, these can of course be replaced with other symbols otherwise.)
+
+    And now for regular expressions of the form $r^+$.
+    The language of $r$ has a \BR-schema $B_1$, which has rules of the form $S_\omega,\square\to\bullet$ and $S_\sigma,S_\tau\to S_{\omega'}$.
+    We then add the rules $S_\omega,S_\sigma\to S_{\omega\sigma}$ and $S_{\omega\sigma},S_\tau\to S_{\omega'}$.
+    \qed
+
+\end{proof}
+
+\begin{example}
+
+    For example, let us look at the language $\c L(b(a^++aab))$.
+    We proceed inductively to generate an initial $\beta$-reducer for this language.
+
+    Firstly, $a^+$ is generated by the rules
+    \[ S_a,\square\to\bullet,\quad S_a,S_a\to S_a \]
+    And $aab$ is generated by the rules
+    \[  S_a,S_a \to S_{aa},\quad S_{aa},S_b\to S_{aab},\quad S_{aab},\square\to\bullet \]
+
+    Thus $a^++aab$ is generated by (only adding the relevant rules):
+    \begin{align*}
+        S_a,\square&\to\bullet,\\
+        S_a,S_a&\to S_{a+aa},\\
+        S_{a+aa},S_a&\to S_{\mbox a_1},\\
+        S_{a+aa},S_b&\to S_{\mbox{aab}_2},\\
+        S_{\mbox a_1},\square&\to\bullet,\\
+        S_{\mbox a_1},S_a&\to S_{\mbox a_1},\\
+        S_{\mbox{aab}_2},\square&\to\bullet
+    \end{align*}
+
+    And finally $b(a^++aab)$ is generated by
+    \begin{align*}
+        S_b,S_a&\to S_{ba},\\
+        S_{ba},\square&\to\bullet,\\
+        S_{ba},S_a&\to S_{b(a+aa)},\\
+        S_{b(a+aa)},S_a&\to S_{b\mbox a_1},\\
+        S_{b(a+aa)},S_b&\to S_{b\mbox{aab}_2},\\
+        S_{b\mbox a_1},\square&\to\bullet,\\
+        S_{b\mbox a_1},S_a&\to S_{b\mbox a_1},\\
+        S_{b\mbox{aab}_2},\square&\to\bullet
+    \end{align*}
+
+\end{example}
+
+So we have shown that the class of regular languages, \REG, is a proper subclass of \BR{} (since $\set{a^nb^n}[n\geq1]$ is in \BR{} yet not regular).
+The question then becomes, what is the relationship between \BR{} and the class of context-free languages \CFL{}, or the class of deterministic context-free languages \DCFL?
+
+\ari{I hope to prove or disprove this (hopefully finding some analogy for the pumping lemma for \BR)}
+We conjecture that $\DCFL=\BR$.
+For example, we are unsure but believe that the language of even palindromes, $\set{ww^r}[w\in\set{a,b}^+]$, is not \BR.
+
+\ari{The following section is just me writing down some ideas, far from finished}
+
+\subsection{Multi-$\beta$-Reducible Languages}
+
+An \MBR-schema is a tuple $(\Sigma,\Gamma,s,\set{\hb_i}_{i=0}^n)$ where $\Sigma,\Gamma,s$ are as before and now
+\[ \hb_i\colon\Gamma\times(\Gamma\cup\set{\epsilon,\square})\partial(\Gamma\cup\set{\epsilon,\bullet})\times\set{0,\dots,n} \]
+For each $i\in\set{0,\dots,n}$ we define $\beta_i$ as before, but instead of returning the string it also returns the index returned by the reduction.
+$\beta^*$ will then start using $\beta_0$, and then when the current $\beta_i$ returns $j$, it switches to $\beta_j$.
+
+This is (we believe) stronger than \BR: $\set{ww^r}[w\in\set{a,b}^+]$ is \MBR.
+We construct an \MBR-schema as follows:
+\[ B = \parens{\set{a,b},\set{A,B,A',B'},[a\to A,b\to B],\set{\hb_0,\hb_1,\hb_2,\hb_3,\hb_4}} \]
+Where we have the reduction rules:
+\begin{align*}
+    \hb_0\colon\quad& A,\epsilon\to A',1\\
+        & B,\epsilon\to B',3\\
+    \hb_1\colon\quad& A,\square\to A',2\\
+    \hb_2\colon\quad& A',A'\to\bullet,2\\
+        &A',\epsilon\to\epsilon,2\\
+        &A',\square\to A',0\\
+    \hb_3\colon\quad& B,\square\to B',2\\
+    \hb_4\colon\quad& B',B'\to\bullet,2\\
+        &B',\epsilon\to\epsilon,2\\
+        &B',\square\to B',0\\
+\end{align*}
+Intuitively, $\hb_0$ detects the first character and ``tags'' it, then $\hb_1$ or $\hb_3$ ``tag'' the end character if it is the same as the beginning character.
+$\hb_2$ and $\hb_4$ ``clean up'' the tags: if the two tagged characters are next to each other, then we can try and accept the word.
+Otherwise delete them both.
+
+But \MBR{} is not contained in \CFL{}: $\set{a^nb^nc^n}[n\geq1]\in\MBR$.
+Indeed, we can construct an \MBR-schema as follows:
+\[ B = \parens{\set{a,b,c},\set{A,B,C,X},[a\to A,b\to B,c\to C],\set{\hb_0,\hb_1,\hb_2,\hb_3}} \]
+And we have the reduction rules:
+\begin{align*}
+    \hb_0\colon\quad& B,A\to X,0\\
+        &C,B\to X,0\\
+        &C,\square\to C,1\\
+    \hb_1\colon\quad& A,\epsilon\to\epsilon,2\\
+    \hb_2\colon\quad& B,\epsilon\to\epsilon,3\\
+    \hb_3\colon\quad& C,\square\to\bullet,3\\
+        &C,\epsilon\to\epsilon,1
+\end{align*}
+Intuitively, $\hb_0$ checks that $A$ comes before $B$ which comes before $C$ (as otherwise it places in an $X$, which is never reduced).
+Then $\hb_1$ matches a single $A$, then calls $\hb_2$ to match a single $B$, which calls $\hb_3$ to match a single $C$.
+
 \bibliographystyle{abbrv}

### 2025-01-18 Paper/paper.tex
Fixed proof on REG<BR, added BR(n)

--- 2024-12-28 Paper/paper.tex
@@
 
+\section{$\beta$-Reducible Languages}
+
+In this section we will try to understand better the class of languages which can be parsed using $\beta$-reduction.
+The definition of $\beta$-reduction in this section will differ slightly from the remainder of this paper, but core ideas will be preserved.
+We will construct a hierarchy of $\beta$-reducible languages, which we will compare with the Chomsky Hierarchy.
+
+\subsection{$\beta$-Reducible Languages}
+
+The simplest class in our hierarchy will be $\beta$-reducible languages, the class of which we will denote by \BR.
+Define a \BR-schema to be a tuple $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ where
+\begin{enumerate}
+    \item $\Sigma$ is a finite alphabet of terminals,
+    \item $\Gamma$ is a finite alphabet of non-terminals (generally $\Sigma$ and $\Gamma$ are disjoint, but we don't require this),
+    \item $\square,\blacksquare$ are symbols not in $\Sigma,\Gamma$,
+    \item $\bullet$ is a symbol (which may be in $\Sigma,\Gamma$) or the empty word,
+    \item $s$ is a state, which is a map $\Sigma\to\Gamma$,
+    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon\Gamma\times(\Gamma\cup\set{\epsilon,\blacksquare})\partial\Gamma\cup\set{\epsilon,\bullet}$.
+\end{enumerate}
+Given a \BR-schema, we define a pre-$\beta$-reducer to be a function $\beta'\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given a string $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta'(\xi)$ as follows:
+\begin{enumerate}
+    \item If $\xi=\bullet$, then $\beta'(\xi)=\xi$
+    \item If $\xi=\sigma\in\Gamma$, then $\beta'(\xi)=\hb(\sigma,\blacksquare)$ if it is defined, and $\sigma$ otherwise,
+    \item If $\xi=\sigma_1\sigma_2\xi'$, then
+        \begin{enumerate}
+            \item $\beta'(\xi)=\hb(\sigma_1,\sigma_2)\xi'$ if it is defined, otherwise
+            \item $\beta'(\xi)=\hb(\sigma_1,\epsilon)\sigma_2\xi'$ if it is defined, otherwise
+            \item $\beta'(\xi)=\sigma_1\beta'(\sigma_2\xi')$.
+        \end{enumerate}
+\end{enumerate}
+And we define a $\beta$-reducer to be a function $\beta\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta(\xi)$ as follows:
+\begin{enumerate}
+    \item if $\xi=\epsilon$ and $\hb(\square,\blacksquare)=\tau$ is defined then $\beta(\xi)=\tau$,
+    \item if $\xi=\sigma\xi'$ and $\hb(\square,\sigma)=\tau$ is defined then $\beta(\xi)=\tau\xi'$,
+    \item otherwise $\beta(\xi)=\beta'(\xi)$.
+\end{enumerate}
+Then we define $\beta^*(\xi)$ to be the fixed point of successive applications of $\beta$ to $\xi$, i.e. $\beta^*(\xi)=\beta^n(\xi)$ where $\beta^n(\xi)=\beta^{n+1}(\xi)$.
+For a word $w=\sigma_1\cdots\sigma_n=\in\Sigma^*$, we define $s(w)=s(\sigma_1)\cdots s(\sigma_n)\in\Gamma^*$.
+The language \textit{generated} (or \textit{accepted}) by a \BR-schema $B$ is defined to be all the words $w\in\Sigma^*$ such that $\beta^*(s(w))=\bullet$.
+
+Finally, let $\sigma\in\Gamma$: define the set $C(\sigma)=\bigcup_{n\geq1}C_n(\sigma)$ where $C_0(\sigma)=\set\sigma$ and $C_{n+1}(\sigma)=\set{\hb(\tau,\epsilon),\hb(\tau,\square)}[\tau\in C_n(\sigma)]$.
+Call a \BR-schema \textit{valid} if $\sigma\notin C(\sigma)$ for all $\sigma\in\Gamma$ and $\hb(\square,\blacksquare)$ is either undefined or $\bullet$.
+A $\beta$-reducible language is one generated by a valid \BR-schema; the class of all $\beta$-reducible languages is denoted \BR.
+The reason for requiring a \BR-schema to be valid is to ensure that the $\beta$-reducer halts.
+Otherwise, suppose $\sigma\in C(\sigma)$, so we could have that $\hb(\sigma,\square)=\sigma$ for example, and then $\beta$ would never halt on the string $\sigma$.
+Validity is not a strong requirement.
+
+According to theorem \ref{thrm:complexity}, all $\beta$-reducible languages can be reduced by their $\beta$-reducer in $O(n)$ time.
+
+For example, $\set{a^nb^n}[n\geq1]$ is in \BR.
+We define $B=(\Sigma,\Gamma,s,\hb)$ as follows:
+\begin{enumerate}
+    \item $\Sigma=\set{a,b}$,
+    \item $\Gamma=\set{S_a,S_b,S_{ab},S_{aab}}$,
+    \item $s(a)=S_a$ and $s(b)=S_b$,
+    \item $\hb$ is defined by the rules:
+        \[ S_a,S_b\to S_{ab},\quad S_a,S_{ab}\to S_{aab},\quad S_{aab},S_b=S_{ab},\quad S_{ab},\blacksquare\to\bullet \]
+\end{enumerate}
+Indeed, a word of the form $a\cdots ab\cdots b$ will be mapped to $S_a\cdots S_aS_b\cdots S_b$, which will then be reduced to $S_a\cdots S_aS_{ab}S_b\cdots S_b\to S_a\cdots S_{aab}S_b\cdots S_b\to S_a\cdots S_{ab}\cdots S_b$, and so on until $S_{ab}$, which is then reduced
+to $\bullet$.
+
+\begin{definition}
+
+    A \emph{finite state machine} is a tuple $A=(\Sigma,Q,q_0,F,\delta)$ where
+    \begin{enumerate}
+        \item $\Sigma$ is an alphabet,
+        \item $Q$ is a set of states,
+        \item $q_0\in Q$ is the initial state,
+        \item $F\subseteq Q$ is the set of final states,
+        \item $\delta\colon Q\times\Sigma\to Q$ is the \emph{transition function}.
+    \end{enumerate}
+    We define $\delta^*\colon Q\times\Sigma^*\longto Q$ as follows:
+    \[ \delta^*(q,\epsilon)=q,\qquad \delta^*(q,\sigma\xi)=\delta^*(\delta(q,\sigma),\xi) \]
+    And then we define the \emph{language} of $A$ to be
+    \[ \c L(A) = \set{\xi\in\Sigma^*}[\delta^*(q_0,\xi)\in F] \]
+
+    A language $L$ is \emph{regular} if there exists a finite state machine $A$ such that $\c L(A)=L$.
+    The class of all regular languages is denoted \REG.
+
+\end{definition}
+
+\begin{theorem}
+
+    Every regular language is $\beta$-reducible.
+
+\end{theorem}
+
+\begin{proof}
+
+    Let $A=(\Sigma,Q,q_0,F,\delta)$ be a finite state machine.
+    We define a \BR-schema $B=(\Sigma,\Gamma,q_0,\blacksquare,\bullet,s,\hb)$ as follows: $\Gamma=\Sigma\cup Q$, $s\colon\sigma\mapsto\sigma$, and for $q\in Q$ and $\sigma\in\Sigma$ we define $\hb(q,\sigma)=\delta(q,\sigma)$.
+    And for $q\in F$ we define $\hb(q,\blacksquare)=\bullet$.
+
+    Simple induction will show that $q\xi\to^*\delta^*(q,\xi)$ (where $\xi\to^*\xi'$ means that $\beta^n(\xi)=\xi'$ for some $n$).
+    Indeed $\delta^*(q,\epsilon)=q$ and $q\to^*q$, and $q\sigma\xi\to\hb(q,\sigma)\xi=\delta(q,\sigma)\xi$ and by induction $\to^*\delta^*(\delta(q,\sigma),\xi)=\delta^*(q,\sigma\xi)$ as required.
+    From this it follows easily that (because $\square=q_0$) $\xi\to^*\delta^*(\xi)$ and thus $\beta^*(\xi)=\bullet$ if and only if $\delta^*(\xi)\in F$, if and only if $\xi\in\c L(A)$.
+    Thus $\c L(A)=\c L(B)$ as required.
+    \qed
+
+\end{proof}
+
+So we have shown that the class of regular languages, \REG, is a proper subclass of \BR{} (since $\set{a^nb^n}[n\geq1]$ is in \BR{} yet not regular).
+The question then becomes, what is the relationship between \BR{} and the class of context-free languages \CFL{}, or the class of deterministic context-free languages \DCFL?
+
+\ari{I hope to prove or disprove this (hopefully finding some analogy for the pumping lemma for \BR)}
+We conjecture that $\DCFL=\BR$.
+For example, we are unsure but believe that the language of even palindromes, $\set{ww^r}[w\in\set{a,b}^+]$, is not \BR.
+
+\ari{The following section is just me writing down some ideas, far from finished}
+
+\subsection{Multi-$\beta$-Reducible Languages}
+
+For $n\geq1$, a $\BR(n)$-schema is a tuple $(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\set{\hb_i}_{i=0}^{n-1})$ where $\Sigma,\Gamma,\square,\blacksquare,\bullet,s$ are as before and now
+\[ \hb_i\colon\Gamma\times(\Gamma\cup\set{\epsilon,\square})\partial(\Gamma\cup\set{\epsilon,\bullet})\times\set{0,\dots,n-1} \]
+For each $i\in\set{0,\dots,n-1}$ we define $\beta_i$ as before, but instead of returning the string it also returns the index returned by the reduction.
+$\beta^*$ will then start using $\beta_0$, and then when the current $\beta_i$ returns $j$, it switches to $\beta_j$.
+
+We define \MBR{} to be $\MBR=\bigcup_{n\geq1}\BR(n)$.
+
+This is (we believe) stronger than \BR: $\set{ww^r}[w\in\set{a,b}^+]$ is \MBR.
+We construct a $\BR(5)$-schema as follows:
+\[ B = \parens{\set{a,b},\set{A,B,A',B'},\square,\blacksquare,\bullet,[a\to A,b\to B],\set{\hb_0,\hb_1,\hb_2,\hb_3,\hb_4}} \]
+Where we have the reduction rules:
+\begin{align*}
+    \hb_0\colon\quad& \square,A\to A',1\\
+        & \square,B\to B',3\\
+    \hb_1\colon\quad& A,\square\to A',2\\
+    \hb_2\colon\quad& A',A'\to\bullet,2\\
+        &A',\epsilon\to\epsilon,2\\
+        &A',\square\to A',0\\
+    \hb_3\colon\quad& B,\square\to B',2\\
+    \hb_4\colon\quad& B',B'\to\bullet,2\\
+        &B',\epsilon\to\epsilon,2\\
+        &B',\square\to B',0\\
+\end{align*}
+Intuitively, $\hb_0$ detects the first character and ``tags'' it, then $\hb_1$ or $\hb_3$ ``tag'' the end character if it is the same as the beginning character.
+$\hb_2$ and $\hb_4$ ``clean up'' the tags: if the two tagged characters are next to each other, then we can try and accept the word.
+Otherwise delete them both.
+
+For the language of palindromes $\set{w\in\set{a,b}^*}[w=w^r]$ we can define a $\BR(3)$-schema where $\bullet=\epsilon$ and
+\begin{align*}
+    \hb_0\colon\quad& \square,A\to\epsilon,1\\
+        & \square,B\to\epsilon,2\\
+    \hb_1\colon\quad& A,\blacksquare\to\epsilon,0\\
+    \hb_2\colon\quad& B,\blacksquare\to\epsilon,0\\
+\end{align*}
+
+But \MBR{} is not contained in \CFL{}: $\set{a^nb^nc^n}[n\geq1]\in\MBR$.
+Indeed, we can construct an \MBR-schema as follows:
+\[ B = \parens{\set{a,b,c},\set{A,B,C,X},[a\to A,b\to B,c\to C],\set{\hb_0,\hb_1,\hb_2,\hb_3}} \]
+And we have the reduction rules:
+\begin{align*}
+    \hb_0\colon\quad& B,A\to X,0\\
+        &C,B\to X,0\\
+        &C,\square\to C,1\\
+    \hb_1\colon\quad& A,\epsilon\to\epsilon,2\\
+    \hb_2\colon\quad& B,\epsilon\to\epsilon,3\\
+    \hb_3\colon\quad& C,\square\to\bullet,3\\
+        &C,\epsilon\to\epsilon,1
+\end{align*}
+Intuitively, $\hb_0$ checks that $A$ comes before $B$ which comes before $C$ (as otherwise it places in an $X$, which is never reduced).
+Then $\hb_1$ matches a single $A$, then calls $\hb_2$ to match a single $B$, which calls $\hb_3$ to match a single $C$.
+
 \bibliographystyle{abbrv}

### 2025-01-21 Paper/paper.tex
fix error

@@
 \section{Introduction}
-\yoni{comment}
 Nowadays, most programming language interpreters follow the following pipeline (see \cite{dragon})
@@
     \item Parsing: convert the parsed tokens into a \textit{parse tree}: a tree which represents the code according to the language's grammar.
-\yoni{comment}
     \item Interpreting: recursively descend the tree and execute the code according to the instructions embedded in the tree.
@@
 We propose a novel method for the process of interpretation, which doesn't use parse trees.
-But our algorithm is not limited to this, and can be altered to produce parse trees as well (see \S\ref{producing-parse-trees}).
+But our algorithm is not limited to this, and can be altered to produce parse trees as well (see $\S$\ref{producing-parse-trees}).
 This means that our algorithm can be utilized as part of the pipeline for compilers and traditional interpreters.
@@
 For example, $\mb{\ttt1}$ and $\mb{\ttt+}$ can be combined together to give a token $\mb{\ttt{1,+}}$.
-\yoni{comment}
 Then if we follow a token of the form $\mb{\ttt{x,+}}$ with a token of the form $\mb{\ttt y}$, the tokens are combined to form a token $\mb{\ttt{x+y}}$.

### 2025-01-21 Paper/paper.tex
undo
new comments

@@
 We propose a novel method for the process of interpretation, which doesn't use parse trees.
-But our algorithm is not limited to this, and can be altered to produce parse trees as well (see $\S$\ref{producing-parse-trees}).
+But our algorithm is not limited to this, and can be altered to produce parse trees as well (see
+
+%%\S\ref{producing-parse-trees}).
 This means that our algorithm can be utilized as part of the pipeline for compilers and traditional interpreters.
@@
 We extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.
-
+\yoni{plz explicitly say that we use $a_b$ to denote $<a,b>$.}
 Suppose $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$, then $\beta(\xi)$ is defined recursively as follows:
@@
 \end{enumerate}
+
+\yoni{I wonder how can this support ternary operators like if then else, integrals, or just custom-defined functions. Maybe like in functional programming, so that functions will also be values?}
 
@@
     And we define our priority function as follows:
-    \yoni{feels out of context, the notion of a priority function
-    was not defined. Also, notice that some tokens get priorities and some don't, which is a bit confusing.}
-    \ari{Both points are now explained in the previous section, is this good?}
     \[ \pi(\mbt x) = \infty\hbox{ for numbers \ttt x},\quad \pi(\mbt+) = 1,\quad \pi(\mbt*) = 2,\quad \pi(\mbt() = \infty,\quad \pi(\mbt)) = 0 \]
@@
 
+\yoni{It would be nice to really show what would have happened if we were to use the previous section's approach.
+The main difference, if i understand correctly, is that this would enforce $\hb(\mb (, \mb{x,)}$ to be $(x,snd)$ instead of $(x,fst)$, right?
+It would be cool to write out which term we would eventually get this way.}
+
 \end{example}
@@
 We also have another set $\values$, the \textit{universe of values}.
+\yoni{plz provide an example for what this universe typically looks like.}
 
@@
 \[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
+
+\yoni{plz intuitively describe what it returns.}
+
 Define $\oP=\oS\times\values$ to be the set of all possible terms,
@@
 a term consists of a type, a priority, and a value.
+\yoni{A term is a common word for something like an expression, separated from priority and type (which is here not really a type). So this naming is problematic.}
 Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the type; $p$ is the priority; and $v$ the value.
+\yoni{Not sure about value either. Let's discuss. Value is typically a number, a string, a concrete function...}
 
@@
     Define $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$.
-    Our initial $\beta$-reducer will is defined as follows:
+    Our initial $\beta$-reducer is defined as follows:
     \begin{itemize}
         \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f\mapsto(v,f))$
+\yoni{This is a bit strange.
+It suggests that $\values\times \values \subseteq \values$, which is impossible.
+We need to properly define the relevant set here, and the domain and range of this function.}
         \item $\hb(\OP\NUM,\NUM)=(\NUM,\snd,((v,f),u\mapsto f(v,u)))$
@@
     \end{itemize}
+\yoni{Here would be a good place to show what is saved compared to the previous approach, maybe there are less lines or something?}
     In the first point $(v,f\mapsto(v,f))$ means that the function to compute the value of the new token is one which takes two values $v,f$ and outputs a new value which is the pair $(v,f)$.
@@
     So we will assume the lexer allocates priorities similar to the previous example in the previous section.
+    \yoni{better to use ref instead of "previous example".}
     This issue will be resolved in the next section.
@@
     We no longer need boxes to distinguish between tokens because the individual tokens and so we will no longer use them.
+    \yoni{"and so" sounds like it misses something before it.}
 
@@
     \end{gather*}
+\yoni{$\OP\NUM_2(2,*)$ should be $\OP\NUM_2((2,*))$ right?}
 
@@
 \ari{We can move the definition up, but I felt like it messed with the flow.}
+\yoni{But also using something and defining it later messes the flow... We can probably find a way that is clear without jumps.}
 
@@
 But I feel like we need all of the definitions to give a good example.}
+\yoni{exactly -- the idea is to break the example to many small examples, so that each concept is clearer.}
 

### 2025-01-29 Paper/paper.tex
Update paper

@@
 We extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.
-\yoni{plz explicitly say that we use $a_b$ to denote $<a,b>$.}
-Suppose $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$, then $\beta(\xi)$ is defined recursively as follows:
+Suppose $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$, where $\sigma_n$ denotes the pair $(\sigma,n)$, a token and a priority.
+Then $\beta(\xi)$ is defined recursively as follows:
 \begin{enumerate}
@@
 
-\begin{example}
+\begin{example} \label{example:partial}
 
@@
 We also have another set $\values$, the \textit{universe of values}.
-\yoni{plz provide an example for what this universe typically looks like.}
+This could be, for example, a set consisting of all numbers, functions $\bb R\times\bb R\longto\bb R$, tuples of numbers, etc.
 
@@
 \[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
-
-\yoni{plz intuitively describe what it returns.}
+That is, the initial $\beta$-reducer takes two types from $\Sigma$ and returns the following:
+\begin{enumerate}
+    \item a new type,
+    \item a function to determine the priority of the new type given the priorities of the original two types,
+    \item and a function to determine the value of the new type given the values of the original two types.
+\end{enumerate}
 
@@
     Define $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$.
+    And we define our universe of values to consist of
+    \begin{enumerate}
+        \item numbers ($\bb R$),
+        \item binary operators (functions $\bb R\times\bb R\to\bb R$),
+        \item pairs consisting of a number and a binary operator.
+    \end{enumerate}
+
     Our initial $\beta$-reducer is defined as follows:
     \begin{itemize}
-        \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f\mapsto(v,f))$
+        \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f\mapsto(v,f))$ (here $v$ is a number, $f$ a binary operator, and the result is a pair),
 \yoni{This is a bit strange.
@@
 We need to properly define the relevant set here, and the domain and range of this function.}
+            \ari{In general this is not an issue, you could for example start with a set $S_0=S$, and define $S_{n+1}=S_n\cup(S_n\times S_n)$, then $\values=\bigcup_nS_n$ has this property.
+            I'm not saying this just to be technical, this is sort of how I imagined $\values$ being defined.
+            In any case, I provided a simpler definition for $\values$ above.}
         \item $\hb(\OP\NUM,\NUM)=(\NUM,\snd,((v,f),u\mapsto f(v,u)))$
@@
 \yoni{Here would be a good place to show what is saved compared to the previous approach, maybe there are less lines or something?}
-    In the first point $(v,f\mapsto(v,f))$ means that the function to compute the value of the new token is one which takes two values $v,f$ and outputs a new value which is the pair $(v,f)$.
 
@@
     In this example allocating initial priorities is not so simple: the priorities depend not on the type of the token in $\Sigma$, but the token in the stream given to the lexer.
-    So we will assume the lexer allocates priorities similar to the previous example in the previous section.
-    \yoni{better to use ref instead of "previous example".}
+    So we will assume the lexer allocates priorities similar to example \ref{example:partial}.
     This issue will be resolved in the next section.
@@
     \[ \NUM_\infty(2)\ \OP_2(*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \]
-    We no longer need boxes to distinguish between tokens because the individual tokens and so we will no longer use them.
-    \yoni{"and so" sounds like it misses something before it.}
+    The boxes are removed from around the tokens because they were an aesthetic quirk we used for readability.
+    Since each token is now easily distinguishable from other tokens, the boxes are no longer necessary.
 
@@
     \item $\Sigma$ is a finite alphabet of terminals,
-    \item $\Gamma$ is a finite alphabet of non-terminals (generally $\Sigma$ and $\Gamma$ are disjoint, but we don't require this),
+    \item $\Gamma$ is a finite alphabet of non-terminals (we don't require $\Sigma,\Gamma$ to be disjoint),
     \item $\square,\blacksquare$ are symbols not in $\Sigma,\Gamma$,
@@
 For example, $\set{a^nb^n}[n\geq1]$ is in \BR.
-We define $B=(\Sigma,\Gamma,s,\hb)$ as follows:
+We define $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ as follows:
 \begin{enumerate}
@@
     \item $\Gamma=\set{S_a,S_b,S_{ab},S_{aab}}$,
+    \item $\square,\blacksquare,\bullet$ are all distinct elements not in $\Sigma,\Gamma$,
     \item $s(a)=S_a$ and $s(b)=S_b$,
@@
         & \square,B\to B',3\\
-    \hb_1\colon\quad& A,\square\to A',2\\
+    \hb_1\colon\quad& A,\blacksquare\to A',2\\
     \hb_2\colon\quad& A',A'\to\bullet,2\\
         &A',\epsilon\to\epsilon,2\\
-        &A',\square\to A',0\\
-    \hb_3\colon\quad& B,\square\to B',2\\
+        &A',\blacksquare\to A',0\\
+    \hb_3\colon\quad& B,\blacksquare\to B',2\\
     \hb_4\colon\quad& B',B'\to\bullet,2\\
         &B',\epsilon\to\epsilon,2\\
-        &B',\square\to B',0\\
+        &B',\blacksquare\to B',0\\
 \end{align*}
@@
         &C,B\to X,0\\
-        &C,\square\to C,1\\
+        &C,\blacksquare\to C,1\\
     \hb_1\colon\quad& A,\epsilon\to\epsilon,2\\
     \hb_2\colon\quad& B,\epsilon\to\epsilon,3\\
-    \hb_3\colon\quad& C,\square\to\bullet,3\\
+    \hb_3\colon\quad& C,\blacksquare\to\bullet,3\\
         &C,\epsilon\to\epsilon,1

### 2025-02-09 Paper/paper.tex
Updated paper, added more examples

@@
 \end{enumerate}
-
-\yoni{I wonder how can this support ternary operators like if then else, integrals, or just custom-defined functions. Maybe like in functional programming, so that functions will also be values?}
 
@@
 
-\yoni{It would be nice to really show what would have happened if we were to use the previous section's approach.
-The main difference, if i understand correctly, is that this would enforce $\hb(\mb (, \mb{x,)}$ to be $(x,snd)$ instead of $(x,fst)$, right?
-It would be cool to write out which term we would eventually get this way.}
+    This example would not be possible using our previous section's approach to $\beta$-reduction, or at least it would be much less efficient.
+    This is because we'd have to define every possible reduction, in particular we'd have to be able to reduce $\mbt($ and $\mbt1$ (since they're both given an initial infinite priority).
+    We could of course do this, and our above example would work.
+    But what of examples like $(\cdots(1+1)\cdots)$?
+    We'd need to reduce an arbitrary number of parentheses.
 
@@
 Here we abstract away the value of each term; the value of the new term given by the initial $\beta$-reducer will be determined by an output function.
-Our alphabet of tokens $\Sigma$, instead of having a token for every number, will just have tokens for types (number, operator, etc.).
+Our alphabet of tokens $\Sigma$, instead of having a token for every number, will just have tokens for \kinds{} (number, operator, etc.).
 We also have another set $\values$, the \textit{universe of values}.
 This could be, for example, a set consisting of all numbers, functions $\bb R\times\bb R\longto\bb R$, tuples of numbers, etc.
-
-\yoni{I would consider using the word "kinds" instead of "types". "kinds" is not great either, but types are stuff like int, real, int->int, which i am not sure is appropriate here.}
-\ari{I understand this, I'm not such a fan of ``kinds'' but I guess using ``types'' isn't really faithful to its understanding in CS.
-I'd like to maybe discuss this with you more when we meet.}
 
@@
 \begin{enumerate}
-    \item a new type,
-    \item a function to determine the priority of the new type given the priorities of the original two types,
-    \item and a function to determine the value of the new type given the values of the original two types.
+    \item a new \kind,
+    \item a function to determine the priority of the new \kind{} given the priorities of the original two \kinds,
+    \item and a function to determine the value of the new \kind{} given the values of the original two \kinds.
 \end{enumerate}
@@
 that is, 
-a term consists of a type, a priority, and a value.
+a term consists of a \kind, a priority, and a value.
 \yoni{A term is a common word for something like an expression, separated from priority and type (which is here not really a type). So this naming is problematic.}
-Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the type; $p$ is the priority; and $v$ the value.
-\yoni{Not sure about value either. Let's discuss. Value is typically a number, a string, a concrete function...}
+Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the \kind; $p$ is the priority; and $v$ the value.
 
@@
     \begin{enumerate}
+        \item \emph{none}: a value which reflects that the token has no value,
         \item numbers ($\bb R$),
@@
         \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f\mapsto(v,f))$ (here $v$ is a number, $f$ a binary operator, and the result is a pair),
-\yoni{This is a bit strange.
-It suggests that $\values\times \values \subseteq \values$, which is impossible.
-We need to properly define the relevant set here, and the domain and range of this function.}
-            \ari{In general this is not an issue, you could for example start with a set $S_0=S$, and define $S_{n+1}=S_n\cup(S_n\times S_n)$, then $\values=\bigcup_nS_n$ has this property.
-            I'm not saying this just to be technical, this is sort of how I imagined $\values$ being defined.
-            In any case, I provided a simpler definition for $\values$ above.}
         \item $\hb(\OP\NUM,\NUM)=(\NUM,\snd,((v,f),u\mapsto f(v,u)))$
@@
     \end{itemize}
-\yoni{Here would be a good place to show what is saved compared to the previous approach, maybe there are less lines or something?}
+    In this approach, as opposed to the approach taken in the previous section, we simplify the reduction rules (especially when it comes to the rules regarding operations).
 
     So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
-    In this example allocating initial priorities is not so simple: the priorities depend not on the type of the token in $\Sigma$, but the token in the stream given to the lexer.
+    In this example allocating initial priorities is not so simple: the priorities depend not on the \kind{} of the token in $\Sigma$, but the token in the stream given to the lexer.
     So we will assume the lexer allocates priorities similar to example \ref{example:partial}.
@@
     \end{gather*}
-\yoni{$\OP\NUM_2(2,*)$ should be $\OP\NUM_2((2,*))$ right?}
+    We remove parentheses for added readability when it leads to no confusion: for example we write $\OP\NUM_2(2,*)$ instead of $\OP\NUM_2((2,*))$.
+    And we don't write the value if it's \emph{none} (e.g. in the case of $\LPAREN_\infty$).
 
@@
 
-We define the following four base sets:
+\subsubsection{Base Sets}
+
+We begin the description of stateful reduction by describing certain sets which must exist for the algorithm.
+We start with the following sets:
 \begin{enumerate}
@@
 The sets $\termp,\terms,\terma$ must be disjoint, and we place no such restriction on $\values$ (it can include elements from $\termp,\terms,\terma$).
-\yoni{The comment about no restrictions for $\values$ is unclear.}
 
@@
 
-The \textit{initial priority function} is a function $\pi\colon\termp\longto\bbZ$.
+\subsubsection{States}
+
+We begin by defining \textit{point-states} to be partial maps $\termp\partial\expi$.
+If $s_1,s_2$ are two point-states then we define their composition to be
+\[ s_1s_2(\sigma) = \begin{cases} s_2(\sigma) & \sigma\in\dom s_2\\ s_1(\sigma) & \sigma\in\dom s_1-\dom s_2\end{cases} \]
+Point-states with finite domain will be denoted by $[\sigma_i\mapsto\varkappa_i]_{i=1}^n$ where $\sigma_1,\dots,\sigma_n$ is the domain and $\varkappa_1,\dots,\varkappa_n$ are their respective images.
+The empty point-state (with the empty domain) will be denoted by $\varnothing$.
+
+A \textit{state} will be a pair of a sequence of point-states and a set of indices.
+This will be denoted by $\bar s=[(s_1,\dots,s_n),\;I=(i_1,\dots,i_k)]$ (we require that $k\leq n$).
+The set of indices represent the \emph{closures} of the state: every function call will push a new point-state and index to the state.
+This way if a function returns another function, it can look at its current closure and return it as well, which would allow the returned function to reference variables in its parent's scope.
+
+We define
+\begin{enumerate}
+    \item $\bar s+s=\bigl[(s_1,\dots,s_n,s),\;I\bigr]$, for $s$ a point-state,
+    \item $\bar s+_cs=\bigl[(s_1,\dots,s_n,s),\;(i_1,\dots,i_k,n+1)\bigr]$, for $s$ a point-state,
+    \item $\pop\bar s=\bigl[(s_1,\dots,s_{n-1}),\;I\bigr]$ if $i_k<n$, otherwise it is defined to be equal to $\bigl[(s_1,\dots,s_{n-1}),\;(i_1,\dots,i_{k-1})\bigr]$,
+    \item $\bar ss=\bigl[(s_1,\dots,s_{n-1},s_ns),\;I\bigr]$ for $s$ a point-state,
+    \item $\bar s(\sigma)=s_1\cdots s_n(\sigma)$ for $\sigma\in\termp$,
+    \item $\bar s_c=s_{i_k}\cdots s_n$,
+    \item if $\sigma\in\termp$ and $\varkappa\in\expi$ then $\bar s\{\sigma\mapsto\varkappa\}=\bigl[(s_1,\dots,s_k[\sigma\mapsto\varkappa],\dots,s_n),\;I\bigr]$ where $k$ is the maximum index such that
+    $\sigma\in\dom s_k$.
+\end{enumerate}
+
+\begin{example} \label{example:state}
+
+    Let $\Sigma^0=\set\NUM$, $\Sigma^1=\set\LIST$, so that $\Sigma=\Sigma^0\cup\Sigma^1$; and $\c A^0=\set{\SEMI,\LBRACK,\RBRACK}$, $\c A^1=\set\OP$, so that $\c A=\c A^0\cup\c A^1$.
+    Now suppose that the printable terms, $\termp$, consists of variable names: $\mb{x_1},\mb{x_2},\dots$, as well as symbols for numbers $\mbt{1},\mbt{2},\dots$, symbols for binary operations $\mbt+,\mbt*$, and the
+    symbols $\mbt;,\mbt[,\mbt]$.
+    For readability, we again return to putting boxes around printable terms (but only \textit{printable} terms).
+
+    The set of values $\values$ is defined as follows:
+    \begin{enumerate}
+        \item \ttt{none} is a value,
+        \item every number is a value,
+        \item every list of non-\ttt{none} values is a value, i.e. if $v_1,\dots,v_n\in\values$ and $v_i\neq\ttt{none}$ then $(v_1,\dots,v_n)\in\values$.
+    \end{enumerate}
+
+    Our initial state, $s_1$ will map numbers $\mbt{n}$ to $\NUM(n)$, binary operations $\mb\circ$ to $\OP(\circ)$, and $\mbt;\mapsto\SEMI,\mbt[\mapsto\LBRACK,\mbt]\mapsto\RBRACK$.
+    The program's state will be $\bar s=[(s_1),\;(1)]$.
+
+    Now suppose our program sets $\mb{x_1}$ to the list $(1,2,3)$, then we will want our program's state to be $\bar s=\bar s[\mb{x_1}\mapsto\LIST\NUM((1,2,3))]$.
+    And suppose now that our program enters a block, then we will set our program's state to $\bar s=\bar s+\varnothing$.
+    Upon exiting the block, we set our program's state to $\bar s=\pop\bar s$.
+
+    Now suppose our program wants to a call a function which has parameters $\mb{x_1},\mb{x_2},\mb{x_3}$ and a closure $s$ with the input $1,2,((3,4),(5,6))$.
+    Then our program's state will be
+    \[ \bar s = (\bar s +_c s)\bigl[\mb{x_1}\mapsto\NUM(1),\mb{x_2}\mapsto\NUM(2),\mb{x_3}\mapsto\LIST(\LIST\NUM)((3,4),(5,6))\bigr] \]
+    We do $+_c$ instead of a normal $+$ because we are creating another scope.
+    Now suppose we define a function inside this scope, then we will need to take $\bar s_c$ and add this to the function so it can access variables from this scope.
+
+\end{example}
+
+\subsubsection{The $\beta$-Reducer}
+
+Now we can describe the initial $\beta$-reducer and its derived $\beta$-reducer.
+
+As before, we have an \textit{initial priority function}, which is a function $\pi\colon\termp\longto\bbZ$.
 This can be naturally extended to $\pi\colon\termp^*\longto\oe^*$ as before: $\pi(\sigma^1\cdots\sigma^n)=\sigma^1_{\rho_1}\cdots\sigma^n_{\rho_n}$ where $\rho_i=\pi(\sigma^i)$.
 
-Finally, our \textit{initial $\beta$-reducer} is a function ($\state$ is a set of functions $\termp\partial\expi$ which we define later)
-\yoni{Using something and defining it later should be avoided...}
-\ari{We can move the definition up, but I felt like it messed with the flow.}
-\yoni{But also using something and defining it later messes the flow... We can probably find a way that is clear without jumps.}
-
-\yoni{Generally, this section is hard to follow. I suggest adding plenty of examples, after every new definition / concept, please add an appropriate small example.}
-\ari{I'm not really sure what kinds of examples to add; I added a big example at the end that we can potentially split up?
-But I feel like we need all of the definitions to give a good example.}
-\yoni{exactly -- the idea is to break the example to many small examples, so that each concept is clearer.}
-
+Finally, our \textit{initial $\beta$-reducer} is a function
 \[ \hb\colon\termi\times\term^\epsilon \partial \termi^\epsilon\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\times\state\partial\values\times\termp^*\times\state) \]
@@
 
-\subsubsection{States}
-
-We begin by defining \textit{point-states} to be partial maps $\termp\partial\expi$.
-If $s_1,s_2$ are two point-states then we define their composition to be
-\[ s_1s_2(\sigma) = \begin{cases} s_2(\sigma) & \sigma\in\dom s_2\\ s_1(\sigma) & \sigma\in\dom s_1-\dom s_2\end{cases} \]
-Point-states with finite domain will be denoted by $[\sigma_i\mapsto\varkappa_i]_{i=1}^n$ where $\sigma_1,\dots,\sigma_n$ is the domain and $\varkappa_1,\dots,\varkappa_n$ are their respective images.
-
-A \textit{state} will be a pair of a sequence of point-states and a set of indices.
-This will be denoted by $\bar s=[(s_1,\dots,s_n),\;I=(i_1,\dots,i_k)]$ (we require that $k\leq n$).
-We define
-\begin{enumerate}
-    \item $\bar s+s=\bigl[(s_1,\dots,s_n,s),\;I\bigr]$, for $s$ a point-state,
-    \item $\bar s+_cs=\bigl[(s_1,\dots,s_n,s),\;(i_1,\dots,i_k,n+1)\bigr]$, for $s$ a point-state,
-    \item $\pop\bar s=\bigl[(s_1,\dots,s_{n-1}),\;I\bigr]$ if $i_k<n$, otherwise it is defined to be equal to $\bigl[(s_1,\dots,s_{n-1}),\;(i_1,\dots,i_{k-1})\bigr]$,
-    \item $\bar ss=\bigl[(s_1,\dots,s_{n-1},s_ns),\;I\bigr]$ for $s$ a point-state,
-    \item $\bar s(\sigma)=s_1\cdots s_n(\sigma)$ for $\sigma\in\termp$,
-    \item $\bar s_c=s_{i_k}\cdots s_n$,
-    \item if $\sigma\in\termp$ and $\varkappa\in\expi$ then $\bar s\{\sigma\mapsto\varkappa\}=\bigl[(s_1,\dots,s_k[\sigma\mapsto\varkappa],\dots,s_n),\;I\bigr]$ where $k$ is the maximum index such that
-    $\sigma\in\dom s_k$.
-\end{enumerate}
-
-\ari{Adding the following example:}
-
 \begin{example}
 
-    Let $\Sigma^0=\set\NUM$, $\Sigma^1=\set\LIST$, $\Sigma=\Sigma^0\cup\Sigma^1$; and finally we have $\c A^0=\set{\COLON,\LBRACK,\RBRACK}$, $\c A^1=\set\OP$, $\c A=\c A^0\cup\c A^1$.
-    Let our set of printable terms consists of numbers as well as $\set{\mbt+,\mbt*,\mbt;,\mbt[,\mbt]}$.
-    For readability, we again return to putting boxes around printable terms (but only \textit{printable} terms).
+    Our base sets and initial state will be defined as in example \ref{example:state}.
     Our initial priority function is
@@
     \[ \mbt[_0\ \mbt1_\infty\ \mbt+_1\ \mbt2_\infty\ \mbt;_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0 \]
-
-    Our initial state will map numbers $\mbt n$ to $\NUM(n)$; operations $\mbt{$\circ$}$ to $\OP(\circ)$; $\mbt;$ to $\COLON$; $\mbt[$ to $\LBRACK$; and $\mbt]$ to $\RBRACK$.
 
@@
         \item $\hb(\OP\NUM,\OP\NUM)=(\OP\NUM,\snd,((v,f),(u,g)\mapsto (f(v,u),g)))$
-        \item $\hb(\sigma,\COLON)=(\sigma,\zero,(v,\_\mapsto v))$
+        \item $\hb(\sigma,\SEMI)=(\sigma,\zero,(v,\_\mapsto v))$
         \item $\hb(\LBRACK,\sigma)=(\LBRACK\sigma,\fst,(\_,v\mapsto(v)))$
@@
         &\LBRACK_0\ \OP\NUM_1(1,+)\ \NUM_\infty(2)\ \mbt;_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
-        &\LBRACK_0\ \OP\NUM_1(1,+)\ \NUM_\infty(2)\ \COLON_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
+        &\LBRACK_0\ \OP\NUM_1(1,+)\ \NUM_\infty(2)\ \SEMI_0\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
         &\LBRACK_0\ \OP\NUM_1(1,+)\ \NUM_0(2)\ \mbt3_\infty\ \mbt*_2\ \mbt4_\infty\ \mbt;_0\ \mbt]_0\cr
@@
         &\LBRACK\NUM_0((3))\ \OP\NUM_2(3,*)\ \NUM_\infty(4)\ \mbt;_0\ \mbt]_0\cr
-        &\LBRACK\NUM_0((3))\ \OP\NUM_2(3,*)\ \NUM_\infty(4)\ \COLON_0\ \mbt]_0\cr
+        &\LBRACK\NUM_0((3))\ \OP\NUM_2(3,*)\ \NUM_\infty(4)\ \SEMI_0\ \mbt]_0\cr
         &\LBRACK\NUM_0((3))\ \OP\NUM_2(3,*)\ \NUM_0(4)\ \mbt]_0\cr
@@
 
+\subsubsection{Priorities}
+
+Now, one may be inclined to ask why we don't add priorities to this $\beta$-reduction.
+It turns out that priorities, under certain assumptions, don't add any power.
+Indeed, suppose we added priorities to our $\beta$-reducer, so now our initial $\beta$-reducer is of the form
+\[ \hb\colon\Gamma\times(\Gamma\cup\set{\epsilon,\blacksquare})\partial(\Gamma\cup\set{\epsilon,\bullet})\times(\bbN\times\bbN\to\bbN) \]
+and we have an initial priority function $\pi\colon\Sigma\to\bbN$.
+Let $F$ be the set of priority functions returned by $\hb$, that is
+\[ F = \set{f\colon\bbN\times\bbN\to\bbN}[\exists\sigma\in\Gamma,\tau\in\Gamma\cup\set{\epsilon,\blacksquare}\colon\hb(\sigma,\tau)=(\alpha,f)] \]
+and let $N=\mathrm{Im}\pi$ be the set of all initial priorities.
+Now let us define $N_0=N$ and $N_{n+1} = N_n\cup\set{f(n,m)}[n\geq m\in N_n, f\in F]$.
+We define $N_\infty=\bigcup_{n\geq0}N_n$.
+If $N_\infty$ is finite, then there exists an equivalent \BR-schema (one without priorities).
+
+The proof of this is simple: define $B=(\Sigma,\Gamma',\square,\blacksquare,\bullet,s',\hb')$ where
+\begin{enumerate}
+    \item $\Gamma'=\Gamma\times N_\infty$ which is still finite.
+        Elements of $\Gamma'$ will be written as $\sigma_n$ instead of $(\sigma,n)$.
+    \item $s'$ maps $\sigma$ to $s(\sigma)_{\pi(\sigma)}$.
+    \item If $\hb(\sigma,\tau)=(\alpha,f)$ then for $n\geq m\in N_\infty$ we define $\hb'(\sigma_n,\tau_m)=\alpha_{f(n,m)}$.
+        Note that by definition $f(n,m)\in N_\infty$ so this is well-defined.
+\end{enumerate}
+It is not hard to see that the $\beta$-reducer and this \BR-schema both generate the same language.
+
+The requirement that $N_\infty$ be finite is a strong condition (for example, $F$ cannot include a function of the form $(n,m)\mapsto n+m$ if $N\neq\set0$).
+But in practice this is not a restrictive condition, for example ILang satisfies this condition: if $F$ consists of functions of the form $k_c$ (the constant map), $\snd,\fst$, then $F_\infty$ will trivially be finite.
+
+\subsubsection{Regular Languages}
+
 \begin{definition}

### 2025-02-09 Paper/paper.tex
Fixed single use of type

@@
 \[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
-That is, the initial $\beta$-reducer takes two types from $\Sigma$ and returns the following:
+That is, the initial $\beta$-reducer takes two \kinds{} from $\Sigma$ and returns the following:
 \begin{enumerate}

### 2025-04-28 Paper/beta-reducible-langs.tex
Added about beta reducability

\documentclass{llncs}

%\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{enumitem}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{graphicx}
%\usepackage{mdframed}
%\usepackage{cleveref}

\def\labelitemi{$\top$}
\def\labelenumi{(\textbf{\arabic{enumi}})}

\usepackage{tikz}
\usetikzlibrary{trees}

\newcounter{algo}

\floatstyle{ruled}
\newfloat{algo}{h}{aux}
\floatname{algo}{Algorithm}

\input preamble

%\newcommand{\yoni}[1]{{\begin{mdframed}[linecolor=red]{\color{red}#1}\end{mdframed}}}
%\newcommand{\ari}[1]{{\begin{mdframed}[linecolor=blue]{\color{blue}#1}\end{mdframed}}}
\def\yoni#1{\par\noindent\fbox{\color{red}#1}\par}
\def\ari#1{\par\noindent\fbox{\color{blue}#1}\par}
\begin{document}

\title{Beta-Reduction}
\author{Ari Feiglin\inst{1}\and Yoni Zohar\inst{1}}
\institute{Bar-Ilan University, Ramat Gan, Israel}

\maketitle

\begin{abstract}

    In this paper we discuss an alternative approach to parsing context-free languages.
    \keywords{Context-Free Languages\and Parsing}

\end{abstract}

\section{$\beta$-Reducability}

In this section we investigate the formal language-theoretical properties of $\beta$-reduction.
In particular we discuss what sorts of languages can be $\beta$-reduced.
The definition of $\beta$-reduction in this section differs slightly from our previous definition, but the core principal
is preserved.
In particular, we will forgo priorities, which as we will see are beneficial implementation-wise, but unnecessary theoretically.

\subsection{The $\beta$-Reduction Hierarchy}

We begin by defining multiple classes of languages.
These are all special cases of the class we will begin to define now.

\begin{definition}

    Let $n>0$ be an integer, then an \emph{$n$-nondeterministic $\beta$-reduction schema} (shortened to $n-\NBR$ schema) is
    a tuple $B=(\Sigma,\Gamma,\square,\top,\hb)$ where
    \begin{enumerate}
        \item $\Sigma$ is a finite alphabet of terminals;
        \item $\Gamma$ is a finite alphabet of variables (we will assume that $\Sigma\subseteq\Gamma$, or that there exists a
            clear mapping of $\Sigma$ into $\Gamma$);
        \item $\square$ is a symbol in $\Gamma$ (the string beginning-marker);
        \item $\top$ is a symbol (potentially in $\Gamma$), or the empty word;
        \item $\hb$ is an \emph{initial $\beta$-reducer}, which is a function
        $$ \hb\colon\Gamma\times(\Gamma\cup\set\epsilon)\to\P((\Gamma\cup\set{\top})^{\leq n}\cup\set\perp) $$
            Where $\perp$ is a symbol distinct from any other previously discussed symbol.
    \end{enumerate}

\end{definition}

Let us focus on the initial $\beta$-reducer.
It takes as input two symbols and outputs a set of strings in $\Gamma\cup\set\top$ of length $\leq n$, and the set may also
contain the symbol $\perp$.
$\top$ is the target string, if we can reduce the string to $\top$ we consider the string part of the language.
And $\perp$ says that you can non-deterministically skip the reduction.

So for example if
$$ \hb(a,b) = \set{c,d,\perp},\quad \hb(b,a) = \set{c} $$
the string $aba$ can be reduced (in a single step) to one of the following:
$$ ca\quad da\quad ac $$
The first two result from reducing the first two characters ($a,b$) to one of their potential products, the third results from
skipping the first reduction (since $\perp\in\hb(a,b)$) and reducing the second two characters ($b,a$).

Let's formally define this process of reduction.
We assume that $\hb$ never returns an empty set (if we want to say that we can't reduce two symbols, we can return $\set\perp$).
Furthermore, we define concatenation of two sets of strings component-wise:
$A\circ B=\set{\xi_1\circ\xi_2}[\xi_1\in A,\xi_2\in B]$.
Where $\circ$ is the concatenation operator, which we will from now on leave out writing explicitly.
Furthermore, concatenation between a string and a set is done in the obvious way: $\xi A=\set\xi A$ and $A\xi=A\set\xi$.
Note that $\varnothing A=A\varnothing=\varnothing$, while $\epsilon A=A\epsilon=A$.

\begin{definition}

    Let $B$ be an $n$-\NBR, then we define its $\beta$-reduction function as follows.
    The function is $\beta\colon(\Gamma\cup\set\top)^*\to\P_{<\omega}((\Gamma\cup\set\top)^*)$ and defined as follows:
    \begin{enumerate}
        \item $\beta(\top)=\set\top$;
        \item $\beta(\sigma)=\hb(\sigma,\epsilon)-\set\perp$ if non-empty, and otherwise $\set\sigma$;
        \item For a string $\sigma_1\sigma_2\xi$, we first add
        $\bigl((\hb(\sigma_1,\sigma_2)-\set\perp)\xi\bigr)\cup\bigl((\hb(\sigma_1,\epsilon)-\set\perp)\sigma_2\xi\bigr)$ to
        the result.
        Then if $\perp\in\hb(\sigma_1,\sigma_2)\cup\hb(\sigma_1,\epsilon)$ we also add $\sigma_1\beta(\sigma_2\xi)$.
    \end{enumerate}
    This extends to a function $\beta\colon\P_{<\omega}((\Gamma\cup\set\top)^*)\to\P_{<\omega}((\Gamma\cup\set\top)^*)$
    in the natural way.
    We then define the relation $\xi\to\xi'$ to mean that $\xi'\in\beta(\xi)$; and we define $\sto$ to be the reflexive and
    transitive closure of $\to$.

    We then define the language of $B$ (or the language \emph{recognized} by $B$) to be
    $$ \L(B) = \set{w\in\Sigma^*}[\square w\sto\top] $$
    We define the class $\nNBR$ to be the class of all languages recognized by $\nNBR$ schemas.
    And we define $\sNBR=\bigcup_n\nNBR$.

\end{definition}

Note that an $\sNBR$-schema may be defined similar to an $\nNBR$-schema, but where the initial $\beta$-reducer returns finite
subsets of $(\Gamma\cup\set\top)^*$ (denoted $\P_{<\omega}((\Gamma\cup\set\top)^*)$).
This is since $\hb$ will return a longest string of some length $n$, and then the schema is an $\nNBR$ schema.

\subsection{Context-Free Languages}

Recall the definition of context-free languages:

\begin{definition}

    A \emph{context-free grammar} is a tuple $G=(\Sigma,V,S,P)$ where
    \begin{enumerate}
        \item $\Sigma$ is a finite alphabet of terminals;
        \item $V$ is a finite alphabet of variables;
        \item $S\in V$ is the initial variable;
        \item $P\subseteq V\times(\Sigma\cup V)^*$ is a set of production rules, where we write $A\vdash_G\alpha$ in place of
        $(A,\alpha)\in P$ (and we may omit the subscript from $\vdash$ when $G$ is understood).
    \end{enumerate}
    We extend the definition of $\vdash$ to a relation $(\Sigma\cup V)^*\times(\Sigma\cup V)^*$, where if
    $\alpha A\beta$ is a string of terminals and variables, $A\in V$ such that $A\vdash\gamma$, then
    $\alpha A\beta\vdash\alpha\gamma\beta$.
    $\vdash^*$ is the reflexive and transitive closure of $\vdash$.
    The language of $G$ is defined to be
    $$ \L(G) = \set{w\in\Sigma^*}[S\vdash^*w] $$
    A language is \emph{context-free} if it is the language of some context-free grammar.
    The class of context-free languages is denoted $\CFL$.

\end{definition}

Our goal of this section is to investigate the relationship between the classes $\CFL$ and $\nNBR,\sNBR$.

\begin{lemma}

    $\CFL\subseteq\sNBR$.

\end{lemma}

\begin{proof}

    Let $G=(\Sigma,V,S,P)$ be a context-free grammar.
    We construct an \sNBR-schema $B=(\Sigma,\Gamma,S,\epsilon,\hb)$, where $\Gamma=V\cup\Sigma\cup\overline\Sigma$ where
    $\overline\Sigma$ is a disjoint copy of $\Sigma$.
    For each production rule $A\vdash\alpha\in P$ we place $\perp,\overline\alpha^r\in\hb(A,\epsilon)$ (where $\overline\alpha$
    is $\alpha$ where $\sigma\in\Sigma$ is replaced with $\overline\sigma\in\overline\Sigma$).
    And for $\sigma\in\Sigma$, $\hb(\overline\sigma,\sigma)=\set{\epsilon}$.

    Note that $A\to^*\overline\alpha\in(V\cup\overline\Sigma)^*$, and so when reducing $A$, we can choose any variable to
    reduce.
    That is, if $A_1\to\alpha_1$ and $A_2\to\alpha_2$, and $A\to^*\gamma_1A_1\gamma_2A_2\gamma_3$, then we have that in a single
    reduction we can either get $\gamma_1\alpha_1\gamma_2A_2\gamma_3$ or we get  $\gamma_1A_1\gamma_2\alpha_2\gamma_3$.
    This is because elements of $\overline\Sigma$ won't be reduced, and all other reductions (of variables) may be skipped, as
    they contain $\perp$.

    Now we claim that for any $n\geq0$ and $A\in V$, if $A\vdash^n\alpha$ iff $A\to^n\overline\alpha^r$.
    This is clear by induction: for $n=0$ this is because $\alpha=A$.
    For $n=1$ this is by definition.
    Now suppose $A\vdash^n\alpha_1B\alpha_3$ and $B\vdash\alpha_2$, so that $A\vdash^{n+1}\alpha_1\alpha_2\alpha_3$.
    Then by induction $A\to^n\alpha_3^rB\alpha_1^r$ and by definition $B\to\alpha_2^r$, so that we have
    $A\to^{n+1}\alpha_3^r\alpha_2^r\alpha_1^r=(\alpha_1\alpha_2\alpha_3)^r$ as required.
    The other direction is proven in a similar way.

    If $w\in\L(G)$, we have that $S\vdash^*w$, and so $Sw\sto\overline w^rw$, which clearly then $\sto\epsilon$.
    So $w\in\L(B)$.

    Now if $Aw\sto\overline\alpha w_2$, then we claim there exists a $w_1$ such that $w=w_1w_2$ and $A\vdash^*w_1\alpha^r$.
    We do this by induction on the number of steps in the reduction, $n$.
    For $n=0,1$ this is clear ($w_1=\epsilon$) in both cases.
    Now suppose $Aw\to^n\overline\alpha w_2$, then $w=w_1w_2$ and $A\vdash^*w_1\alpha^r$.
    If the next reduction is in $\overline\alpha$, this must be a variable reduction, which means that $A\vdash^*w_1(\alpha')^r$
    and $Aw\to^{n+1}\overline\alpha' w_2$, so $w'_1=w_1$ still satisfies the requirement.
    Otherwise, the reduction is of the form
    $$ Aw\to^n\overline\alpha'\overline\sigma\sigma w'_2 \to\overline\alpha'w'_2 $$
    In which case $w=w_1w_2=w_1\sigma w'_2$, so that $w'_1$ must be $w_1\sigma$.
    Now, notice that $A\vdash^*w_1\alpha^r=w_1(\alpha'\sigma)^r=w_1\sigma(\alpha')^r=w'_1(\alpha')^r$ as required.

    So we have that $Sw\sto\epsilon$, so $w_2=\epsilon$ and thus $w_1=w$, meaning $S\vdash^*w$.
    Thus if $w\in\L(B)$, $w\in\L(G)$, as required.
    \qed

\end{proof}

Notice that the parameter of the $\sNBR$-schema is dependent only on the length of the production strings in $G$.
That is $B$ is an $\nNBR$-schema where $n=\max\set{\abs\alpha}[A\vdash_G\alpha]$.
Every context-free grammar has an equivalent in Chomsky normal form (CNF), whose $n$ is $2$.
Thus $\CFL\subseteq\pNBR2$.

In fact, $\sNBR=\pNBR2$; this is not difficult to prove.
Take an $\nNBR$-schema $B=(\Sigma,\Gamma,\square,\top,\hb)$ and define a $\pNBR2$-schema
$B_2=(\Sigma,\Gamma_2,\square,\top,\hb_2)$ as follows: $\Gamma_2$ will contain $\Gamma$ as well as every prefix of strings
returned by $\hb$ (each prefix is distinct, if two strings have the same prefix, or the same string is returned by two different
reductions, we consider them distinct).
Given such a prefix $\xi$, denote the symbol in $\Gamma_2$ by $\mbox\xi$.
Then if $\sigma_1\cdots\sigma_n\in\hb(\tau_1,\tau_2)$, we define
$$ \hb_2(\tau_1,\tau_2) = \mbox{\sigma_1\cdots\sigma_n},\quad
\hb_2(\mbox{\sigma_1\cdots\sigma_k},\epsilon)=\set{\mbox{\sigma_1\cdots\sigma_{k-1}}\sigma_k}\hbox{ for $k\leq n$} $$
If $\perp\hb(\tau_1,\tau_2)$, then $\perp\in\hb_2(\tau_1,\tau_2)$ as well.

\begin{lemma}

    $\sNBR=\pNBR2$

\end{lemma}

Though, the most important result is as follows:

\begin{lemma}

    $\CFL\subseteq\pNBR1$

\end{lemma}

\begin{proof}

    Let $G$ be a context-free grammar in Chomsky normal form.
    We define $B=(\Sigma,\Gamma,\square,\epsilon,\hb)$, where $\Gamma=\Sigma\cup V\cup\set\square$ and for every rule
    $A\vdash_GXY$, we add the reduction $X,Y\to_BA$.
    And for every rule $A\vdash_G\sigma$, we add the reduction $\sigma,\epsilon\to_BA$.
    Then for every two symbols $X,Y$ we add the reduction $X,Y\to_B\vdash$.
    If $S\vdash_G\epsilon$ is a rule, then we also add the reduction $\square,\epsilon\to_B\epsilon$.
    We also add the rule $\square,S\to_B\epsilon$.

    Now, we claim that for $\alpha,\gamma\in\Gamma^*$, $\alpha\vdash\gamma$ if and only if $\gamma\to\alpha$.
    Indeed, if $\alpha=\alpha_1A\alpha_2$ and we use the rule $A\to XY$, then $\gamma=\alpha_1XY\alpha_2$.
    And we can use the reduction $X,Y\to A$ to get $\gamma\to\alpha_1A\alpha_2=\alpha$ (since every reduction contains $\perp$,
    we can arbitrarily skip strings).
    And if $\alpha=\alpha_1A\alpha_2$ and we use the rule $A\to\sigma$, then $\gamma=\alpha_1\sigma\alpha_2$ and of course we
    have $\gamma\to\alpha$.

    Proving the other direction is similar.
    Thus we have $\alpha\vdash\gamma$ if and only if $\gamma\to\alpha$.
    Thus $\alpha\vdash^*\gamma$ if and only if $\gamma\to^*\alpha$.
    In particular $S\vdash^*\gamma$ (i.e. $\gamma\in\L(G)$) if and only if $\gamma\to^*S$.
    This means that $\square\gamma\to^*\square S\to\epsilon$, so $\gamma\in\L(B)$.
    If $\square$ is reduced before $\gamma$ gets to $S$, we will have the case $\square\gamma\to^*S$, which won't reduce.
    So in either case we have that $S\vdash^*\gamma\iff\square\gamma\to^*S$.
    This works for $\gamma=\epsilon$ as well.
    So we have $\L(G)=\L(B)$ as required.
    \qed

\end{proof}

\subsection{Deterministic $\beta$-Reduction}

\begin{definition}

    An $\nNBR$-schema $B$ is \emph{deterministic} if for every $\sigma\in\Gamma,\tau\in\Gamma\cup\set\epsilon$, we have that
    $\lvert\hb(\sigma,\tau)\rvert=1$.
    That is $\hb(\sigma,\tau)=\set\alpha$ or $\hb(\sigma,\tau)=\set\perp$.
    We call a deterministic $\nNBR$-schema an $\nBR$-schema.
    Equivalently an $\nNBR$-schema is one where $\hb$ is a partial function
    $$ \hb\colon\Gamma\times(\Gamma\cup\set\epsilon) \partial (\Gamma\cup\set\top)^{\leq n} $$
    (Where an image not being defined is equivalent to defining it to be $\set\perp$.)

    The class of languages accepted by $\nBR$-schemas is denoted $\nBR$.
    Let $\sBR=\bigcup_{n>0}\nBR$.

\end{definition}

Our goal in this section is to understand the relationship between $\nBR$ and the class of deterministic context-free languages.
We denote this class by $\DCFL$.
Recall that a deterministic context-free language is one that can be accepted by a deterministic pushdown automaton.
Our previous proof techniques do not work for the deterministic case, as they rely on the ability to non-deterministically
choose whether or not to reduce.

Though notice that our proof that $\sNBR=\pNBR2$ converts deterministic schemas to deterministic schemas.
So we immediately have that

\begin{lemma}

    $\sBR=\pBR2$

\end{lemma}

We will use an equivalent definition of deterministic context-free languages to show that $\DCFL\subseteq\pBR2$.

\begin{definition}

    An \emph{LR parser} is a tuple $P=(\Sigma,Q,V,\action,\goto,q_0,\square)$ where
    \begin{enumerate}
        \item $\Sigma$ is a finite alphabet of terminals;
        \item $Q$ is a finite set of states (disjoint from $\Sigma$);
        \item $V$ is a finite set of variables (disjoint from $\Sigma$ and $Q$);
        \item $\action$ is a partial function
        \begin{multline*}
            \action\colon Q\times\Sigma\cup\set\square\partial\\
            \set{\shift(q)}[q\in Q]\cup\set{\reduce(A,n)}[A\in V,n\in{\bb Z}_{\geq0}]\cup\set\accept
        \end{multline*}
        ($\shift,\reduce,\accept$ are simply distinct symbols.)
        \item $\goto$ is a partial function
        $$ \goto\colon Q\times V \partial Q $$
        \item $q_0\in Q$ is the initial state;
        \item $\square$ is a symbol not in $\Sigma\cup Q\cup V$, the string endmarker.
    \end{enumerate}

\end{definition}

\begin{definition}

    A \emph{configuration} of an LR parser $P$ is a pair $c\in Q^*\times(\Sigma\cup\set\square)^*$, i.e. a pair
    of the form $(p_0\cdots p_n\;;\;\sigma_1\cdots\sigma_m)$.
    We define $P(c)$ as follows: suppose $c=(q_0\cdots q_m\;;\;\sigma_1\cdots\sigma_n\square)$, then
    \begin{enumerate}
        \item if $\action(q_m,\sigma_1)=\shift(p)$ then
        $$ P(c) = (q_0\cdots q_mp\;;\;\sigma_2\cdots\sigma_n\square) $$
        \item if $\action(q_m,\sigma_1)=\reduce(A,k)$ then let $p=\goto(q_{m-k},A)$ and
        $$ P(c) = (q_0\cdots q_{m-k}p\;;\;\sigma_1\cdots\sigma_n\square) $$
        if $\goto(p_{m-k},A)$ is undefined, $P(c)=\bot$ (a special configuration invariant under $P$);
        \item if $\action(q_m,\sigma_1)=\accept$ then $P(c)=\top$ (also invariant under $P$);
        \item if $\action(q_m,\sigma_1)$ is undefined then $P(c)=\bot$.
    \end{enumerate}
    The language of $P$ is
    $$ \L(P) = \set{w\in\Sigma^*}[\hbox{there exists an $n>0$ such that $P^n(q_0\;;\;w\square)=\top$}] $$

\end{definition}

A well-known theorem is the following:

\begin{theorem}

    A language $\L$ is deterministic context-free if and only if there exists an LR parser such that $\L(P)=\L$.

\end{theorem}

\begin{lemma}

    $\DCFL\subseteq\sBR$

\end{lemma}

\begin{proof}

    Let $P$ be an LR parser.
    We define a $\sBR$-schema $(\Sigma,\Gamma,q_0,\top,\hb)$ where $\Gamma$ consists of $\Sigma\cup Q$ and for every
    $\reduce(A,n)$ in the image of $\action$, $M_{k,A}$ for $k\leq n$ and $A\in V$, and we add $\perp,\top$ as well.

    For every $q\in Q$ and $\sigma\in\Sigma$,
    \begin{enumerate}
        \item if $\action(q,\sigma)=\shift(p)$ then $\hb(q,\sigma)=qp$;
        \item if $\action(q,\sigma)=\reduce(A,n)$, then $\hb(q,\sigma)=qM_{n,A}\sigma$;
        \item if $\action(q,\sigma)=\accept$, then $\hb(q,\sigma)=\top$;
        \item if $\action(q,\sigma)$ is undefined, then $\hb(q,\sigma)=\bot$;
        \item $\hb(q,M_{n,A})=M_{n-1,A}$ if $n>0$;
        \item $\hb(q,M_{0,A}=q\goto(q,A)$ if defined, otherwise $\perp$;
        \item $\hb(\sigma,\top)=\hb(\top,\sigma)=\top$.
    \end{enumerate}
    We also add $\hb(q,\epsilon)$ according to $\action(q,\square)$.
    We can assume that this is not a $\shift$ operation.

    It's not hard to see that for a given configuration $c=(v;w)$, we have that $P(c)=(v',w')$ then $vw\to^* v'w'$.
    And if $vw\to^*v'w'$ then $P^k(c)=(v',w')$ for some $k\geq0$.
    Thus we get $\L(P)=\L(B)$ as required.
    \qed

\end{proof}

Since $\sBR=\pBR2$, we get

\begin{lemma}

    $\DCFL\subseteq\pBR2$

\end{lemma}

\end{document}

### 2025-04-28 Paper/paper.tex
Added about beta reducability

@@
 \usepackage{graphicx}
-\usepackage{mdframed}
+%\usepackage{mdframed}
 %\usepackage{cleveref}
@@
 
-\newcommand{\yoni}[1]{{\begin{mdframed}[linecolor=red]{\color{red}#1}\end{mdframed}}}
-\newcommand{\ari}[1]{{\begin{mdframed}[linecolor=blue]{\color{blue}#1}\end{mdframed}}}
+%\newcommand{\yoni}[1]{{\begin{mdframed}[linecolor=red]{\color{red}#1}\end{mdframed}}}
+%\newcommand{\ari}[1]{{\begin{mdframed}[linecolor=blue]{\color{blue}#1}\end{mdframed}}}
+\def\yoni#1{\par\noindent\fbox{\color{red}#1}\par}
+\def\ari#1{\par\noindent\fbox{\color{blue}#1}\par}
 \begin{document}
@@
     \item $s$ is a state, which is a map $\Sigma\to\Gamma$,
-    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon\Gamma\times(\Gamma\cup\set{\epsilon,\blacksquare})\partial\Gamma\cup\set{\epsilon,\bullet}$.
+    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon(\Gamma\cup\set\square)\times(\Gamma\cup\set{\epsilon,\blacksquare})\partial\Gamma\cup\set{\epsilon,\bullet}$.
 \end{enumerate}
@@
 
-According to theorem \ref{thrm:complexity}, all $\beta$-reducible languages can be reduced by their $\beta$-reducer in $O(n)$ time.
+According to theorem \gobble\ref{thrm:complexity}, all $\beta$-reducible languages can be reduced by their $\beta$-reducer in $O(n)$ time.
 
@@
 
-\ari{The following section is just me writing down some ideas, far from finished}
-
-\subsection{Multi-$\beta$-Reducible Languages}
-
-For $n\geq1$, a $\BR(n)$-schema is a tuple $(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\set{\hb_i}_{i=0}^{n-1})$ where $\Sigma,\Gamma,\square,\blacksquare,\bullet,s$ are as before and now
-\[ \hb_i\colon\Gamma\times(\Gamma\cup\set{\epsilon,\square})\partial(\Gamma\cup\set{\epsilon,\bullet})\times\set{0,\dots,n-1} \]
-For each $i\in\set{0,\dots,n-1}$ we define $\beta_i$ as before, but instead of returning the string it also returns the index returned by the reduction.
-$\beta^*$ will then start using $\beta_0$, and then when the current $\beta_i$ returns $j$, it switches to $\beta_j$.
-
-We define \MBR{} to be $\MBR=\bigcup_{n\geq1}\BR(n)$.
-
-This is (we believe) stronger than \BR: $\set{ww^r}[w\in\set{a,b}^+]$ is \MBR.
-We construct a $\BR(5)$-schema as follows:
-\[ B = \parens{\set{a,b},\set{A,B,A',B'},\square,\blacksquare,\bullet,[a\to A,b\to B],\set{\hb_0,\hb_1,\hb_2,\hb_3,\hb_4}} \]
-Where we have the reduction rules:
-\begin{align*}
-    \hb_0\colon\quad& \square,A\to A',1\\
-        & \square,B\to B',3\\
-    \hb_1\colon\quad& A,\blacksquare\to A',2\\
-    \hb_2\colon\quad& A',A'\to\bullet,2\\
-        &A',\epsilon\to\epsilon,2\\
-        &A',\blacksquare\to A',0\\
-    \hb_3\colon\quad& B,\blacksquare\to B',2\\
-    \hb_4\colon\quad& B',B'\to\bullet,2\\
-        &B',\epsilon\to\epsilon,2\\
-        &B',\blacksquare\to B',0\\
-\end{align*}
-Intuitively, $\hb_0$ detects the first character and ``tags'' it, then $\hb_1$ or $\hb_3$ ``tag'' the end character if it is the same as the beginning character.
-$\hb_2$ and $\hb_4$ ``clean up'' the tags: if the two tagged characters are next to each other, then we can try and accept the word.
-Otherwise delete them both.
-
-For the language of palindromes $\set{w\in\set{a,b}^*}[w=w^r]$ we can define a $\BR(3)$-schema where $\bullet=\epsilon$ and
-\begin{align*}
-    \hb_0\colon\quad& \square,A\to\epsilon,1\\
-        & \square,B\to\epsilon,2\\
-    \hb_1\colon\quad& A,\blacksquare\to\epsilon,0\\
-    \hb_2\colon\quad& B,\blacksquare\to\epsilon,0\\
-\end{align*}
-
-But \MBR{} is not contained in \CFL{}: $\set{a^nb^nc^n}[n\geq1]\in\MBR$.
-Indeed, we can construct an \MBR-schema as follows:
-\[ B = \parens{\set{a,b,c},\set{A,B,C,X},[a\to A,b\to B,c\to C],\set{\hb_0,\hb_1,\hb_2,\hb_3}} \]
-And we have the reduction rules:
-\begin{align*}
-    \hb_0\colon\quad& B,A\to X,0\\
-        &C,B\to X,0\\
-        &C,\blacksquare\to C,1\\
-    \hb_1\colon\quad& A,\epsilon\to\epsilon,2\\
-    \hb_2\colon\quad& B,\epsilon\to\epsilon,3\\
-    \hb_3\colon\quad& C,\blacksquare\to\bullet,3\\
-        &C,\epsilon\to\epsilon,1
-\end{align*}
-Intuitively, $\hb_0$ checks that $A$ comes before $B$ which comes before $C$ (as otherwise it places in an $X$, which is never reduced).
-Then $\hb_1$ matches a single $A$, then calls $\hb_2$ to match a single $B$, which calls $\hb_3$ to match a single $C$.
+\subsection{Variants of \BR}
+
+In this section we define variants of $\beta$-reducible languages, which are all special cases (including the initial variant
+already defined) of the following definition:
+
+\begin{definition}
+
+    A \emph{nondeterministic $n$-\BR{} schema} (or an $n$-\NBR{} schema), for $n\in\bN$, is a tuple
+    $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ (like before), where
+    \begin{enumerate}
+        \item $\Sigma$ is a finite alphabet of terminals;
+        \item $\Gamma$ is a finite alphabet of non-terminals (not necessarily disjoint from $\Sigma$);
+        \item $\square,\blacksquare$ are symbols not in $\Sigma,\Gamma$;
+        \item $\bullet$ is a symbol (perhaps in $\Sigma$ or $\Gamma$), or the empty word;
+        \item $s$ is a state, which is a map $\Sigma\to\Gamma$;
+        \item $\hb$ is an initial $\beta$-reducer, which is a function
+        $$ \hb\colon(\Gamma\cup\set\square)\times(\Gamma\cup\set{\epsilon,\blacksquare})\to
+        \P((\Gamma\cup\set{\epsilon,\bullet})^n\cup\set\perp) $$
+    \end{enumerate}
+
+\end{definition}
+
+\begin{definition}
+
+    Given a nondeterministic $n$-\NBR{} schema $B$, we define the function
+    $\beta'\colon(\Gamma\cup\set\bullet)^*\to\P((\Gamma\cup\set\bullet)^*)$ as follows:
+    \begin{enumerate}
+        \item $\beta'(\bullet)=\set\bullet$;
+        \item $\beta'(\sigma)=\hb(\sigma,\blacksquare)$ if defined, otherwise $\set\sigma$;
+        \item if $\xi=\sigma_1\sigma_2\xi'$, then
+        \begin{enumerate}
+            \item if $\perp\notin\hb(\sigma_1,\sigma_2)$, $\beta'(\xi)=\hb(\sigma_1,\sigma_2)\xi'$
+            (where the concatenation of sets is defined word-wise) if defined, and if $\perp$ is in the reduction, then take
+            the union of this with $\sigma_1\
+            otherwise
+            \item $\beta'(\xi)=\hb(\sigma_1,\epsilon)\xi'$ if defined, otherwise
+            \item $\beta'(\xi)=\sigma_1\beta'(\sigma_2\xi')$.
+        \end{enumerate}
+    \end{enumerate}
+    Then we define the function $\beta\colon(\Gamma\cup\set\bullet)^*\to\P((\Gamma\cup\set\bullet)^*)$ as follows:
+    \begin{enumerate}
+        \item $\beta(\epsilon)=\hb(\square,\blacksquare)$ if defined, otherwise $\set\epsilon$;
+        \item $\beta(\sigma\xi)=\hb(\square,\sigma)\xi$ if defined, otherwise:
+        \item $\beta(\xi)=\beta'(\xi)$.
+    \end{enumerate}
+    This naturally extends to a function $\beta\colon\P((\Gamma\cup\set\bullet)^*)\to\P((\Gamma\cup\set\bullet)^*)$.
+    We then define the \emph{language of $B$} to be
+    $$ \L(B) = \set{w\in\Sigma^*}[\bullet\in\beta^k(s(w))\hbox{ for some $k\geq0$}] $$
+
+\end{definition}
+
+The class $n$-\NBR{} is defined to be the class of all languages recognized by a nondeterministic $n$-\BR{} schema.
+Then we define \sNBR{} to be the union of all $n$-\NBR{} for $n\geq1$.
+
+\begin{definition}
+
+    An $n$-\NBR{} schema is \emph{deterministic} if
+
+\end{definition}
 

### 2025-05-18 Paper/beta-reducible-langs.tex
Beta paper update

@@
 
-    Let $n>0$ be an integer, then an \emph{$n$-nondeterministic $\beta$-reduction schema} (shortened to $n-\NBR$ schema) is
+    Let $n>0$ be an integer, then an \emph{$n$-nondeterministic $\beta$-reduction schema} (shortened to $\nNBR$ schema) is
     a tuple $B=(\Sigma,\Gamma,\square,\top,\hb)$ where
@@
             clear mapping of $\Sigma$ into $\Gamma$);
-        \item $\square$ is a symbol in $\Gamma$ (the string beginning-marker);
-        \item $\top$ is a symbol (potentially in $\Gamma$), or the empty word;
+        \item $\square$ is a symbol in $\Gamma$ (the string ends-marker);
+        \item $\top$ is a symbol in $\Gamma$, or the empty word;
         \item $\hb$ is an \emph{initial $\beta$-reducer}, which is a function
-        $$ \hb\colon\Gamma\times(\Gamma\cup\set\epsilon)\to\P((\Gamma\cup\set{\top})^{\leq n}\cup\set\perp) $$
+        $$ \hb\colon\Gamma\times(\Gamma\cup\set\epsilon)\to\P(\Gamma^{\leq n}\cup\set\perp) $$
         Where $\perp$ is a symbol distinct from any other previously discussed symbol.
+        Equivalently, we can view $\hb$ as a relation
+        $$ \hb \subseteq \Gamma\times(\Gamma\cup\set\epsilon)\times\bigl(\Gamma^{\leq n}\cup\set\perp\bigr) $$
     \end{enumerate}
@@
 Let us focus on the initial $\beta$-reducer.
-It takes as input two symbols and outputs a set of strings in $\Gamma\cup\set\top$ of length $\leq n$, and the set may also
+It takes as input two symbols and outputs a set of strings in $\Gamma$ of length $\leq n$, and the set may also
 contain the symbol $\perp$.
@@
     Let $B$ be an $n$-\NBR, then we define its $\beta$-reduction function as follows.
-    The function is $\beta\colon(\Gamma\cup\set\top)^*\to\P_{<\omega}((\Gamma\cup\set\top)^*)$ and defined as follows:
+    The function is $\beta\colon\Gamma^*\to\P_{<\omega}(\Gamma^*)$ and defined as follows:
     \begin{enumerate}
-        \item $\beta(\top)=\set\top$;
         \item $\beta(\sigma)=\hb(\sigma,\epsilon)-\set\perp$ if non-empty, and otherwise $\set\sigma$;
@@
     \end{enumerate}
-    This extends to a function $\beta\colon\P_{<\omega}((\Gamma\cup\set\top)^*)\to\P_{<\omega}((\Gamma\cup\set\top)^*)$
+    This extends to a function $\beta\colon\P_{<\omega}(\Gamma^*)\to\P_{<\omega}(\Gamma^*)$
     in the natural way.
@@
     We then define the language of $B$ (or the language \emph{recognized} by $B$) to be
-    $$ \L(B) = \set{w\in\Sigma^*}[\square w\sto\top] $$
+    $$ \L(B) = \set{w\in\Sigma^*}[\square w\square\sto\top] $$
     We define the class $\nNBR$ to be the class of all languages recognized by $\nNBR$ schemas.
@@
 Note that an $\sNBR$-schema may be defined similar to an $\nNBR$-schema, but where the initial $\beta$-reducer returns finite
-subsets of $(\Gamma\cup\set\top)^*$ (denoted $\P_{<\omega}((\Gamma\cup\set\top)^*)$).
+subsets of $\Gamma^*$ (denoted $\P_{<\omega}(\Gamma^*)$); or equivalently $\hb$ is a finite subset
+$$ \hb \subseteq \Gamma\times(\Gamma\cup\set\epsilon)\times\Gamma^* $$
 This is since $\hb$ will return a longest string of some length $n$, and then the schema is an $\nNBR$ schema.
 
-\subsection{Context-Free Languages}
+Furthermore, we don't gain any power if we define $\L(B)$ as one of the following:
+\begin{align*}
+    \L(B) &= \set{w\in\Sigma^*}[\square w\sto\top],\\
+    \L(B) &= \set{w\in\Sigma^*}[w\square\sto\top],\\
+    \L(B) &= \set{w\in\Sigma^*}[w\sto\top]
+\end{align*}
 
-Recall the definition of context-free languages:
+Notice that $\sNBR=\pNBR2$; this is not difficult to prove.
+Take an $\nNBR$-schema $B=(\Sigma,\Gamma,\square,\top,\hb)$ and define a $\pNBR2$-schema
+$B_2=(\Sigma,\Gamma_2,\square,\top,\hb_2)$ as follows: $\Gamma_2$ will contain $\Gamma$ as well as every prefix of strings
+returned by $\hb$ (each prefix is distinct, if two strings have the same prefix, or the same string is returned by two different
+reductions, we consider them distinct).
+Given such a prefix $\xi$, denote the symbol in $\Gamma_2$ by $\mbox\xi$.
+Then if $\sigma_1\cdots\sigma_n\in\hb(\tau_1,\tau_2)$, we define
+$$ \hb_2(\tau_1,\tau_2) = \mbox{\sigma_1\cdots\sigma_n},\quad
+\hb_2(\mbox{\sigma_1\cdots\sigma_k},\epsilon)=\set{\mbox{\sigma_1\cdots\sigma_{k-1}}\sigma_k}\hbox{ for $k\leq n$} $$
+If $\perp\hb(\tau_1,\tau_2)$, then $\perp\in\hb_2(\tau_1,\tau_2)$ as well.
+
+\begin{lemma}
+
+    $\sNBR=\pNBR2$
+
+\end{lemma}
+
+\subsection{Context-Free Languages and $\pNBR1$}
 
@@
 
-Our goal of this section is to investigate the relationship between the classes $\CFL$ and $\nNBR,\sNBR$.
-
-\begin{lemma}
-
-    $\CFL\subseteq\sNBR$.
-
-\end{lemma}
-
-\begin{proof}
-
-    Let $G=(\Sigma,V,S,P)$ be a context-free grammar.
-    We construct an \sNBR-schema $B=(\Sigma,\Gamma,S,\epsilon,\hb)$, where $\Gamma=V\cup\Sigma\cup\overline\Sigma$ where
-    $\overline\Sigma$ is a disjoint copy of $\Sigma$.
-    For each production rule $A\vdash\alpha\in P$ we place $\perp,\overline\alpha^r\in\hb(A,\epsilon)$ (where $\overline\alpha$
-    is $\alpha$ where $\sigma\in\Sigma$ is replaced with $\overline\sigma\in\overline\Sigma$).
-    And for $\sigma\in\Sigma$, $\hb(\overline\sigma,\sigma)=\set{\epsilon}$.
-
-    Note that $A\to^*\overline\alpha\in(V\cup\overline\Sigma)^*$, and so when reducing $A$, we can choose any variable to
-    reduce.
-    That is, if $A_1\to\alpha_1$ and $A_2\to\alpha_2$, and $A\to^*\gamma_1A_1\gamma_2A_2\gamma_3$, then we have that in a single
-    reduction we can either get $\gamma_1\alpha_1\gamma_2A_2\gamma_3$ or we get  $\gamma_1A_1\gamma_2\alpha_2\gamma_3$.
-    This is because elements of $\overline\Sigma$ won't be reduced, and all other reductions (of variables) may be skipped, as
-    they contain $\perp$.
-
-    Now we claim that for any $n\geq0$ and $A\in V$, if $A\vdash^n\alpha$ iff $A\to^n\overline\alpha^r$.
-    This is clear by induction: for $n=0$ this is because $\alpha=A$.
-    For $n=1$ this is by definition.
-    Now suppose $A\vdash^n\alpha_1B\alpha_3$ and $B\vdash\alpha_2$, so that $A\vdash^{n+1}\alpha_1\alpha_2\alpha_3$.
-    Then by induction $A\to^n\alpha_3^rB\alpha_1^r$ and by definition $B\to\alpha_2^r$, so that we have
-    $A\to^{n+1}\alpha_3^r\alpha_2^r\alpha_1^r=(\alpha_1\alpha_2\alpha_3)^r$ as required.
-    The other direction is proven in a similar way.
-
-    If $w\in\L(G)$, we have that $S\vdash^*w$, and so $Sw\sto\overline w^rw$, which clearly then $\sto\epsilon$.
-    So $w\in\L(B)$.
-
-    Now if $Aw\sto\overline\alpha w_2$, then we claim there exists a $w_1$ such that $w=w_1w_2$ and $A\vdash^*w_1\alpha^r$.
-    We do this by induction on the number of steps in the reduction, $n$.
-    For $n=0,1$ this is clear ($w_1=\epsilon$) in both cases.
-    Now suppose $Aw\to^n\overline\alpha w_2$, then $w=w_1w_2$ and $A\vdash^*w_1\alpha^r$.
-    If the next reduction is in $\overline\alpha$, this must be a variable reduction, which means that $A\vdash^*w_1(\alpha')^r$
-    and $Aw\to^{n+1}\overline\alpha' w_2$, so $w'_1=w_1$ still satisfies the requirement.
-    Otherwise, the reduction is of the form
-    $$ Aw\to^n\overline\alpha'\overline\sigma\sigma w'_2 \to\overline\alpha'w'_2 $$
-    In which case $w=w_1w_2=w_1\sigma w'_2$, so that $w'_1$ must be $w_1\sigma$.
-    Now, notice that $A\vdash^*w_1\alpha^r=w_1(\alpha'\sigma)^r=w_1\sigma(\alpha')^r=w'_1(\alpha')^r$ as required.
-
-    So we have that $Sw\sto\epsilon$, so $w_2=\epsilon$ and thus $w_1=w$, meaning $S\vdash^*w$.
-    Thus if $w\in\L(B)$, $w\in\L(G)$, as required.
-    \qed
-
-\end{proof}
-
-Notice that the parameter of the $\sNBR$-schema is dependent only on the length of the production strings in $G$.
-That is $B$ is an $\nNBR$-schema where $n=\max\set{\abs\alpha}[A\vdash_G\alpha]$.
-Every context-free grammar has an equivalent in Chomsky normal form (CNF), whose $n$ is $2$.
-Thus $\CFL\subseteq\pNBR2$.
-
-In fact, $\sNBR=\pNBR2$; this is not difficult to prove.
-Take an $\nNBR$-schema $B=(\Sigma,\Gamma,\square,\top,\hb)$ and define a $\pNBR2$-schema
-$B_2=(\Sigma,\Gamma_2,\square,\top,\hb_2)$ as follows: $\Gamma_2$ will contain $\Gamma$ as well as every prefix of strings
-returned by $\hb$ (each prefix is distinct, if two strings have the same prefix, or the same string is returned by two different
-reductions, we consider them distinct).
-Given such a prefix $\xi$, denote the symbol in $\Gamma_2$ by $\mbox\xi$.
-Then if $\sigma_1\cdots\sigma_n\in\hb(\tau_1,\tau_2)$, we define
-$$ \hb_2(\tau_1,\tau_2) = \mbox{\sigma_1\cdots\sigma_n},\quad
-\hb_2(\mbox{\sigma_1\cdots\sigma_k},\epsilon)=\set{\mbox{\sigma_1\cdots\sigma_{k-1}}\sigma_k}\hbox{ for $k\leq n$} $$
-If $\perp\hb(\tau_1,\tau_2)$, then $\perp\in\hb_2(\tau_1,\tau_2)$ as well.
-
-\begin{lemma}
-
-    $\sNBR=\pNBR2$
-
-\end{lemma}
-
-Though, the most important result is as follows:
-
-\begin{lemma}
+\begin{theorem}
 
@@
 
-\end{lemma}
+\end{theorem}
 
@@
 
+\subsection{Rewrite Systems}
+
+We first begin by taking a different perspective on $\beta$-reduction.
+
+\begin{definition}
+
+    An \emph{$n$-$\beta$-rewrite system} ($\BRW n$-system) is a tuple $B=(\Sigma,V,\hb)$, where
+    \begin{enumerate}
+        \item $\Sigma$ is a finite alphabet;
+        \item $V$ is a finite set of variables, disjoint from $\Sigma$;
+        \item $\hb$ is the \emph{initial $\beta$-function} a function
+        $$ \hb\colon \Gamma\times(\Gamma\cup\set\epsilon)\to\P(\Gamma^{\leq n}) $$
+        Where $\Gamma=\Sigma\cup V$.
+        Equivalently, a $\hb$ is a relation
+        $$ \hb \subseteq \Gamma\times(\Gamma\cup\set\epsilon)\times\Gamma^{\leq n} $$
+    \end{enumerate}
+    We can also call an $\BRW n$-system a $\sBRW$-system.
+
+\end{definition}
+
+\begin{definition}
+
+    Let $B$ be a $\BRW n$-system, then we define the binary relation $\to_B$ on $\Gamma^*$ as follows: we write
+    $\alpha\to_B\gamma$ if $\alpha=\alpha_1\sigma\tau\alpha_2$, where $\sigma\in\Gamma$ and $\tau\in\Gamma\cup\set\epsilon$ and
+    $\gamma=\alpha_1\hb(\sigma\tau)\alpha_2$.
+    We then define $\to_B^*$ to be the transitive and reflexive closure of $\to$.
+
+    Further, we define $\tto_B$ and $\tto^*_B$ to be restrictions of $\to_B$ and $\to^*_B$ to $\Sigma^*$.
+
+    Call two $\sBRW$s $B_1$ and $B_2$ \emph{equivalent} if ${\tto_{B_1}^*}={\tto_{B_2}^*}$
+
+\end{definition}
+
+Notice that unlike with $\sNBR$-schemas, in $\sBRW$-systems all reductions may be done non-deterministically.
+E.g. given a string $abc$, we can reduce either $a,b$ or $b,c$.
+As we will see, this doesn't impose any practical limitations for $n\geq2$.
+
+Using a proof similar to the one for $\sNBR$-schemas, we get the following result:
+
+\begin{lemma}
+
+    Every $\sBRW$-system is equivalent to a $\BRW2$-system.
+
+\end{lemma}
+
+\begin{definition}
+
+    A \emph{semi-Thue system} (or a \emph{string rewriting system}) is a pair $T=(\Sigma,R)$ where $\Sigma$ is an alphabet
+    (not necessarily finite), and $R$ is a binary relation on $\Sigma^*$.
+    Instead of $(u,v)\in R$ we write $u\vdash_Tv$.
+    We extend this relation to other $\Sigma$-strings as follows: if $u\vdash_Tv$ then for any $\alpha,\gamma\in\Sigma^*$ we
+    declare $\alpha u\gamma\vdash_T\alpha v\gamma$.
+    We let $\vdash_T^*$ be the reflexive and transitive closure of $\vdash_T$.
+
+    If both $\Sigma$ and $R$ are finite, we call $T$ \emph{finite}.
+
+\end{definition}
+
+Finite semi-Thue systems are Turing complete.
+This is clear since every formal grammar forms a finite semi-Thue system.
+
+\begin{lemma}
+
+    $\sBRW$ (and thus $\BRW2$) is Turing complete.
+
+\end{lemma}
+
+\begin{proof}
+
+    It is sufficient to show that for every finite semi-Thue system $T=(\Sigma,R)$, there exists a $\sBRW$-system $B$ such that
+    ${\tto_B^*}={\vdash_T^*}$.
+
+    Let $B=(\Sigma,V,\hb)$, where for every string $u$ in $R$ we add all the prefixes of $u$ to $V$.
+    We can assume that for each $u$ of $R$ the substrings added are unique (and we view multiple occurrences of $u$ in $R$ as
+    distinct).
+    We write elements of $V$ in boxes to distinguish them.
+
+    Now for $u\vdash_Tv$, let us suppose that $u\neq\epsilon$.
+    So let $u=\sigma_1\cdots\sigma_n$ and $v=\tau_1\cdots\tau_m$, then we add rules
+    $$
+    \sigma_1,\epsilon\to_B\mbox{\sigma_1},\qquad
+    \mbox{\sigma_1\cdots\sigma_k},\sigma_{k+1}\to_B\mbox{\sigma_1\cdots\sigma_{k+1}}
+    $$
+    and
+    $$
+    \mbox{\sigma_1\cdots\sigma_n},\epsilon\to_B\tau_1\cdots\tau_m
+    $$
+
+    For rules of the form $\epsilon\vdash_Tv$, we add rules $\sigma,\epsilon\to\sigma v$ for every $\sigma\in\Sigma$.
+
+    Clearly if $u\vdash_Tv$ then $u\tto_B^*v$, and vice versa (for the converse we of course have $u\vdash_T^*v$).
+    Thus we have the desired result.
+    \qed
+
+\end{proof}
+
+\begin{theorem}
+
+    Every recursively enumerable language is $\pNBR2$, and conversely every $\sNBR$ language is recursively enumerable.
+
+\end{theorem}
+
+If $T$ is a semi-Thue system, it has an \emph{inverse}: a semi-Thue system $T^r$ where
+$$ u\vdash_{T^r}v \iff v\vdash_Tu $$
+
+\begin{proof}
+
+    Given a recursively enumerable language $L$, there exists some finite semi-Thue system which generates it (meaning
+    $u\in L$ iff $S\vdash_T^*u$ for some $S\in\Sigma$).
+    Then if we look at the inverse semi-Thue system, $u\in L$ iff $u\vdash_{T^r}^*S$.
+    Let $B$ be the $\BRW2$-system equivalent to $T^r$.
+    Then we make it into a $\pNBR2$-schema by setting $\Gamma=\Sigma\cup V$, and $\top=S$.
+    Then we have that
+    $$ u\in L \iff u\vdash_{T^r}^*S \iff u\tto_B^*\top \iff u\in\L(B) $$
+    as required.
+
+    The converse is obvious: every $\sNBR$-schema can obviously be simulated by a nondeterministic Turing machine.
+    \qed
+
+\end{proof}
+
+So we see that for $n\geq2$, whether or not $\BRW2$-systems can always or can only sometimes skip a reduction is immaterial:
+both are Turing complete.
+This justifies us not adding $\bot$ to $\sBRW$-systems.
+
 \subsection{Deterministic $\beta$-Reduction}
@@
     That is $\hb(\sigma,\tau)=\set\alpha$ or $\hb(\sigma,\tau)=\set\perp$.
+    Furthermore, we require that if $\hb(\sigma,\tau)$ is defined (not equal to $\set\perp$) for some $\tau$,
+    $\hb(\sigma,\epsilon)$ is undefined.
+
     We call a deterministic $\nNBR$-schema an $\nBR$-schema.
     Equivalently an $\nNBR$-schema is one where $\hb$ is a partial function
-    $$ \hb\colon\Gamma\times(\Gamma\cup\set\epsilon) \partial (\Gamma\cup\set\top)^{\leq n} $$
+    $$ \hb\colon\Gamma\times(\Gamma\cup\set\epsilon) \partial \Gamma^{\leq n} $$
     (Where an image not being defined is equivalent to defining it to be $\set\perp$.)
+    such that if $\hb(\sigma,\tau)$ is defined, $\hb(\sigma,\epsilon)$ is not.
 
@@
         \item $\hb(q,M_{n,A})=M_{n-1,A}$ if $n>0$;
-        \item $\hb(q,M_{0,A}=q\goto(q,A)$ if defined, otherwise $\perp$;
+        \item $\hb(q,M_{0,A})=q\goto(q,A)$ if defined, otherwise $\perp$;
         \item $\hb(\sigma,\top)=\hb(\top,\sigma)=\top$.
     \end{enumerate}
-    We also add $\hb(q,\epsilon)$ according to $\action(q,\square)$.
+    We also add $\hb(q,\square)$ according to $\action(q,\square)$.
     We can assume that this is not a $\shift$ operation.
@@
 
+Notice here that the $\sBR$-schema defined in the lemma above is a $\pBR3$-schema.
+But under the assumption that there are no actions of the form $\reduce(A,0)$ (which corresponds to a rule
+$A\to\epsilon$ in a grammar), we can alter it to be a $\pBR2$-schema.
+We replace rule $2$ with $\hb(q,\sigma)=M_{n-1,A}\sigma$.
+Now, we define the following variation of $\pBR2$-schemas:
+
+\begin{definition}
+
+    A $\pBR2$-schema $B$ is \emph{non-increasing} if for all $\sigma\in\Gamma$,
+    $\abs{\hb(\sigma,\epsilon)}\leq1$ if defined.
+    That is, the string being reduced can never increase in length.
+
+\end{definition}
+
+\begin{definition}
+
+    Let $B$ be a non-increasing $\pBR2$-schema.
+    It is \emph{decreasing} if the following two conditions hold:
+    \begin{enumerate}
+        \item for all $\sigma\in\Gamma$, $\beta^n(\sigma)$ never equals $\sigma$ for all $n\geq1$;
+        \item for all $\sigma,\tau\in\Gamma$, $\beta^n(\sigma\tau)$ never begins with $\sigma$.
+    \end{enumerate}
+
+\end{definition}
+
+It is immediate from this definition that a decreasing $\pBR2$-schema will always terminate on every input
+string.
+
+\begin{lemma}
+
+    Every non-increasing $\pBR2$-schema has an equivalent decreasing $\pBR2$-schema.
+
+\end{lemma}
+
+\begin{proof}
+
+    Let $B$ be a non-increasing $\pBR2$-schema, we construct an equivalent decreasing $\pBR2$-schema $B'$.
+    Our first step we do iteratively, starting with a $\pBR2$-schema $B_1$, until $B_{\abs\Gamma}$.
+    First we add a new symbol $\perp$ to $\Gamma_1$, it has no reduction rules.
+
+    Suppose we have constructed $B_i$ ($0\leq i<\abs\Gamma$, $B_0=B$).
+    For each $\sigma\in\Gamma$ define
+    $$ n_\sigma = \min\set{n\geq1}[\beta^n(\sigma)=\sigma] $$
+    If all $n_\sigma$s are equal to $\infty$, then set $B_{i+1}=B_i$.
+    Otherwise let $\sigma$ has the minimal $n_\sigma$, which means that in the reduction
+    $$ \sigma = \sigma_0 \to \sigma_1 \to \cdots \to \sigma_n = \sigma $$
+    all the $\sigma_i$s are distinct, except for $i=0,n$.
+
+    Now, add new symbols $\sigma'_1,\dots,\sigma'_{n-1}$ to $\Gamma_{i+1}$, with the reduction rules
+    \begin{align*}
+            \hb_{i+1}(\sigma'_j,\epsilon)&=\sigma'_{j+1},\\
+            \hb_{i+1}(\tau,\sigma'_j) &= \hb(\tau,\sigma_j),\\
+            \hb_{i+1}(\sigma'_j,\tau) &= \hb(\sigma_j,\tau)
+    \end{align*}
+    for $0\leq j\leq n-2$ ($\sigma'_0=\sigma$), as well as the rule $\hb_{i+1}(\sigma'_{n-1},\epsilon)=\bot$.
+    We have thus created $B_{i+1}$.
+
+    Now, $B^0=B_{\abs\Gamma}$ has the first property of a non-increasing schema.
+    That is, if $\beta^n(\sigma)$ never equals $\sigma$.
+    Furthermore, $B^0$ and $B_0=B$ are equivalent.
+
+    We now construct $B'$ from $B^0$.
+    For every reduction of the form
+    $$ \sigma_0\tau_0 \to \sigma_1\tau_1 \to \cdots \to \sigma_n\tau_n = \sigma_0\tau_n $$
+    we add characters $\sigma'_1,\dots,\sigma'_{n-1}$.
+    We add the reduction rules
+    \begin{align*}
+            \hb'(\sigma'_i,\tau_i) &= \sigma'_{i+1}\tau_{i+1},\\
+            \hb'(\tau,\sigma'_i) &= \hb(\tau,\sigma_i),\\
+            \hb'(\sigma'_{n-1},\tau_{n-1}) &= \bot
+    \end{align*}
+    This is clearly decreasing, and equivalent to $B^0$.
+    \qed
+
+\end{proof}
+
+\begin{lemma}
+
+    Every decreasing $\pBR2$ schema takes $O(\abs\Gamma\cdot n^2)$ reductions until it terminates.
+
+\end{lemma}
+
+\begin{proof}
+
+    We claim that it  must take at most $\abs\Gamma\cdot n$ reductions to reduce the string by one symbol.
+
+    Let us call reductions of the form $\sigma_1,\sigma_2\to\tau_1\tau_2$ $2$-reductions and $\sigma\to\tau$
+    $1$-reductions.
+
+    Let us create a counter $n(i)$ for the number of $1$- and $2$- reductions at each index.
+    Successive $1$- or $2$-reductions of an index can never place a symbol at that index that has already been
+    there, since the schema is decreasing.
+    This means that $n(i)\leq\abs\Gamma$.
+    And the number of reductions in total is just $\sum_in(i)\leq n\abs\Gamma$, as required.
+
+    Then it takes at most $n$ reductions of a symbol to terminate.
+    Which means that it takes at most $O(\abs\Gamma\cdot n^2)$ reductions.
+    \qed
+
+\end{proof}
+
+\begin{lemma}
+
+    There exists an algorithm which executes any given decreasing $\pBR2$-schema in $O(n^2)$ time.
+    This algorithm is also optimal.
+
+\end{lemma}
+
+\begin{proof}
+
+    Since an execution of a decreasing $\pBR2$-schema requires $O(n^2)$ reductions, any algorithm which
+    executes a $\pBR2$-schema takes $\Omega(n^2)$ time.
+    Thus if we have an algorithm which runs in $O(n^2)$ time, it must be optimal.
+
+    Our algorithm is as follows:
+
+    \def\flag{\mathit{flag}}
+    \algorithm
+        \Function{Execute}{$B,w$}
+            \State $\flag\gets1$
+            \State $i\gets0$
+            \While{$\flag$}
+                \lIf{$\hb(w[i],w[i+1])=\sigma_1\sigma_2$}
+                    Replace $w[i],w[i+1]$ with $\sigma_1,\sigma_2$
+                \lElseIf{$\hb(w[i],w[i+1])=\sigma$}
+                    Replace $w[i],w[i+1]$ with $\sigma$
+                \lElseIf{$\hb(w[i],\epsilon)=\sigma$}
+                    Replace $w[i]$ with $\sigma$
+                \If{any of the above conditions were true, and $\sigma\neq\epsilon$}
+                    \lIf{$\sigma\neq\epsilon$} $i\gets i-1$
+                \Else
+                    \State $i\gets i+1$
+                    \lIf{$i=\abs w$} $\flag\gets0$
+                \EndIf
+            \EndWhile
+            \State\Return $w$
+        \EndFunc
+    \ealgorithm
+
+\end{proof}
+
+Unfortunately, notice that this is not enough to show that $\DCFL\subseteq\pBR1$.
+This is since we conjecture that $\pBR1\subseteq\DCFL$ yet decreasing $\pBR2$-schemas can still recognize non-$\DCFL$ languages.
+In fact, they can recognize non-$\CFL$ languages!
+For example, $\set{a^nb^nc^n}[n\geq0]$ is not $\CFL$ yet is $\pBR2$.
+
+\begin{proposition}
+
+    $\set{a^nb^nc^n}[n\geq0]$ is $\pBR2$.
+
+\end{proposition}
+
+\begin{proof}
+
+    Define $\Gamma=\set{a,b,c,X,\square}$, and $\hb$ as follows:
+    \begin{enumerate}
+        \item $\hb(a,b)=X$: this matches the first $a$ and $b$;
+        \item $\hb(a,X)=Xa$; this moves the $X$ out of the way (eventually the the leftmost side of the string) to allow
+            rule $1$ to repeat;
+        \item $\hb(X,c)=\epsilon$: this matches $X$ with a $c$;
+        \item $\hb(\square,\square)=\top$.
+    \qed
+    \end{enumerate}
+
+\end{proof}
+
+Using a similar proof we can show that $\set{\sigma_1^n\cdots\sigma_m^n}[n\geq0]$ is $\pBR2$ for
+$\Gamma=\set{\sigma_1,\cdots,\sigma_m}$.
+
+\subsubsection{Priorities}
+
+Now, one may be inclined to ask why we don't add priorities to this $\beta$-reduction.
+It turns out that priorities, under certain assumptions, don't add any power.
+Indeed, suppose we added priorities to our $\beta$-reducer, so now our initial $\beta$-reducer is of the form
+$$ \hb\colon\Gamma\times(\Gamma\cup\set\epsilon)\to\P_{<\omega}((\Gamma^{\leq n}\times(\bbZ^2\to\bbZ^n))\cup\set\bot) $$
+and we have an initial priority function $\pi\colon\Sigma\to\bbZ$.
+Let $F$ be the set of priority functions returned by $\hb$, that is
+$$ F = \set{f\colon\bbZ\times\bbZ\to\bbZ^n}[(\xi,f)\in\hb(\sigma,\tau)\hbox{ for some $\sigma,\tau$}] $$
+and let $N=\mathrm{Im}\pi$ be the set of all initial priorities.
+Now let us define $N_0=N$ and $N_{n+1} = N_n\cup\set{k}[n\geq m\in N_n, f\in F,k\in f(n,m)]$.
+We define $N_\infty=\bigcup_{n\geq0}N_n$.
+If $N_\infty$ is finite, then there exists an equivalent \BR-schema (one without priorities).
+
+The proof of this is simple: define $B=(\Sigma,\Gamma',\square,\blacksquare,\bullet,s',\hb')$ where
+\begin{enumerate}
+    \item $\Gamma'=\Gamma\times N_\infty$ which is still finite.
+        Elements of $\Gamma'$ will be written as $\sigma_n$ instead of $(\sigma,n)$.
+    \item $s'$ maps $\sigma$ to $s(\sigma)_{\pi(\sigma)}$.
+    \item If $\hb(\sigma,\tau)=(\alpha,f)$ then for $n\geq m\in N_\infty$ we define $\hb'(\sigma_n,\tau_m)=\xi_{f(n,m)}$.
+        Where by $\xi_{f(n,m)}$ we mean that if $f(n,m)=(n_1,\dots,n_k)$ and $\xi=\sigma^1\cdots\sigma^k$ then
+        $\xi_{f(n,m)}=\sigma^1_{n_1}\cdots\sigma^k_{n_k}$.
+        Note that by definition $n_i\in N_\infty$ so this is well-defined.
+\end{enumerate}
+It is not hard to see that the $\beta$-reducer and this \BR-schema both generate the same language.
+
+The requirement that $N_\infty$ be finite is a strong condition (for example, $F$ cannot include a function of the form $(n,m)\mapsto n+m$ if $N\neq\set0$).
+But in practice this is not a restrictive condition, for example ILang satisfies this condition: if $F$ consists of functions of the form $k_c$ (the constant map), $\snd,\fst$, then $F_\infty$ will trivially be finite.
+
 \end{document}

### 2025-07-13 Paper/string-reduction-schemas.tex
SRSs

\documentclass{llncs}

%\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{enumitem}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{mdframed}
%\usepackage{cleveref}

\def\labelitemi{$\bullet$}
\def\labelenumi{(\textbf{\arabic{enumi}})}

\usepackage{tikz}
\usetikzlibrary{trees}

\newcounter{algo}

\floatstyle{ruled}
\newfloat{algo}{h}{aux}
\floatname{algo}{Algorithm}

\input preamble

\newcommand{\yoni}[1]{{\begin{mdframed}[linecolor=red]{\color{red}#1}\end{mdframed}}}
\newcommand{\ari}[1]{{\begin{mdframed}[linecolor=blue]{\color{blue}#1}\end{mdframed}}}
%\def\yoni#1{\par\noindent\fbox{\color{red}#1}\par}
%\def\ari#1{\par\noindent\fbox{\color{blue}#1}\par}
\begin{document}

\title{String Reduction Schemas}
\author{Ari Feiglin\inst{1}\and Yoni Zohar\inst{1}}
\institute{Bar-Ilan University, Ramat Gan, Israel}

\maketitle

\begin{abstract}

    In this paper we introduce the concept of \emph{string reduction schemas} in varying levels of strength, creating a new hierarchy of classes of formal languages.
    We compare this hierarchy with the well-known Chomsky Hierarchy, and show various inclusions and equalities.

    The main result of this paper is to show that $\DCFL$, the class of deterministic context-free languages, is contained in $\SR1$.
    This is significant as $\SR1$ schemas can be naturally produced by humans and parsers which work with them are simple to implement.
    Thus string reduction schema are useful in the theory of parsing.
    We further demonstrate this by utilizing $\SR1$ schema to create parsers in OCaml as well as in \TeX, demonstrating the ease of implementation.

    \keywords{String Reduction Schema\and Context-Free Language\and Parsing}

\end{abstract}

\section{Introduction}

We begin our discussion by briefly giving an overview of parsing.

\ari{TODO: write introduction}

\section{A note on notation}

While we don't use much special notation in this paper, it nevertheless is worthwhile to discuss some potentially non-standard notation we do use.

First we have some language-theoretic notational definitions:
\begin{itemize}
    \item Given two words $u,v$, we denote their concatenation by juxtaposition (i.e. $uv$).
    \item Given a set of symbols $\Sigma$, we denote $\Sigma^*$ the free monoid over $\Sigma$ (i.e. the set of finite-length words over $\Sigma$).
    \item $\epsilon$ denotes the empty word, which is the unit in the $\Sigma^*$ (i.e. $u\epsilon=\epsilon u=u$ for every $u\in\Sigma^*$).
    \item Given a set of symbols $X$ and an integer $n\geq0$, we denote by $X^n$ the set of $n$-length words in $X$.
    For $n=0$, $X^0=\set\epsilon$.
    \item Similarly, $X^{\leq n}$ denotes the set of words over $X$ of length $\leq n$.
    Equivalently, $X^{\leq n}=\bigcup_{m\leq n}X^m$.
    \item Similarly, $X^{(0,n]}=\bigcup_{0<m\leq n}X^m$.
\end{itemize}

We also have some set-theoretic notational definitions:
\begin{itemize}
    \item We denote by $A\partial B$ the type of partial functions from $A$ to $B$ (i.e. functions which need not produce an output for every input).
    \item $\P(A)$ is the powerset of $A$: the set of all subsets of $A$.
    \item $\P_{<\omega}(A)$ is the finitary powerset of $A$: the set of all finite subsets of $A$.
    \item $\P_{>0}(A)$ is the non-empty powerset of $A$: the set of all non-empty substets of $A$.
    \item Let $X$ be a set and $x$ be some symbol (which in context cannot be mistaken for a set), then we write $X\cup x$ for $X\cup\set x$.
\end{itemize}

\section{Nondeterministic string reduction schemas}

We begin our introduction of string reduction schema by defining their most general variation.
As discussed, string reduction schema deviate from the norm of using grammars as input and instead use string rewriting systems.
This has the benefit of being straightforward to implement programmatically.

\begin{definition}

    A \emph{$\pNSR{n,m}$ schema} (short for $(n,m)$-nondeterministic string rewriting schema) is a tuple
    $$ B = (\Sigma,\Gamma,\square,\A,\ib) $$
    where
    \begin{enumerate}
        \item $\Sigma$ is the alphabet of \emph{input terminals};
        \item $\Gamma$ is the alphabet of \emph{intermediary variables} (or just \emph{variables}).
        It must contain $\Sigma$ (although in general we find it more useful to define an embedding of $\Sigma$ into $\Gamma$);
        \item $\square\in\Gamma$ is the \emph{string endmarker symbol};
        \item $\A\subseteq\Gamma$ is the set of \emph{acceptance symbols};
        \item $\ib$ is the \emph{initial reduction function}, a function:
        $$ \ib\colon\Gamma^{(0,n]}\longto\P_{>0}(\Gamma^{\leq m}\cup\perp) $$
        That is, $\ib$ accepts strings over $\Gamma$ of length $\leq n$ and returns a set of strings of length $\leq m$, as well as potentially the symbol $\perp$ (which is a symbol not in any other set of symbols we are to
        consider).
    \end{enumerate}

\end{definition}

Of course, if we neglect $\perp$, then $\ib$ can just be viewed as a string rewrite rule $\to$ where the lhs of $\to$ must be strings of length $\leq n$ and the rhs must be strings of length $\leq m$.
The idea of this a $\pNSR{n,m}$ schema is to define a method to rewrite strings.
This is done by iterating over an input string from left to right and rewriting any substring non-deterministically which matches an input of $\ib$.
$\perp$ is a symbol telling us we can skip over this substring (nondeterministically).

For example, if $\Gamma=\set{a,b}$ and $\ib(ab)=\set{a,\perp}$, $\ib(aa)=\set{b}$, and $\ib(ba)=\set{a}$ (assume $\ib(w)=\set\perp$ for all other words $w$).
Then we can rewrite (we underline the substrings being rewritten)
$$ a\underline{ba}a \to \underline{aa}a \to \underline{ba} \to a $$
Notice that in the first rewrite we skip over $\underline{ab}aa$, as we can since $\perp\in\ib(ab)$.

Let us now formalize this idea.

\begin{definition}

    Let $B=(\Sigma,\Gamma,\square,\A,\ib)$ be an $\pNSR{n,m}$ schema.
    We define its \emph{reduction function} to be the function
    $$ \beta\colon \Gamma^* \longto \P_{<\omega}(\Gamma^*) $$
    defined in the following way.
    Let $S$ be the set of all triplets of words $t_i,u_i,v_i$ where
    \begin{enumerate}
        \item $w=t_iu_iv_i$;
        \item $t_i$ has no subwords $t'$ such that $\perp\notin\ib(t')$;
        \item $\ib(u_i)\neq\set\perp$;
    \end{enumerate}
    then we define
    $$ \beta(w) = \set{t_iu'v_i}[t_iu_iv_i\in S,\,u'\in\ib(u_i)-\perp] $$

    Now it is indeed simpler to view this as a string rewrite relation.
    So let us define $\to_B$ to be the binary relation on $\Gamma^*$ defined by $u\to_Bv$ if and only if $v\in\beta(u)$.
    Let $\to_B^*$ be its reflexive and transitive closure.
    We define the language of $B$ (or the language \emph{accepted} by $B$) to be
    $$ \L(B) = \set{w\in\Sigma^*}[\square w\square\to^*\top\in\A] $$

    The class of all languages accepted by an $\pNSR{n,m}$-schema is denoted $\pNSR{n,m}$.

\end{definition}

\ari{TODO: prove $\pNSR{n,m}=\pNSR{2,2}=\RE$ and $\CFL\subseteq\pNSR{2,1}$}

\section{Deterministic string reduction schemas}

\begin{definition}

    A $\pNSR{n,m}$ schema $B$ is \emph{deterministic} if $\abs{\ib(w)}=1$ for every $w\in\Gamma^{\leq n}$, and if $\ib(w)$ is defined, $\ib(w')$ is undefined for all subwords $w'$ of $w$.
    We call such a schema a $\pSR{n,m}$-schema.

\end{definition}

Notice that if $B$ is a $\pNSR{n,m}$ schema, then $\ib(w)$ is either a singleton of the form $\set u$ or $\set\perp$.
Thus we can consider $\ib$ to be a partial function
$$ \ib\colon\Gamma^{\leq n}\partial\Gamma^{\leq m} $$
where $\ib(w)$ is undefined if (for the previous $\ib$), $\ib(w)=\set\perp$.
Then our definition of $\beta$ stays similar, except now $S$ is the set of triplets $t_i,u_i,v_i$ where
\begin{enumerate}
    \item $w=t_iu_iv_i$;
    \item $t_i$ has no subwords $t'$ such that $\ib(t')$ is defined;
    \item $\ib(u_i)$ is defined.
\end{enumerate}

Now, why can't a subword of a word in the domain of $\ib$ have its image defined?
Suppose $w=uw'v$ and $\ib(w),\ib(w')$ both be defined.
Assume that $\ib(u)$ is undefined, then $\beta(w)$ may have a size greater than $1$.
This is because we can reduce both $w$ as a whole as well as $w'$.
That is, $\ib(w),u\ib(w')v\in\beta(w)$.

This cannot happen in our definition of deterministic schema:

\begin{lemma}

    Let $B$ be a $\pSR{n,m}$ schema.
    Then for every $w\in\Gamma^*$, $\beta(w)$ has at most one word.

\end{lemma}

\ari{TODO: formally prove this}

\ari{TODO: show $\pSR{n,m}=\pSR{2,2}$ and describe relation between $\pSR{2,2}$ and $\CFL$}

%\section{$\SR1$ and $\SRe1$ schemas}
\section{$\SR1$ schemas}

Our main focus of this section is $\SR1$ schemas.
%Now, recall that for a deterministic schema, subwords of defined words cannot be defined (where a word being defined means it has an image under $\ib$).
%In the context of $\SR1$ schemas, this is quite restricting, as it means that if $\ib(ab)$ is defined, $\ib(a)$ cannot be.
%For this reason we define $\SRe1$ schemas which allow for $\ib(ab)$ and $\ib(a)$ to be defined.
%Intuitively, $\ib(a)$ is only ``used'' if $\ib(ab)$ cannot be.

%\begin{definition}
%
%    An $\SRe1$ schema is a $\pNSR{2,1}$ schema such that $\ib(w)$ has a size of one for every $w\in\Gamma^{(0,2]}$.
%
%\end{definition}
%
%Given a $\SRe1$ schema $B$, we define $\beta(w)$ similarly as before except $S$ is now defined as triplets $t_i,u_i,v_i$ where
%\begin{enumerate}
%    \item $w=t_iu_iv_i$;
%    \item $t_i$ has no $\ib$-defined subwords;
%    \item $\ib(u_i)$ is defined;
%    \item if $u_i=a\in\Gamma$ and $v_i=bv'_i$, then $ab$ is not $\ib$-defined.
%\end{enumerate}

The goal of this section is to show that every deterministic context-free language is $\SR1$.
To do so, we first define what we mean by a deterministic context-free language.

\subsection{Deterministic context-free languages}

We borrow the following definition from Sipser.

\begin{definition}

    A \emph{deterministic pushdown automaton} (\emph{DPDA}) is a tuple $P=(\Sigma,\Gamma,Q,q_0,F,\delta)$
    where
    \begin{enumerate}
        \item $\Sigma$ is the alphabet of \emph{terminals};
        \item $\Gamma$ is the alphabet of \emph{stack symbols};
        \item $Q$ is the set of \emph{states};
        \item $q_0\in Q$ is the \emph{initial state};
        \item $F\subseteq Q$ is the set of \emph{accepting states};
        \item $\delta$ is the \emph{transition function}:
        $$ \delta\colon Q\times\Sigma_\epsilon\times\Gamma_\epsilon\partial Q\times\Gamma_\epsilon $$
        such that for every $q\in Q,a\in\Sigma,A\in\Gamma$, exactly one of the following is defined:
        $$ \delta(q,a,A),\ \delta(q,a,\epsilon),\ \delta(q,\epsilon,A),\ \delta(q,\epsilon,\epsilon) $$
    \end{enumerate}

\end{definition}

\begin{definition}

    Let $P$ be a DPDA, then we define a \emph{configuration} of $P$ to be an element of
    $Q\times\Sigma^*\times\Gamma^*$.
    We define the relation $\vdash_P$ on the set of configurations like so.
    Given a configuration $(q,aw,AW)$ (where $a\in\Sigma_\epsilon$, $A\in\Gamma_\epsilon$), if
    $\delta(q,a,A)=(p,B)$, then
    $$ (q,aw,AW) \vdash_P (p,w,BW) $$

    Let $\vdash^*_P$ be the transitive and reflexive closure of $\vdash_P$.
    Then the language of $P$ (or the language \emph{recognized} by $P$) is
    $$ \L(P) = \set{w\in\Sigma^*}[(q_0,w,\epsilon)\vdash^*_P(f,\epsilon,W)\hbox{ for some $f\in F$ and
    $W\in\Gamma^*$}] $$

    The class of languages recognized by a DPDA is denoted $\DCFL$.

\end{definition}

We take $P$ to be a DPDA, and we hold it constant for the remainder of this section.
Our goal is to show that there exists a $\SR1$ schema $B$ such that $\L(B)=\L(P)$.

Notice that transitions in a DPDA are one of the following:
\begin{enumerate}
    \item $\delta(q,a,A)=(p,B)$
    \item $\delta(q,a,A)=(p,\epsilon)$
    \item $\delta(q,a,\epsilon)=(p,A)$
    \item $\delta(q,a,\epsilon)=(p,\epsilon)$
    \item $\delta(q,\epsilon,A)=(p,B)$
    \item $\delta(q,\epsilon,A)=(p,\epsilon)$
    \item $\delta(q,\epsilon,\epsilon)=(p,A)$
    \item $\delta(q,\epsilon,\epsilon)=(p,\epsilon)$
\end{enumerate}
We can cut some of these out.
That is,
\begin{enumerate}
    \item $\delta(q,a,A)=(p,B)$ is unnecessary.
    We can add another state $p_B$ and replace this with two transitions: $\delta(q,a,A)=(p_B,\epsilon)$ and $\delta(p_B,\epsilon,\epsilon)=(p,B)$.
    \setcounter{enumi}{2}
    \item $\delta(q,a,\epsilon)=(p,A)$ is unnecessary.
    We can add another state $p_A$ and replace this with two transitions: $\delta(q,a,\epsilon)=(p_A,\epsilon)$ and $\delta(p_A,\epsilon,\epsilon)=(p,A)$.
    \setcounter{enumi}{4}
    \item $\delta(q,\epsilon,A)=(p,B)$ is unnecessary for the same reason as $(1)$.
    \setcounter{enumi}{7}
    \item $\delta(q,\epsilon,\epsilon)=(p,\epsilon)$ is unnecessary, as it is the only transition between $q$ and $p$, so we can simply unify $q$ and $p$.
\end{enumerate}
So we are left with four types of transitions:
\begin{enumerate}
        \setcounter{enumi}{1}
    \item a read \& pop transition (read a character from the stream, and accordingly pop a character from the stack and change states);
        \setcounter{enumi}{3}
    \item a read transition (read a character from the stream, and accordingly change states);
        \setcounter{enumi}{5}
    \item a pop transition (pop a character from the stack without reading from the stream, and change states accordingly);
    \item a push transition (push a character onto the stack without reading from the stream, and change states accordingly).
\end{enumerate}

Notice the following interesting, yet simple, observation.
Define $\widehat P$ to be the same DPDA as $P$, except we now consider $\epsilon$-moves as moves with an actual letter of the alphabet (call it, say $\hep$).
For $w\in\Sigma^*$, let us define $E(w)$ to be all the possible ways of placing $\hep$ in $w$:
$$ E(w) = \set{\hep^{n_0}a_0\hep^{n_1}\cdots\hep^{n_k}a_k\hep^{n_{k+1}}}[w=a_0\cdots a_k,\,n_i\geq0] $$
Then $w\in\L(P)$ if and only if there exists a $\widehat w\in E(w)$ such that $\widehat w\in\L(\widehat P)$.

Let's formalize the concept of $\widehat P$.
Given a DPDA $P=(\Sigma,\Gamma,Q,q_0,F,\delta)$, define $\widehat P=(\Sigma\cup\hep,\Gamma,Q,q_0,F,\widehat\delta)$ where for $q\in Q,a\in\Sigma,A\in\Gamma_\epsilon$, $\widehat\delta(q,a,A)=\delta(q,a,A)$.
And instead of $\epsilon$-transitions we have explicit $\hep$-transitions:
$$ \widehat\delta(q,\hep,A) = \delta(q,\epsilon,A) $$
Let us denote by $\hvdash$ the derivation relation of $\widehat P$ (contrasting with $P$'s, which we use $\vdash$ for).
We call $\widehat P$ the \emph{$\epsilon$-free counterpart of $P$}.

It would be nice to have a function mapping $w\in\L(P)$ to $\epsilon(w)\in\L(\widehat P)$.
We can construct such a function inductively.
Since $P$ is deterministic, for every configuration $c_0$, there is at most one configuration $c_1$ such that $c_0\vdash c_1$.
Thus for a given configuration $c_0$ we can construct a unique path $c_0\vdash c_1\vdash\cdots\vdash c_n\nvdash$.
We define $\epsilon(c_0)$ by induction on $n$.
\begin{enumerate}
    \item For $n=0$, meaning $(p,w,W)\nvdash$, we define $\epsilon(p,w,W)=w$.
    \item We split into two cases for $n>0$:
        \begin{enumerate}
            \item if $(p,w,W)\vdash(q,w,W')$, we define $\epsilon(p,w,W)=\hep\epsilon(q,w,W')$.
            \item if $(p,aw,W)\vdash(q,w,W')$, we define $\epsilon(p,aw,W)=a\epsilon(q,w,W')$.
        \end{enumerate}
\end{enumerate}
And we let $\epsilon(w)=\epsilon(q_0,w,\epsilon)$.

We prove the correctness of this definition in the following lemmas:

\begin{lemma}

    $(p,w,W)\vdash(q,w',W')$ iff $(p,\epsilon(p,w,W),W)\hvdash(q,\epsilon(q,w',W'),W')$

\end{lemma}

\begin{proof}

    If $(p,w,W)\vdash(q,w',W')$, we split into cases:
    \begin{enumerate}
        \item If $w'=w$, then $\epsilon(p,w,W)=\hep\epsilon(q,w,W')$ and the transition is an $\epsilon$-transition.
         So
         $$ (p,\epsilon(p,w,W),W) = (p,\hep\epsilon(q,w,W'),W) \hvdash (q,\epsilon(q,w,W'),W') $$
        \item If $w=aw'$ then $\epsilon(p,w,W)=a\epsilon(q,w,W')$ and
        $$ (p,\epsilon(p,w,W),W)=(p,a\epsilon(q,w,W'),W') \hvdash (q,\epsilon(q,w,W'),W') $$
    \end{enumerate}

    And conversely, suppose $(p,w,W)\vdash(p',w'',W'')$
    Then we just showed that $(p,\epsilon(p,w,W),W)\hvdash(p',\epsilon(p',w'',W''),W'')$.
    Thus $p'=q$, $W''=W'$.
    Then we just split into cases regarding if $w=w''$, or $w=aw''$.
    \qed

\end{proof}

Immediately from this lemma we get by induction on $n\geq0$ that
$$ (p,w,W)\vdash^n(q,w',W') \iff (p,\epsilon(p,w,W),W)\hvdash^n(q,\epsilon(q,w',W'),W') $$
Now, $(p,w,W)\nvdash$ if and only if $\epsilon(p,w,W)=w$ by definition.
So we get that
$$ (p,w,W)\vdash^n(q,w',W')\nvdash \iff (p,\epsilon(p,w,W),W)\hvdash^n(q,w',W')\hnvdash $$
In particular, for $f\in F$ (since we are assuming there are no $\epsilon$-transitions from final states):
$$ (q_0,w,\epsilon)\vdash^*(f,\epsilon,W) \iff (q_0,\epsilon(w),\epsilon) \hvdash^* (f,\epsilon,W) $$
Thus we get the desired result:

\begin{theorem}

    $w\in\L(P)$ if and only if $\epsilon(w)\in\L(\widehat P)$.

\end{theorem}

We now prove a related result.

\begin{lemma}

    For $\hat w\in E(w)$, if $(p,\hat w,W)\hvdash(q,\hat u,U)$ then $(p,w,W)\vdash(q,u,U)$.
    By induction this means that $(p,\hat w,W)\hvdash^*(q,\hat u,U)$ implies $(p,\hat w,W)\hvdash(q,u,U)$.

\end{lemma}

\begin{proof}

    Notice that substrings of strings from $E(w)$ are strings in $E(u)$ for some substring $u$ of $w$, so this statement is well-formed.
    Now, if $\hat w=\hep\hat u$ then we have $(p,\hep\hat u,W)\hvdash(q,\hat u,U)$, so by definition this means that (since $\hat u\in E(w)$), $(p,u,W)\vdash(q,u,U)$.
    And if $\hat w=a\hat u$ then we have the desired result as well, since the transitions of $\widehat P$ and $P$ on elements of $\Sigma$ are identical.
    \qed

\end{proof}

\begin{theorem}

    $w\in\L(P)$ if and only if there exists a $\hat w\in E(w)$ such that $\hat w\in\L(\widehat P)$.

\end{theorem}

\begin{proof}

    The first direction is true due to the above theorem, taking $\hat w=\epsilon(w)$.
    For the converse, suppose $\hat w\in\L(\widehat P)$, meaning $(q_0,\hat w,\epsilon)\hvdash^*(f,\epsilon,W)$.
    By the above lemma, this means $(q_0,w,\epsilon)\vdash^*(f,\epsilon,W)$ and so $w\in\L(P)$.
    \qed

\end{proof}

Summarizing,

\begin{corollary}

    The following are equivalent:
    \begin{enumerate}
        \item $w\in\L(P)$
        \item $\epsilon(w)\in\L(\widehat P)$
        \item for some $\hat w\in E(w)$, $\hat w\in\L(\widehat P)$.
    \end{enumerate}

\end{corollary}

%Let $P$ be a DPDA.
%Let us define the \emph{set of transition symbols} to be
%$$ \T = \set{(p\vdash q+W),\;(p\vdash q-A)}[p,q\in Q,W\in\Gamma^*,A\in\Gamma] $$
%Furthermore, we define the \emph{projections} $\pi_b,\pi_e\colon\T\to Q$ where $\pi_b(p\vdash q\circ X)=p$ and $\pi_e(p\vdash q\circ X)=q$.
%We extend these to $\T^+\to Q$ where
%$$ \pi_b(t_1\cdots t_n) = \pi_b(t_1),\qquad \pi_e(t_1\cdots t_n) = \pi_e(t_n) $$

Let $P$ be a DPDA.
Let us define the \emph{set of transition symbols} to be
$$ \T = \set{(p\vdash q),\;(p\vdash q+A),\;(p\vdash q-A)}[p,q\in Q,A\in\Gamma] $$
Furthermore, we define the \emph{projections} $\pi_b,\pi_e\colon\T\to Q$ where $\pi_b(p\vdash q\circ X)=p$ and $\pi_e(p\vdash q\circ X)=q$.
We extend these to $\T^+\to Q$ where
$$ \pi_b(t_1\cdots t_n) = \pi_b(t_1),\qquad \pi_e(t_1\cdots t_n) = \pi_e(t_n) $$

We also define the \emph{stack function} $\S\colon\T^*\times\Gamma^*\partial\Gamma^*$ recursively as follows:
\begin{enumerate}
    \item $\S(\epsilon;W)=W$
    \item if $\bar t$ ends with $\pi_e(\bar t)=q$, then
    \begin{enumerate}
        \item $\S(\bar t(q\vdash p);W)=\S(\bar t;W)$
        \item $\S(\bar t(q\vdash p+W');W)=\S(\bar t;W'W)$
        \item $\S(\bar t(q\vdash p-A);AW)=\S(\bar t;W)$
    \end{enumerate}
\end{enumerate}
Otherwise $\S$ is left undefined.

We extend $\S$ to $\S\colon\P(\T)^*\times\Gamma^*\to\P(Q\times\Gamma^*\times Q)$ by
$$ \S(T_1\cdots T_n;W) = \set{(\pi_b(\bar t),\,\S(\bar t;W),\,\pi_e(\bar t))}[\bar t=t_1\cdots t_n,\,t_i\in T_i] $$
For $\epsilon$, we let
$$ \S(\epsilon;W) = \set{(q,W,q)}[q\in Q] $$
this agrees (pretty much) with our definition.

An immediate consequence of our definitions is

\begin{lemma}

    If $\bar S,\bar T\in\P(\T)^*$
    $$ (p_1,W_1,p_2)\in \S(\bar S;W_0) \hbox{ and } (p_2,W_2,p_3)\in \S(\bar T;W_1) $$
    then $(p_1,W_2,p_3)\in \S(\bar S\bar T;W_0)$.

\end{lemma}

Given a letter $a\in\Sigma$, we define $\uline a\in\P(\T)$ by
%\begin{align*}
%    \uline a &{}= \set{p\vdash q+W}[(p,a,\epsilon)\vdash(p',\epsilon,\epsilon)\vdash^*(q,\epsilon,W)]\cup\\
%    &\hphantom{{}={}}\set{p\vdash q-A}[(p,a,A)\vdash(q,\epsilon,\epsilon)]
%\end{align*}
\begin{align*}
    \uline a &{}= \set{p\vdash q}[\delta(p,a,\epsilon)=(q,\epsilon)]\cup\\
    &\hphantom{{}={}} \set{p\vdash q-A}[\delta(p,a,A)=(q,\epsilon)]
\end{align*}
Similarly we define
%\begin{align*}
%    \uline\epsilon &{}= \set{p\vdash q+W}[(p,\epsilon,\epsilon)\vdash^+(q,\epsilon,W)]\cup\\
%    &\hphantom{{}={}}\set{p\vdash q-A}[(p,\epsilon,A)\vdash(q,\epsilon,\epsilon)]
%\end{align*}
\begin{align*}
    \uline\hep &= \set{p\vdash q+A}[\delta(p,\epsilon,\epsilon)=(q,A)]\cup\\
    &\hphantom{{}={}} \set{p\vdash q-A}[\delta(p,\epsilon,A)=(q,\epsilon)]
\end{align*}
Given a word $w=a_1\cdots a_n\in(\Sigma\cup\hep)^*$, we define $\uline w=\uline a_1\cdots\uline a_n\in\P(\T)^*$.

These definitions come together in the following lemma:

\begin{lemma}   \label{lem:stackandpda}

    For $w\in(\Sigma\cup\hep)^+$,
    $$ (p,W',q)\in \S(\uline{w};W) \iff (p,w,W)\hvdash^*(q,\epsilon,W') . $$

\end{lemma}

\begin{proof}

    First suppose that $(p,W',q)\in \S(\uline w;W)$.
    For the base case of $w$ having a length of one, we split into two cases:
    \begin{enumerate}
        \item if $w=a\in\Sigma$, then $(p,W',q)\in \S(\uline a;W)$ if and only if $p\vdash q\in\uline a$ and $W'=W$, or if $p\vdash q- A\in\uline a$ and $W=AW'$.
        In the former, this means that $\delta(p,a,\epsilon)=(q,\epsilon)$ and so $(p,a,W)\hvdash(q,\epsilon,W=W')$ as required.
        In the latter case, this means that $\delta(p,a,A)=(q,\epsilon)$ and so $(p,a,W=AW')\hvdash(q,\epsilon,W')$ as required.
        \item if $w=\hep$, then $(p,W',q)\in \S(\uline\hep;W)$ if and only if $p\vdash q+ A\in\uline\hep$ and $W'=AW$, or if $p\vdash q- A\in\uline\hep$ and $W=AW'$.
        Then we do the same as before and chase definitions.
    \end{enumerate}
    We now induct.
    If $(p,W',q)\in \S(\uline w\uline a;W)$, suppose $t_i\in\uline w_i$ and $s\in\uline a$ such that $W'=\S(\bar ts;W)$ and $\pi_b(\bar t)=p$ and $\pi_e(s)=q$.
    Now, let $W''=\S(\bar t;W)$ and $p'=\pi_e(\bar t)$.
    Then $(p,W'',p')\in \S(\uline w;W)$, so by induction $(p,w,W)\hvdash^*(p',\epsilon,W'')$ and so $(p,wa,W)\hvdash^*(p',a,W'')$.
    Furthermore, we know that $(p',W',q)\in \S(\uline a;W'')$ and so by the base case $(p',a,W'')\hvdash(q,\epsilon,W')$.
    Together this tells us that $(p,wa,W)\hvdash^*(q,\epsilon,W')$ as required.

    Now we assume the converse: that $(p,w,W)\hvdash^*(q,\epsilon,W')$.
    Again we induct on the length of $w$ (which is the length of the derivation).
    For the base case we have that $(p,a,W)\hvdash(q,\epsilon,W')$.
    Then $\delta(p,a,\_)=(q,\_)$, and splitting the options here into cases on $\_$, we see that indeed $(p,W',q)\in \S(\uline a;W)$.

    Now suppose that
    $$ (p,wa,W) \hvdash (q',a,W'') \hvdash (q,\epsilon,W') $$
    Then we have that $(p,w,W)\hvdash^*(q',a,W'')$ so by induction $(p,W'',q')\in \S(\uline w;W)$.
    And by the base case we have that $(q',W',q)\in \S(\uline a;W'')$, so then $(p,W',q)\in \S(\uline w\uline a;W)$ as required.
    \qed

\end{proof}

In particular we have that for $f\in F$ and $\hat w\in E(w)$ (strictly speaking, we need to quantify by saying ``there exists a $\hat w\in E(w)$''):
$$ (q_0,W,f) \in \S(\uline{\hat w};\epsilon) \iff (q_0,\hat w,\epsilon)\hvdash^*(f,\epsilon,W) $$
and thus
$$ w \in \L(P) \iff \hat w \in \L(\widehat P) \iff (q_0,W,f) \in \S(\uline{\hat w};\epsilon) $$

\subsection{The main theorem}

We define the \emph{transition operator} $\tr\colon\T\times\T\partial\T$ as follows:
\begin{align*}
    (p\vdash q+ A),\,(p\vdash r- A) &\mapsto (p\vdash r)\\
    (p\vdash q+ A),\,(q\vdash r) &\mapsto (p\vdash r+ A)\\
    (p\vdash q),\,(q\vdash r+ A) &\mapsto (p\vdash r+ A)\\
    %(p\vdash q- A),\,(q\vdash r) &\mapsto (p\vdash r- A)\\
    %(p\vdash q),\,(q\vdash r- A) &\mapsto (p\vdash r- A)\\
    (p\vdash q),\,(q\vdash r) &\mapsto (p\vdash r)
\end{align*}
This extends to $\tr\colon\P(\T)\times\P(\T)\to\P(\T)$ by $\tr(S_1,S_2)=\set{\tr(s_1,s_2)}[s_i\in S_i]$.

The following lemma is immediate:

\begin{lemma}

    If $\tr(t_n,t_{n+1})$ is defined, then for every $W\in\Gamma^*$,
    $$ \S(t_1\cdots t_{n-1}\tr(t_n,t_{n+1});W) = \S(t_1\cdots t_nt_{n+1}; W) $$
    and the lhs is defined iff the rhs is.

\end{lemma}

This implies

\begin{lemma}

    If $\tr(t_i,t_{i+1})$ is defined for any $i$, then for every $W\in\Gamma^*$,
    $$ \S(t_1\cdots \tr(t_i,t_{i+1})\cdots t_n; W) = \S(t_1\cdots t_it_{i+1}\cdots t_n;W) $$

\end{lemma}

An immediate lemma is the following:

\begin{lemma}

    $\pi_\X(t,s)=\pi_\X(\tr(t,s))$ for $\X\in\set{b,e}$.
    Thus if $\bar t\in T^+$ and $\bar s\in T^+$ is the result of applying $\tr$ to $\bar t$ at an arbitrary position, then $\pi_\X(\bar t)=\pi_\X(\bar s)$ for $\X\in\set{b,e}$.

\end{lemma}

We define a $\SR1$ schema $\widehat B$ which will accept $\L(\widehat P)$.
We define
$$ \widehat B = (\Sigma\cup\hep, \hat\Lambda, \square, \set\top, \hib) $$
where
$$ \hat\Lambda = \P(\T) \cup \P(Q\top) \cup \set{\square,\top} $$
where $Q\top=\set{p\top}[p\in Q]$.
And we define the initial reduction function as follows:
\begin{enumerate}
    \item $\hib(S_1,S_2)=\tr(S_1,S_2)$ for $S_i\in\P(\T)$ if this is non-empty; otherwise leave it undefined.
    \item $\hib(S,\square)=\set{p\top}[(p\vdash f+ A)\in S\hbox{ or }(p\vdash f)\in S,\hbox{ where $f\in F$}]$ for $S\in\P(\T)$, if this is non-empty; otherwise leave it undefined.
    \item $\hib(S_1,S_2)=\set{p\top}[(p\vdash q+ A)\in S_1\hbox{ or }(p\vdash q)\in S_1\hbox{ and }q\top\in S_2]$ for $S_1\in\P(\T)$ and $S_2\in\P(Q\top)$, if this is non-empty; otherwise leave it undefined.
    \item $\hib(\square,S)=\top$ if $S\in\P(Q\top)$ and $q_0\top\in S$.
\end{enumerate}

\begin{lemma}

    For $\bar S\in\P(\T)^*$, $\square\bar S\square\hto^*\square\hb^*(\bar S)\square$.

\end{lemma}

\begin{proof}

    $\square,S$ is only reduced if $S\in\P(Q\top)$.
    And in general we have that if $\bar S\hto^*\bar T$ then $\bar S\square\hto^*\bar T\square$.
    And since $\square\bar S\hto^*\square\hb^*(\bar S)$ (since reducing $\bar S$ does not produce elements of $\P(Q\top)$), we get the desired result.
    \qed

\end{proof}

\begin{lemma}

    For $\bar S\in\P(\T)^*$, let $\bar T=\hb^*(\bar S)$.
    \begin{enumerate}
        \item if $\bar T=T$, then for $t\in T$ if $\S(t)$ is defined then $t$ is not of the form $p\vdash q- A$.
        \item otherwise, for any $t_i\in T_i$ for which $\S(\bar t)$ is defined, $t_i$ is of the form $p\vdash q+ A$.
    \end{enumerate}

\end{lemma}

\begin{proof}

    $\S(p\vdash q- A)$ is undefined, so the first point is trivial.

    Now if $\bar T$ has a length greater than $1$, let $i$ be the first index where $p\vdash q=t_i$ or $p\vdash q- A=t_i$.
    If $p\vdash q=t_i$, then if $i=1$, we can look to $t_{i+1}$, which must be either $q\vdash r$ or $q\vdash r+ A$, and in any case $\tr$ is defined and so $\bar T$ can be reduced, in contradiction.
    If $i>1$, then similarly we look to the left of $i$, and this must be of the form $r\vdash p+ A$ (since $i$ is the first index in which something other than $p\vdash q+ A$ occurs), and again $\tr$ is defined.

    Similarly if $p\vdash q- A=t_i$ then $i>1$, and in order for the stack to be defined $t_{i-1}=r\vdash p+ A$, so $\tr$ can be defined.
    \qed

\end{proof}

\begin{lemma}

    For $\bar S\in\P(\T)^*$, $\S(\bar S)=\S(\hb(\bar S))$

\end{lemma}

This is immediate since $\hb$ only applies $\tr$ on $\bar S$, which doesn't affect $\S$.

\begin{lemma}

    For $\bar S\in\P(\T)^*$, $\square\bar S\square\hto^*\top$ if and only if $(q_0,W,f)\in \S(\bar S)$.

\end{lemma}

\begin{proof}

    Let $\bar T=\hb^*(\bar S)$, then we know that $\square\bar S\square\hto^*\square\bar T\square$ and $\S(\bar S)=\S(\bar T)$.

    For $\bar T=T$, then $\square T\square\hto^*\top$ if and only if $q_0\vdash f$ or $q_0\vdash f+ A$ is in $T$.
    And in either case $(q_0,W,f)\in\S(\bar T)$.
    Now, if $\bar T=\bar T'T$, then $\square\bar T'T\square\hto^*\top$ means that $p\vdash f+ A\in T$ so that it can reduce with $\square$.
    Then $\bar T'$ reduces with $\hib(T,\square)$ if and only if $q\vdash p+ A$ is in the end of $\bar T'$ and $p\vdash f+ B$ is in $T$.
    And so on inductively we see that this means that $\bar T$ starts with $q_0\vdash q_1+ A_0$ and then has $q_1\vdash q_2+ A_1$ and so on until $q_n\vdash f+ A_n$.
    Thus $(q_0,A_0\cdots A_n,f)\in\S(\bar T)=\S(\bar S)$.

    And conversely if $(q_0,A_0\cdots A_n,f)\in\S(\bar T)=\S(\bar S)$, then this means that $\bar T$ consists of $T_0\cdots T_n$ such that $q_i\vdash q_{i+1}+ A_i\in T_i$.
    Thus $\square\bar T\square\hto^*\top$ as required.
    \qed

\end{proof}

\begin{theorem}

    The following are equivalent:
    \begin{enumerate}
        \item $w\in\L(P)$
        \item $\epsilon(w)\in\L(\hat B)$
        \item there exists a $\hat w\in E(w)$ such that $\hat w\in\L(\hat B)$
    \end{enumerate}

\end{theorem}

\begin{proof}

    We know that for $\hat w\in E(w)$, $\hat w\in\L(\hat B)$ if and only if $\square\uline{\hat w}\square\hto^*\top$ iff $(q_0,W,f)\in\S(\uline{\hat w})$ iff $\hat w\in\L(\hat P)$.
    Since this is for an arbitrary $\hat w\in E(w)$, this proves the three equivalences.
    \qed

\end{proof}

%Notice that we can actually refine our definition of $\hat\Lambda$ in $\hat B$.
%Since we start with strings from $\Sigma\cup\hep$, $\hat\Lambda$ must contain $\uline a,\uline\hep$ in order to have a representation of $\Sigma\cup\hep$.
%Now notice that applying $\hib$ to values from $\P(\T)$ gives values in $\set{(p\vdash q),(p\vdash q+ A)}$.
%So we can define $\hat\Lambda$ to be
%$$ \hat\Lambda = \uline{\Sigma\cup\hep} \cup \P\set{(p\vdash q),(p\vdash q+ A)}[p,q\in Q,A\in\Gamma] \cup \P(Q\top) \cup \set{\square,\top} $$

We define the relation $\vdash$ on $Q$, where $p\vdash q$ means that $(p,\epsilon,\epsilon)\vdash(q,\epsilon,W)$
And $\vdash^*$ is its reflexive and transitive closure.

We now define our $\SR1$ schema, $B=(\Sigma,\Lambda,\square,\A,\ib)$.
We begin by defining
$$ \Lambda = \P(\T) \cup \P(Q\top) $$
Now we define $\square=\set{f\top}[f\in F]\in\P(Q\top)$, and we define $\ib$ as follows:
\begin{enumerate}
    \item for $S\in\P(\T)$, $\ib(S,\epsilon)=S\cup\tr(S,\uline\hep)\cup\tr(\uline\hep,S)$ if this is not equal to $S$;
    \item for $S_1,S_2\in\P(\T)$, $\ib(S_1,S_2)=\tr(S_1,S_2)$ if non-empty;
    \item for $S_1\in\P(\T),S_2\in\P(Q\top)$,
    $$ \ib(S_1,S_2)=\set{p\top}[\hbox{$p\vdash q$ or $p\vdash q+A\in S_1$ and $q\vdash^*s$ such that $s\top\in S_2$}] $$
\end{enumerate}
And we define $S\in\A$ for $S\in\P(Q\top)$ if and only if $q_0\vdash^*p$ for some $p\top\in S$.

Now we claim that $w\in\L(B)$ iff $\epsilon(w)\in\L(\hat B)$.

\end{document}

### 2025-07-13 Paper/string-reduction-schemas.tex
DCFL <= 1-SR proof

@@
         \item $\S(\bar t(q\vdash p);W)=\S(\bar t;W)$
-        \item $\S(\bar t(q\vdash p+W');W)=\S(\bar t;W'W)$
-        \item $\S(\bar t(q\vdash p-A);AW)=\S(\bar t;W)$
+        \item $\S(\bar t(q\vdash p+A);W)=A\S(\bar t;W)$
+        \item if $\S(\bar t;W)=AW'$ then $\S(\bar t(q\vdash p-A);W)=W'$
     \end{enumerate}
@@
 
-We define the \emph{transition operator} $\tr\colon\T\times\T\partial\T$ as follows:
+We define the \emph{transition reduction operator} $\tr\colon\T\times\T\partial\T$ as follows:
 \begin{align*}
@@
 
-We define a $\SR1$ schema $\widehat B$ which will accept $\L(\widehat P)$.
-We define
-$$ \widehat B = (\Sigma\cup\hep, \hat\Lambda, \square, \set\top, \hib) $$
-where
-$$ \hat\Lambda = \P(\T) \cup \P(Q\top) \cup \set{\square,\top} $$
-where $Q\top=\set{p\top}[p\in Q]$.
-And we define the initial reduction function as follows:
+%Notice that we can actually refine our definition of $\hat\Lambda$ in $\hat B$.
+%Since we start with strings from $\Sigma\cup\hep$, $\hat\Lambda$ must contain $\uline a,\uline\hep$ in order to have a representation of $\Sigma\cup\hep$.
+%Now notice that applying $\hib$ to values from $\P(\T)$ gives values in $\set{(p\vdash q),(p\vdash q+ A)}$.
+%So we can define $\hat\Lambda$ to be
+%$$ \hat\Lambda = \uline{\Sigma\cup\hep} \cup \P\set{(p\vdash q),(p\vdash q+ A)}[p,q\in Q,A\in\Gamma] \cup \P(Q\top) \cup \set{\square,\top} $$
+
+We define $\S_\epsilon$ similar to $\S$, accept we allow $\epsilon$-transitions.
+Specifically, $\S_\epsilon\colon Q\times\T^*\times\Gamma^*\partial\P_{<\omega}(\Gamma^*\times Q)$ where first we define
+$$ \S_\e(p_0,\e;W)=\set{(W',q)}[(p_0,\e,W)\vdash^*(q,\e,W')] $$
+and we recurse:
 \begin{enumerate}
-    \item $\hib(S_1,S_2)=\tr(S_1,S_2)$ for $S_i\in\P(\T)$ if this is non-empty; otherwise leave it undefined.
-    \item $\hib(S,\square)=\set{p\top}[(p\vdash f+ A)\in S\hbox{ or }(p\vdash f)\in S,\hbox{ where $f\in F$}]$ for $S\in\P(\T)$, if this is non-empty; otherwise leave it undefined.
-    \item $\hib(S_1,S_2)=\set{p\top}[(p\vdash q+ A)\in S_1\hbox{ or }(p\vdash q)\in S_1\hbox{ and }q\top\in S_2]$ for $S_1\in\P(\T)$ and $S_2\in\P(Q\top)$, if this is non-empty; otherwise leave it undefined.
-    \item $\hib(\square,S)=\top$ if $S\in\P(Q\top)$ and $q_0\top\in S$.
+    \item for $(p\vdash q)$, if $(W',p)\in\S_\e(p_0,\bar t;W)$ and $(q,\epsilon,W')\vdash^*(s,\epsilon,W'')$ then $(W'',s)\in\S_\e(p_0,\bar t(p\vdash q);W)$.
+    \item for $(p\vdash q+A)$, if $(W',p)\in\S_\e(p_0,\bar t;W)$ and $(q,\epsilon,AW')\vdash^*(s,\epsilon,W'')$, then $(W'',s)\in\S_\e(p_0,\bar t(p\vdash q+A);W)$.
+    \item for $(p\vdash q-A)$, if $(AW',p)\in\S_\e(p_0,\bar t;W)$ and $(q,\epsilon,W')\vdash^*(s,\epsilon,W'')$, then $(W'',s)\in\S_\e(p_0,\bar t(p\vdash q+A);W)$.
 \end{enumerate}
+We extend $\S_\e$ to $\P_{<\omega}(\T)^*\times\Gamma^*\to\P_{<\omega}(Q\times\Gamma^*\times Q)$ where
+$$ \S_\e(\bar S;W) = \set{(p_0,W',p_e)}[p_0\in Q,\,(W',p_e)\in\S_\e(p_0,\bar S;W)] $$
+
+Furthermore, for $\bar S=S_1\cdots S_k\in\P(\T)^*$, let us define
+$$ \uline E(\bar S) = \set{\uline\hep^{n_0}S_1\uline\hep^{n_1}\cdots S_k\uline\hep^{n_k}}[n_i\geq0] $$
+so that $\uline E(\uline w)=\uline{E(w)}$.
+
+An immediate result by definition is given two strings of the same length $\bar S$ and $\bar T$, if we define their union $\bar S\cup\bar T$ pointwise (meaning $(\bar S\cup\bar T)_i=S_i\cup T_i$), then we have
+$$ \S_\e(\bar S\cup\bar T;W) = \S_\e(\bar S;W)\cup\S_\e(\bar T;W) $$
 
@@
 
-    For $\bar S\in\P(\T)^*$, $\square\bar S\square\hto^*\square\hb^*(\bar S)\square$.
+    $$ \S_\e(\bar S;W) = \bigcup_{\hbar S\in\uline E(\bar S)}\S(\hbar S;W) $$
+    in particular
+    $$ \S_\e(\uline w;\e) = \bigcup_{\hat w\in E(w)}\S(\hat{\uline w};\e) $$
 
@@
 
-    $\square,S$ is only reduced if $S\in\P(Q\top)$.
-    And in general we have that if $\bar S\hto^*\bar T$ then $\bar S\square\hto^*\bar T\square$.
-    And since $\square\bar S\hto^*\square\hb^*(\bar S)$ (since reducing $\bar S$ does not produce elements of $\P(Q\top)$), we get the desired result.
+    This is done via induction on $\bar S$.
+    For $\bar S=\e$, if $(p_0,W',p_e)\in\S_\e(\e;W)$ then $(p_0,\e,W)\vdash^n(p_e,\e,W')$ then $(p_0,W',p_e)\in\S(\hep^n;W)$.
+    And so on, the idea being that when $(q,\e,W')\vdash^n(s,\e,W'')$, we replace this with $\hep^n$.
     \qed
@@
 
-    For $\bar S\in\P(\T)^*$, let $\bar T=\hb^*(\bar S)$.
-    \begin{enumerate}
-        \item if $\bar T=T$, then for $t\in T$ if $\S(t)$ is defined then $t$ is not of the form $p\vdash q- A$.
-        \item otherwise, for any $t_i\in T_i$ for which $\S(\bar t)$ is defined, $t_i$ is of the form $p\vdash q+ A$.
-    \end{enumerate}
+    For $w\in\Sigma^*$,
+    $$ (p,W',q)\in\S_\e(\uline w;W) \iff (p,w,W)\vdash^*(q,\e,W') $$
 
@@
 
-    $\S(p\vdash q- A)$ is undefined, so the first point is trivial.
-
-    Now if $\bar T$ has a length greater than $1$, let $i$ be the first index where $p\vdash q=t_i$ or $p\vdash q- A=t_i$.
-    If $p\vdash q=t_i$, then if $i=1$, we can look to $t_{i+1}$, which must be either $q\vdash r$ or $q\vdash r+ A$, and in any case $\tr$ is defined and so $\bar T$ can be reduced, in contradiction.
-    If $i>1$, then similarly we look to the left of $i$, and this must be of the form $r\vdash p+ A$ (since $i$ is the first index in which something other than $p\vdash q+ A$ occurs), and again $\tr$ is defined.
-
-    Similarly if $p\vdash q- A=t_i$ then $i>1$, and in order for the stack to be defined $t_{i-1}=r\vdash p+ A$, so $\tr$ can be defined.
-    \qed
+    This is immediate due to the previous lemma and lemma \ref{lem:stackandpda}.
 
 \end{proof}
-
-\begin{lemma}
-
-    For $\bar S\in\P(\T)^*$, $\S(\bar S)=\S(\hb(\bar S))$
-
-\end{lemma}
-
-This is immediate since $\hb$ only applies $\tr$ on $\bar S$, which doesn't affect $\S$.
-
-\begin{lemma}
-
-    For $\bar S\in\P(\T)^*$, $\square\bar S\square\hto^*\top$ if and only if $(q_0,W,f)\in \S(\bar S)$.
-
-\end{lemma}
-
-\begin{proof}
-
-    Let $\bar T=\hb^*(\bar S)$, then we know that $\square\bar S\square\hto^*\square\bar T\square$ and $\S(\bar S)=\S(\bar T)$.
-
-    For $\bar T=T$, then $\square T\square\hto^*\top$ if and only if $q_0\vdash f$ or $q_0\vdash f+ A$ is in $T$.
-    And in either case $(q_0,W,f)\in\S(\bar T)$.
-    Now, if $\bar T=\bar T'T$, then $\square\bar T'T\square\hto^*\top$ means that $p\vdash f+ A\in T$ so that it can reduce with $\square$.
-    Then $\bar T'$ reduces with $\hib(T,\square)$ if and only if $q\vdash p+ A$ is in the end of $\bar T'$ and $p\vdash f+ B$ is in $T$.
-    And so on inductively we see that this means that $\bar T$ starts with $q_0\vdash q_1+ A_0$ and then has $q_1\vdash q_2+ A_1$ and so on until $q_n\vdash f+ A_n$.
-    Thus $(q_0,A_0\cdots A_n,f)\in\S(\bar T)=\S(\bar S)$.
-
-    And conversely if $(q_0,A_0\cdots A_n,f)\in\S(\bar T)=\S(\bar S)$, then this means that $\bar T$ consists of $T_0\cdots T_n$ such that $q_i\vdash q_{i+1}+ A_i\in T_i$.
-    Thus $\square\bar T\square\hto^*\top$ as required.
-    \qed
-
-\end{proof}
-
-\begin{theorem}
-
-    The following are equivalent:
-    \begin{enumerate}
-        \item $w\in\L(P)$
-        \item $\epsilon(w)\in\L(\hat B)$
-        \item there exists a $\hat w\in E(w)$ such that $\hat w\in\L(\hat B)$
-    \end{enumerate}
-
-\end{theorem}
-
-\begin{proof}
-
-    We know that for $\hat w\in E(w)$, $\hat w\in\L(\hat B)$ if and only if $\square\uline{\hat w}\square\hto^*\top$ iff $(q_0,W,f)\in\S(\uline{\hat w})$ iff $\hat w\in\L(\hat P)$.
-    Since this is for an arbitrary $\hat w\in E(w)$, this proves the three equivalences.
-    \qed
-
-\end{proof}
-
-%Notice that we can actually refine our definition of $\hat\Lambda$ in $\hat B$.
-%Since we start with strings from $\Sigma\cup\hep$, $\hat\Lambda$ must contain $\uline a,\uline\hep$ in order to have a representation of $\Sigma\cup\hep$.
-%Now notice that applying $\hib$ to values from $\P(\T)$ gives values in $\set{(p\vdash q),(p\vdash q+ A)}$.
-%So we can define $\hat\Lambda$ to be
-%$$ \hat\Lambda = \uline{\Sigma\cup\hep} \cup \P\set{(p\vdash q),(p\vdash q+ A)}[p,q\in Q,A\in\Gamma] \cup \P(Q\top) \cup \set{\square,\top} $$
 
@@
 
-Now we claim that $w\in\L(B)$ iff $\epsilon(w)\in\L(\hat B)$.
+\begin{lemma}
+
+    For $\bar S\in\P(\T)^*$, $\S_\e(\bar S)=\S_\e(\beta\bar S)$.
+
+\end{lemma}
+
+\begin{proof}
+
+    We must split into two cases: the first when only a single symbol of $\bar S$ is reduced, and the second when two symbols are reduced.
+
+    In the first case, suppose $\bar S=S_1\cdots S_i\cdots S_n$, and $\beta\bar S=S_1\cdots\ib S_i\cdots S_n$.
+    Now we know that $\ib(S_i)=S_i\cup\tr(S_i,\uline\hep)\cup\tr(\uline\hep,S_i)$, thus
+    $$ \S_\e(\beta\bar S) = \S_\e(\bar S)\cup\S_\e(S_1\cdots\tr(S_i,\uline\hep)\cdots S_n)\cup\S_\e(S_1\cdots\tr(\uline\hep,S_i)\cdots S_n) $$
+    We see then that $\S_\e(\bar S)\subseteq\S_\e(\beta\bar S)$.
+    Conversely, let us look at $E_1=\uline E(S_1\cdots\tr(S_i,\uline\hep)\cdots S_n)$.
+    Since $\S$ is $\tr$-invariant, taking $\S$ of $E_1$ is equal to $\S$ over $\uline E(S_1\cdots S_i\uline\hep\cdots S_n)\subseteq E(\bar S)$.
+    Thus we see that $\S_\e(S_1\cdots\tr(S_i,\uline\hep)\cdots S_n)\subseteq\S_e(\bar S)$, and similar for $\tr(\uline\hep,S_i)$, as required.
+
+    For the second case, this is just due to $\S$ being $\tr$-invariant.
+    \qed
+
+\end{proof}
+
+\begin{lemma}
+
+    For $\bar S\in\P(\T)^*$, $\bar S\square\to^*\top\in\A$ if and only if $(q_0,W,f)\in \S_\e(\bar S;\e)$.
+
+\end{lemma}
+
+\begin{proof}
+
+    Let $\bar T=\hb^*(\bar S)$, then we know that $\bar S\square\to^*\bar T\square$ and $\S_\e(\bar S)=\S_\e(\bar T)$.
+
+    Now, we know that $\bar T\square\to^*\top\in\A$ if and only if $p_i\vdash q_i+A_i\in T_i$ such that
+    \begin{enumerate}
+        \item $A_i\in\Gamma_\e$
+        \item $q_i\vdash^* p_{i+1}$
+        \item $q_0\vdash^* p_1$ and $q_n\vdash^* f\in F$
+    \end{enumerate}
+    And it is not hard to see that this is equivalent.
+    \qed
+
+\end{proof}
+
+In particular we have that $w\in\L(B)$ if and only if $(q_0,W,f)\in\S_\e(\uline w;\e)$ if and only if $(q_0,w,\e)\vdash^*(f,\e,W')$ if and only if $w\in\L(P)$ as required.
+
+\begin{theorem}
+
+    $B$ and $P$ are equivalent.
+
+\end{theorem}
+
+\begin{theorem}
+
+    $\DCFL\subseteq\SR1$
+
+\end{theorem}
 
